### 2020-06-12 ms_text.tex
Start to build

\title{Acceleration of plague outbreaks in the second pandemic}

\author{David J.\,D.\ Earn$^{a,c,e}$, 
Junling Ma$^{f}$, 
Hendrik N.\ Poinar$^{b,c,d,e}$,\\
Jonathan Dushoff$^{a,c,e}$,
and Benjamin M.\ Bolker$^{a,c,e}$\\
\href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
\small $^a$Department of Mathematics \& Statistics,\\
\small $^b$McMaster Ancient DNA Centre, Department of Anthropology,\\
\small $^c$Department of Biology, $^d$Department of Biochemistry,\\
\small and $^e$Michael G.\ deGroote Institute for Infectious Disease Research,\\
\small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
\small $^f$Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}

\begin{document}

\linenumbers

\maketitle

%%\begin{abstract}
{\bfseries
\noindent
Historical records reveal the temporal patterns of a sequence of plague epidemics in London, England, from the 14th to 17th centuries.  Analysis of these records shows for the first time that later epidemics spread significantly faster (``accelerated'').  Between the Black Death of 1348 and the later epidemics that culminated with the Great Plague of 1665, we estimate that the epidemic growth rate increased \foldval.
%%[\rdiffest $\times$, 95\% CI: \rdifflwr $\times$ to \rdiffupr $\times$].  
%%The cause of this change is unknown; despite rapid recent advances in paleogenomics, it is hard to confirm or reject the possibility of functional evolutionary changes in circulating strains of the plague bacterium \ypestis over this time period.
Currently available data do not provide enough information to infer the mode of plague transmission in any given epidemic; nevertheless, order-of-magnitude estimates of epidemic parameters suggest that the observed slow growth rates in the 14th century are inconsistent with direct (pneumonic) transmission.  
We discuss the potential roles of demographic and ecological factors such as climate change or human or rat population density in driving the observed acceleration.
}  
%%\end{abstract}

\bigskip

\leftline{plague $|$ London $|$ epidemic growth rate}

\bigbreak

\noindent\fbox{\parbox{0.98\textwidth}{{\bfseries Significance Statement:}

Epidemics of plague devastated Europe's population throughout the Medieval and Renaissance periods.  Genetic studies of small numbers of skeletal remains indicate that the causative agent of all these epidemics was the bacterium \emph{Yersinia pestis}, but such analyses cannot identify overall patterns of transmission dynamics.
Analysis of thousands of archival records from London, England, reveal that plague epidemics spread much faster in the 17th century than in the 14th century.}}

\bigbreak

\dropcap{P}lague epidemics have afflicted human populations since at least the 6th century \cite{McNe76,Wagn+14}.  These events have had dramatic and long-lasting effects on human demography and behaviour, especially those outbreaks associated with the ``second pandemic'' (14th--19th centuries) in Europe and Asia \cite{Pepy1669,Lath85,Defo1722,McNe76}, and have inspired many
theoretical studies of the ecology and evolution of infectious disease \cite{KermMcKe27,AndeMay91,KeelGill00,Baca12,Silv16,LewnTown16,Yue+16,Dean+18}.
We are now in the third pandemic (Modern Plague), with outbreaks continuing to occur in some parts of the world \cite{PerrFeth97,Neer+10,WHO10,ECDC17,Robe17}. Plague also remains a source of concern due to the bioterror potential of the causative agent, \emph{Yersinia~pestis} \cite{Ligo06,CDCplagueEmergencyFAQ}.

Recent advances in paleogenomics have definitively established that historical plague pandemics were caused by \ypestis
 \cite{Bos+11,Bene11}, as proposed in the 19th century after Yersin discovered the bacterium's link to bubonic plague \cite{Butl14}.
 Researchers have reconstructed the evolutionary history of plague and other pathogens by sequencing and reconstructing nearly complete pathogen genomes from persistent DNA fragments \cite{Schu+11, Bos+11,Deva+14}.
 The strain isolated from victims of the Black Death (London 1348) is  remarkably similar to extant human strains (Modern Plague): 
the core genomes\footnote{The core genome is about 80\% of the full genome, shared between all sequenced strains of \ypestis.} of these strains are $\approx 99.99\%$ similar \cite{Bos+11}, which makes it challenging to 
identify important evolutionary or ecological patterns
from genomic investigations alone.  Here, we complement genetic studies by exploring more traditional (historical, demographic, and epidemiological) sources of information
from a 300-year span of plague outbreaks in the same location (London), revealing a striking change in plague transmission dynamics over the course of the Renaissance period, namely a \foldval increase in the initial growth rate of outbreaks.

We quantify this change without making any assumptions about the underlying transmission processes, exploiting methodology that we have developed previously for this purpose \cite{Ma+14}.  We then consider how this inference can contribute to the debate concerning whether plague transmission was primarily indirect (via rat fleas) or direct (pneumonic human-to-human).  We argue that strictly pneumonic transmission in the 14th century is implausible, but that beyond this the best that can be done at present is to highlight the biological complexities and uncertainties that limit the potential for further inferences.

\mysection{Data}

The city of London, England, is unusual in the extent to which patterns of death and disease can be reconstructed from extant documents. We have analyzed three sources of data (\cref{F:plot3sources}):
\begin{description}
\item[London Bills of Mortality] The London Bills of Mortality \cite{Crei65,Tien+11} (LBoM) provide weekly counts of deaths by cause during each of the known plague epidemics in the 16th and 17th centuries.  The LBoM thus give us information specifically about mortality attributable to plague for each epidemic, which we will use as a proxy for plague incidence \cite{KermMcKe27,Ma+14}.  When available, the LBoM are our most reliable records (except for the epidemic of 1593; see \supp).
\item[Parish registers] After 1538 \cite[p.\,54]{Slac90}, London parish registers provide information on mortality.
%%, although these are not categorized by cause. 
We use weekly counts of deaths published online by Cummins, which are estimated to cover 80\% of total deaths \cite{Cumm+16}. During plague epidemics, most deaths were likely due to plague; we also adjust for non-plague mortality by estimating a baseline death rate from the years before and after an epidemic (see \Methodslink).
\hypertarget{mainsec:Data.wills}{}
\item[Wills and testaments] Before 1538, no direct tabulations of mortality for London are available. However, we do have detailed information on last wills and testaments, which can be used as proxies for mortality \cite{Slac90}. In particular, we have digitized and tabulated daily counts of wills recorded in the Court of Husting \cite{Shar1889} for the 14th century; these give us data for the plague epidemics of 1348, 1361, 1368 and 1375 (monthly counts for three of these epidemics were previously published by Cohn \cite{Cohn03}). We use online data from the Prerogative Court of Canterbury (PCC) \cite{Archives2018} 
for the plague epidemics of the 16th and 17th century.  Plague epidemics in London are mentioned in historical documents from the 15th century \cite[Ch.\,IV]{Crei65}, but during this period wills recorded in the Courts of Husting and Canterbury were too sparse to enable identification of epidemics.

Two dates are associated with each will: the date on which it was written, and the date on which it was probated (\ie accepted by the court after the death of the testator). Traditionally, demographers have used annual counts of wills by probate date, as testators are known to have died by these dates. While this approach works well for inferring long-term trends in annual mortality, probate dates are inappropriate for reconstructing detailed time series of mortality over short time scales, as there is a variable and irregular lag between the (unknown) date of the death and the probate date; most of the wills associated with 14th century epidemics were probated several months after they were written \cite{Bush19} (see \cref{F:willsVSprobates}).

Instead, we use the counts of wills \emph{written} in a given time interval
as a proxy for plague incidence. The date on which a will was written may precede the testator's death by a long period, but during severe epidemics will dates were likely to have been correlated with the fear of infection (and hence with disease incidence), just as internet searches for influenza symptoms can predict 21st century epidemic patterns \cite{Gins+09}.

When using wills as proxies for mortality or incidence, we must keep in mind that: 1) a much smaller proportion of the population is sampled (the will counts in \cref{F:plot3sources} are about 10 times smaller than the counts based on parish registers); 2) the subpopulation of individuals who wrote wills is strongly biased towards merchants and others who owned property (primarily males).
\end{description}

For the period since 1540, the existence of multiple sources allows us to cross-check results---in particular, to test our assertion that aggregated counts of wills by date written are an adequate proxy for mortality rates---and avoid relying on any questionable or poorly sampled segments of data \cite{RoosCurt18} (see \supp). In 1563 and 1593, the number of PCC wills is too small to extract a signal from the noise, but the data for the later three outbreaks verify that the numbers of wills written follow the plague epidemic patterns evident in the mortality records (\cref{F:timeseries}). Based on cross-correlation analysis and differences between epidemic peak times, will counts appear to have lagged mortality recorded in parish registers and LBoM by approximately 3--5 weeks; our inferences about epidemic growth rates are apparently robust to this lag (see \supp).

The LBoM indicate that there were 18 other plague epidemics in London between 1563 and 1666; these were all of much smaller magnitude and are discussed only in \supp.  No plague epidemics have occurred in London since the end of the Great Plague in 1666: a total of 77 deaths from plague are recorded in the LBoM after 1666, never more than 5 in a single week.  The last plague deaths recorded in the LBoM occurred during the week of 6 May 1679.

%
\newcommand{\pixht}{0.3\textheight}
\begin{figure}
\begin{center}
\mbox{
\hspace{1.2cm}
\includegraphics[height=\pixht]{images/WillExample1644_page86_1.pdf}
%% \includegraphics[height=\pixht]{WillExample1644_page86_1.eps}
\hspace{0.4cm}
\includegraphics[height=\pixht]{images/cripplegate_wellcome_parish.jpg}
%% \includegraphics[height=\pixht]{cripplegate_wellcome_parish.eps}
\hspace{0.4cm}
\includegraphics[height=\pixht]{images/bill1665adjust.pdf}
%% \includegraphics[height=\pixht]{bill1665adjust.eps}
}
\\ \vspace{0.5cm}
\includegraphics[width=\textwidth]{analysis/plots/plot_all.pdf}
%% \includegraphics[width=\textwidth]{plot_all.eps}
\end{center}
\vspace{-0.5cm}
\caption[]{
  \emph{Top left:} Part of a will proved in the Prerogative Court of Canterbury (PCC), dated 18 December 1644 \cite{Archives2018}.
  \emph{Top centre:} A parish register page from St Giles without Cripplegate, August 1665 \cite[courtesy of the Wellcome Collection: \url{https://wellcomecollection.org/works/uutd7scc}]{Bell24}.
  \emph{Top right:} One of the London Bills of Mortality (LBoM), for the week beginning 26 September 1665 (photo by Claire Lees).
\emph{Bottom:}
Mortality in London, England, 1340--1380 and 1540--1680, aggregated 4-weekly, plotted on a log scale.  The three distinct sources of data (see \cref{tab:datasources}) are: 
last wills and testaments of Londoners whose wills were probated in the the Court of Husting (14th c.{}) or the PCC (16th-17th c.{}),
weekly aggregations of burials listed in extant parish registers \cite{Slac90}, 
and weekly plague deaths listed in the LBoM \cite{Crei65}.
Major plague years are highlighted in yellow.  Aside from these and various minor plague years, there are notably unusual patterns during an epidemic of ``sweating sickness'' in 1551 \cite[p.\,70]{Slac90}, the ``influenza'' epidemic of 1557--9 \cite[p.\,70]{Slac90} (which coincided with the end of the reign of ``Bloody'' Mary~I and the ascension of Elizabeth~I in 1558 [indicated by a crown icon]), and the absence of a monarch during the Interregnum from 30 January 1649 to 29 May 1660.
}
\label{F:plot3sources}
\end{figure}
%

\mysection{Approach}

\paragraph{Estimation of epidemic growth rates}

While our collection of high-resolution mortality time series during plague epidemics allows us to estimate epidemic growth rates, we are constrained by the limitations of our data---the time series for each epidemic are short and noisy, especially during the 14th century (\eg the epidemics of 1368 and 1375 peaked at a total of 5 wills written per week). Thus we chose to use simple phenomenological models with a small number of parameters that can be fitted to short time series (starting size, growth rate, and total size: see \Methodslink), and do not attempt to separate different sources of variability or estimate the influence of various mechanistic processes (King \etal's exhortation to use epidemic models that separate process from observation noise \cite{King+2015} applies primarily to forecasting, which we are not attempting).
Furthermore, we primarily consider the initial growth rate $r$, which---unlike the basic reproduction number $\rzero$---can be estimated reliably from mortality data alone without any knowledge of the disease characteristics or natural history \cite{Ma+14}.

We estimate the epidemic growth rate for each epidemic as the maximum likelihood estimate of the initial growth rate of a logistic function fitted to the cumulative death curve (using first differences to avoid over-confidence; see \Methodslink\ for more detail), along with likelihood profile confidence intervals (CIs). To summarize the overall difference between epochs, we then use the estimates for each epidemic as observations in a linear mixed model including fixed effects of epoch (``early''/14th c.\ \emph{vs.}\ ``late''/16th-17th c.{}) and data source, and a random effect of outbreak year (see \hyperlink{outbreak.year}{below}), where each estimate was weighted by its certainty.

Finally, to robustly quantify the statistical significance of the difference between epochs, we ran a permutation test. We computed the null distribution of the between-epoch difference in growth rate by scrambling the estimates and associated CIs randomly across epochs. For each of the 126 possible permutations, we refitted the mixed model and extracted the estimated among-epoch difference in growth rate.

\paragraph{Estimating $\rzero$ and attack rate}

Our analysis focuses on the epidemic growth rate $r$, rather than the more commonly investigated basic reproduction number $\rzero$ (the expected number of secondary cases caused by a primary case in a completely susceptible population \cite{AndeMay91}), because estimates of $\rzero$ depend on information about the life history of the pathogen as well as on the epidemic curve itself \cite{WallLips07}. However, if we can estimate $\rzero$ then we can predict epidemic outcomes that we are unable to predict from $r$ alone; then, by comparing observed with theoretical outcomes under different scenarios, we can make inferences about modes of transmission.

In particular, if the host population is homogeneously mixed, we can use $\rzero$ to estimate the expected final size $Z$ (the proportion of the population infected by the end of an epidemic, also called the \emph{attack rate}) \cite{KermMcKe27,MaEarn06}. The \emph{observed attack rate} based on mortality records depends in turn on the fraction of infected individuals who die from disease. We can thus compare theoretical expectations of observed attack rates under different transmission modes with historical information about observed attack rates (see \hyperlink{implications}{Implications}).

We can use the distribution of the generation interval (the times elapsed between the moment a host is infected and the times at which they infect others) to compute $\rzero$ from $r$ \cite{WallLips07}. If we only know the \emph{mean} generation interval $\gentime$, we can still approximate $\rzero \approx r \gentime +1$, by assuming that the generation intervals are exponentially distributed; this approximation underestimates $\rzero$ in the typical case where the generation interval distribution is less variable than an exponential distribution with the same mean.

Beyond the information on generation interval that we use to derive $\rzero$ from $r$, the observed attack rate depends on additional characteristics of an epidemic. If the population does not mix homogeneously, the final size will typically decrease (\eg \cite[Fig.\,6]{Roll+15}). Since the \emph{case fatality proportion} (CFP, the proportion of plague cases who die) is not 100\%, the observed attack rate is less than the final size. Observed attack rates also decrease if some individuals are infected without showing symptoms (asymptomatic infections) \cite{Mars+67}, or without symptoms identified as plague (sub-clinical infections). Finally, the observed attack rate depends on the initial proportion of the population susceptible, which will be lower if some individuals have previously acquired immunity after previously non-fatal plague infections.

We can use independent estimates of the generation interval to estimate $\rzero$ (and hence the theoretical attack rate) for London epidemics under different transmission modes. By considering the additional (extreme) assumptions that \hypertarget{4assumptions}{} (1) the population was homogeneously mixed; (2) the population was initially 100\% susceptible; (3) there were no asymptomatic infections; and (4) all infected individuals died (\ie the CFP was 100\%), all of which increase the observed attack rate, we can calculate an upper bound on the total mortality we would expect to observe in an epidemic with a particular transmission mode.

\mysectionsamepage{Results}

The fitted models do as good a job as one could hope for in capturing the initial patterns of epidemic growth evident in the noisy will counts, and agree very well with the patterns of growth in the mortality time series that are available in the later epoch (\cref{F:timeseries}).

The growth rates for the 14th century epidemics are lower than those for the 16th-17th centuries (\cref{F:combdata}); they may also be more variable than in the later period, although some of the appearance of variability is due to larger uncertainties (presumably due to the relatively poor sampling in the early epoch). The mixed model quantifies the dramatic increase in growth rate (``acceleration'') of epidemics between epochs.
The growth rates increased \foldval in the late \emph{vs.}\ early epochs: $r$ increased \rdiffest $\times$ [95\% CI $(\rdifflwr\times, \rdiffupr\times)$] (see \Methodslink). The permutation test results support the statistical significance of this difference; the observed order of outbreaks yields the most extreme difference in growth rates out of all \nperm\ possible orderings, giving a two-tailed $p$ value of $\permfrac=\permpval$.

%
%%\begin{figure}[h!]
\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/timeseries.pdf}
%% \includegraphics[width=\textwidth]{timeseries.eps}
\end{center}
\caption[]{Observed time series (points) and \hyperlink{phenmod}{phenomenological model} fits (lines; see \Methodslink) that yield the estimated growth rates (listed in \cref{tab:params} and plotted in \cref{F:combdata}).  The data sources (\cref{tab:datasources}) are described in the main text and the caption to \cref{F:plot3sources}.  For visual comparison, wills were aggregated weekly to match the frequency of mortality observations (fits to wills were based on the original daily counts).
  \cref{F:plotwills14c,F:plot3sourcesmajors} display the data during the major plague epidemics on linear rather than logarithmic scales.
  Vertical dashed lines show 1 April, 1 July, and 1 October of each year.
  In 14th c. data (top row), zero-count weeks are shown along the bottom edge of the graph.
}
\label{F:timeseries}
\end{figure}
%

%
%%\begin{figure}[h!]
\begin{figure}
\begin{center} 
\includegraphics[width=6.5in]{analysis/plots/combdata.pdf} 
%% \includegraphics[width=6.5in]{combdata.eps} 
\end{center} \caption[Estimated initial growth rates]{Estimated initial growth rates ($r$) and 95\% profile CIs on a logarithmic scale, for each of the epidemics shown in \cref{F:timeseries}.  Center panels show mixed-model estimates (see \Methodslink) of overall average growth rate for each epoch (early:\ 14th c.\ \emph{vs.}\ late:\ 16th-17th c.{})\ and data source.  To aid interpretation, the additional vertical scales show the implied intrinsic reproductive number ($\rzerop$) and epidemic final size ($\Zp$) under the assumption that the generation interval distribution \cite{WallLips07} was the same as that estimated from 20th century pneumonic plague epidemics \cite{GaniLeac04}. All estimates and CIs are listed in \cref{tab:params} (which also lists the associated doubling times).}
\label{F:combdata}
\end{figure}
%

\begin{table}[h!] 
  \caption[Estimated $r$ and $1/r$]{Maximum likelihood estimates (MLEs) of the initial exponential growth rate ($r$) and doubling time ($(\log2)/r$) with their 95\% confidence intervals (CIs), obtained from the time series shown in \cref{F:timeseries} (see \Methodslink).  The goodness of fit measure ($R^2$) is the proportional reduction in the mean squared error, \ie $1-d^2/\sigma^2$ where $d^2$ is the model mean squared error and $\sigma^2$ is the (population) variance of the data; predictions and observations are aggregated to weekly before computing the $R^2$ for wills, for consistency among data sources. The implied basic reproduction numbers ($\rzerop$) and attack rates ($\Zp$), assuming the generation interval distribution for Modern pneumonic plague \cite{GaniLeac04}, are also shown.
All MLEs and associated CIs are shown in \cref{F:combdata}.
}
\begin{center}
  \footnotesize
  %%\rowcolors{2}{white}{verylightgray}
  \rowcolors{2}{white}{gray}
  \input{analysis/tables/fakeparams.tex}

\end{center}
\label{tab:params}
\end{table}

\FloatBarrier
\mysection{Discussion}

Plague epidemics in London appear to have been faster in the 16th-17th centuries than in the 14th. Our analysis shows that this difference is very likely real, rather than driven by changes in the types of documentary evidence that are available; in particular, we have shown that inferences from will counts are consistent with those from death counts when both are available, in the 17th century.

Why the later plague epidemics were faster than the earlier ones is not clear.  Nevertheless, to provide some context, we consider below a number of mechanisms which \emph{could} have contributed to increases in epidemic growth rates.  We then consider what our estimated epidemic growth rates suggest about the mode of transmission during the various London plague epidemics.

\subsection*{Possible causes of acceleration}

What factors could have caused plague epidemics in London in the 16th and 17th centuries to accelerate relative to the earlier plagues in the 14th century?

\paragraph{Pathogen or host evolution}

Faster growth could be a consequence of the pathogen evolving to infect hosts more effectively, or to remain infectious for longer.  However, such evolution may be challenging to demonstrate, particularly in light of the strong genetic similarity between the Second Pandemic and Modern Plague strains \cite{Bos+11}. Evolution of host resistance has also been suggested as a cause of changes in plague dynamics \cite{LewnTown16}, though we have no particular mechanisms in mind for how the evolution of resistance over the course of centuries could accelerate epidemics.

\paragraph{Shifts in transmission mode}

Modes of transmission may have differed between epochs, which could account for differences in epidemic growth rates.  The reservoir host of \ypestis in London was rat populations where it was primarily spread by flea vectors such as {{\em X.~cheopis}} \cite[p.\,54]{PerrFeth97}. This transmission mode gives rise to bubonic plague in humans, as a consequence of zoonotic spillover; epidemic growth rate is driven by spread among rats and fleas, with human infection and mortality as a secondary consequence. In contrast, when \ypestis infection in humans spreads to  the lungs, it gives rise to pneumonic plague, which can then spread directly from human to human like other severe respiratory infections.
It has also been more recently hypothesized that human ectoparasites (including lice and human fleas) could have supported indirect plague transmission without involving rats \cite{Dean+18,Laro+18}.

Pneumonic plague probably has shorter generation intervals than bubonic plague \cite{GaniLeac04,LaPerriere+2009}, and is often thought to spread much faster at the population level \cite{GaniLeac04}; one \emph{potential} explanation for our observations is that 14th century epidemics were primarily bubonically transmitted while 16th-17th century epidemics were primarily pneumonic.  With currently available data, we are unable to confirm or reject this hypothesis, but we consider below whether any inferences about transmission mode can be made at present.

\paragraph{Ecological and demographic changes}

The complete multi-species ecology of \ypestis is extremely complicated \cite{DubyYesz16}.  We focus on humans because the only data available for the epochs studied here concern humans, and not on the rest of the ecological community supporting \ypestis.  Human population size and population density in London increased enormously from the 14th to 17th centuries \cite{Finl81,FinlShea86,Kryl11} (\cref{tab:pop,F:london_pop}), and density increases were exacerbated in the 17th century by the isolation of the sick and their contacts in pest houses [A.~Carmichael, \emph{pers.\ comm.}].  Since bubonic plague epidemics are driven by spillover from rat-flea epidemics, human population density would only have a direct effect on the speed of pneumonic epidemics.  However, changes in human population (and living conditions) almost certainly affect rat, and flea, density \cite{McCo03}.  In addition, higher rat densities make it more likely that fleas departing dying rats end up on susceptible rat hosts.  The magnitudes of these effects are extremely difficult to estimate, but in \supp we explore them in a (crude) quantitative manner, without attempting to model the truly complex population dynamics that occur even during a single epidemic \cite{KeelGill00b,LaPerriere+2009}.

\paragraph{Environmental change}

The external environment represents the third side of the host-parasite-environment ``disease triangle'' \cite{Scho07}.  Northern European climates changed significantly between the 14th and 16th--17th centuries (the coldest period of the ``Little Ice Age'' occurred in the 17th century \cite{Matt39} \cite[Figure 1]{Jone+13}).  Furthermore, late-epoch epidemics occurred later in the calendar year, and hence in different seasons (\cref{F:timeseries}).  A variety of studies link climate (\eg overall wetness/dryness) or weather (\eg annual precipitation or temperature) to plague incidence or transmission \cite{Cava71,Xu+11,Xu+14,Pham+09,YueLee18,BenA+10,Scho+11}.  Climatic changes may have been responsible for the changes in plague epidemics in London between 1348 and 1666, but it is challenging to make reliable inferences due to the lack of consensus about climate and weather variations in Medieval/Renaissance Europe \cite{Nesj+08}.

\hypertarget{implications}{}
\subsection*{Implications of growth rate estimates for transmission mode}
\label{sec:implications}

One might hope to exploit ancient DNA \cite{Bos+11,Wagn+14} to identify the transmission mode during London plague epidemics.  Unfortunately, there are no apparent genetic differences between strains that have caused indirect or direct transmission, so it will be difficult to determine from genetic analyses whether one mode or the other was dominant in any given Medieval/Renaissance epidemic.

Our growth rate estimates provide a starting point from which we can begin to explore the extents to which direct or indirect transmission played a role in the various London plague epidemics.  Current data are too sparse and limited to make precise predictions \cite{Ma+14,Park+18}, but we consider the extremes of estimated parameter ranges that would favour each mode of transmission and ask whether the resulting predictions of attack rates are consistent with the rough knowledge of epidemic sizes available from the literature.

\paragraph{In the early epoch, primarily pneumonic transmission is implausible.}

If we assume pneumonic transmission, and use generation interval estimates from 20th century outbreaks \cite{GaniLeac04} (\ie assuming that the natural history of infection in London was similar to that of Modern pneumonic plague), in the early epoch we find $\rzerop=\rzeroearlyest$ [CI (\rzeroearlylwr, \rzeroearlyupr)], which implies an attack rate of $\Zp=\Zearlyest\%$ [CI (\Zearlylwr\%, \Zearlyupr\%)] (left axes in \cref{F:combdata}).  This estimate of total mortality is the largest possible for the early-epoch epidemics (violations of any of the \hyperlink{4assumptions}{four extreme assumptions discussed in Approach} would lower the total mortality).  Total mortality of \Zearlyest\% is implausibly low in comparison to typical estimates for 14th century plague epidemics (\eg ``[b]etween the years 1346 and 1352, [plague] caused the death of \ldots\ one-third of the world's population at that time'' \cite{Ligo06}) and certainly inconsistent with the careful and comprehensive analysis of Creighton for the 1348--49 epidemic in London specifically (``the mortality was about one-half the population'' \cite[p.\,128]{Crei65}).  It is therefore unlikely that 14th century plague epidemics in London were pneumonic.

\paragraph{In the early epoch, growth rates appear to be consistent with bubonic plague.}

What if we assume bubonic transmission? Unfortunately, the expected total human mortality from a bubonic plague epidemic depends on several poorly known aspects of rat ecology. For the generation interval $\gentime$ for bubonic plague we use a rough estimate of 18 days ($\approx 0.05\,$yr) based on the flea incubation period of 9--26 days \cite{LaPerriere+2009}, which dominates the time scales of other processes involved in rat-flea plague dynamics (see \supp\ for further detail). In combination with a growth rate estimate of $r \approx \rearlyest/$yr, we obtain $\rzerob \approx \rzeroearlybest$ and $\Zb \approx \Zearlybest\%$. However, this estimate of the final size gives the fraction of rats, not humans, infected over the course of the epidemic. Translating this rat final size to the final size of the human epidemic requires us to quantify the amount of spillover from rats to humans. In particular, we need to know the rat-to-human ratio and the number of infected humans per infected rat, which can both be \emph{crudely} estimated as 1:1 \cite{KeelGill00b} (though these ratios will vary during the course of plague epidemics).  
Given an estimated CFP of 70--80\% for early epoch plagues \cite{McEv88}, our (very rough) mortality predictions are reasonably consistent with the observed human mortality rates.

\paragraph{In the late epoch, primarily pneumonic transmission is unlikely but cannot be ruled out.}

Now suppose that the late-epoch plagues were pneumonic with a generation interval similar to that of Modern pneumonic plague.  The implied reproduction number (\cref{F:combdata}) would be $\rzerop=\rzerolateest$ [CI (\rzerolatelwr, \rzerolateupr)],
which yields a maximum attack rate of \Zlateest\%  [CI (\Zlatelwr\%, \Zlateupr\%)], compared to 
the estimated $\approx 20\%$ of the population
who died in each of the late epoch epidemics \cite[p.\,4]{Cumm+16}.
Given that the CFP for (Modern) pneumonic plague is nearly 100\%
\cite{KoolWein05,WHOplaguePage,WHOplagueFactSheet,CDCplagueEmergencyFAQ},
reconciling the predicted attack rates with observed mortality would require strong effects of heterogeneity; a large fraction of individuals resistant to plague due to prior non-fatal plague infections; a large proportion of asymptomatic or sub-clinical infections; or a decrease in transmission rate as the plague spread (perhaps due to behavioural avoidance of transmission).
The effects of heterogeneity and behavioural avoidance may indeed have been stronger in the late epoch, as a result of concentration of the wealthy in central London and tendency for them to flee London during plague epidemics \cite[p.\,4]{Cumm+16}. While we still consider direct (pneumonic) transmission in  late-epoch London plague epidemics unlikely because of the strong effects that would be needed to reduce the expected mortality rates to the observed levels, we cannot rule it out.

\paragraph{In the late epoch (as in the early epoch), growth rates appear to be consistent with bubonic plague.}

Similar calculations with the estimated $r$ for the late epoch and the
approximate generation interval for bubonic plague give estimated $\rzerob=\rzerolatebest$ [CI $(\rzerolateblwr, \rzerolatebupr)$] and predicted rat attack rates of $\Zb=\Zlatebest\%$ [CI $(\Zlateblwr\%, \Zlatebupr\%)$]. The human $\Zb$ would be identical, if we make the same assumptions as above. This attack rate is even higher than the \Zlateest\% predicted for pneumonic plague, but the CFP for bubonic plague was probably only 30--60\% in the 16th-17th centuries (\eg \cite{Ligo06,WHOplagueFactSheet}), which could bring the total mortality roughly in line with the observed $\approx 20\%$. The lower CFP of bubonic plague would also lead to a build-up in the population of individuals who had survived previous plague epidemics and retained immunity (reducing the susceptible population available to be attacked), which in combination with heterogeneity and asymptomatic infections makes bubonic transmission plausible for the late epoch.

\paragraph{Other possibilities.}

As mentioned above, it is possible that human ectoparasites could have supported indirect plague transmission \cite{Dean+18,Laro+18}. While the biological details of this transmission mode are still uncertain, we can guess that the generation interval of this transmission mode would be intermediate between direct and indirect rat-flea transmission, giving rise to estimates of $\rzero$ and $Z$ intermediate between those discussed above. Investigations of plague biology in fleas have also raised the possibility that rat transmission may have shorter generation intervals than previously thought, which would lower estimated $\rzero$ values for bubonic plague \cite{Eise+06,Eise+07} (but cf.~\cite{Hinn+17}).

\subsection*{COVID-19 context}

The world is currently in the midst of a pandemic of COVID-19 disease, caused by  the SARS-CoV-2 virus \cite{WHO2020March11}. The infection fatality ratio (IFR) for COVID-19 is uncertain due to case detection limitations, but the evidence accumulated to date about this IFR \cite{MeyeLea2020,Basu2020} indicates that it is substantially lower than for the 1918 influenza \cite{Earn+02,JohnMuel02}, and much lower than for the historical plague epidemics studied here. Nevertheless, COVID-19 has had a much greater impact than seasonal influenza epidemics, which cause of order 500,000 deaths worldwide annually \cite{Earn+02,Dush+06}.

Epidemic growth rates (and doubling times) have been important elements of the COVID-19 public health discourse \cite{Park+2020}.  The quality and quantity of COVID-19 data far exceed that of the historical plague data we have analyzed in this paper.  In particular, many countries report daily counts of newly reported cases, hospitalizations and deaths. The techniques we have used to estimate initial growth of plague outbreaks are consequently easier to apply to modern data streams, and should yield more accurate estimates.  In addition, improved information about generation intervals will make it easier to convert growth rate estimates into estimates of the basic reproduction number $\R_0$.  Of course, modern data are by no means perfect, and there are still many issues --- including estimating generation intervals while accounting for disease dynamics, and assessing the importance of asymptomatic and pre-symptomatic transmission --- that can lead to overconfident or incorrect estimates of disease parameters for COVID-19 \cite{Park+2020,Park+20b,Park+20c}.

\subsection*{Conclusion}

We have estimated epidemic growth rates for all the recorded plague epidemics in London, England, during the second pandemic (1348--1666). The major plagues of the 16th and 17th centuries grew much faster than the early plagues of the 14th century when the Black Death first invaded human populations.  This conclusion is based solely on estimated growth rates---in particular, our estimates do not rely on assumptions about the mode of transmission or natural history of infection.  The cause of this substantial ``acceleration'' is currently unclear.

We considered the implications of our estimated growth rates with respect to the mode of transmission and concluded that direct transmission (pneumonic plague) was unlikely to have been the primary mode in the early epoch. Definite conclusions about indirect transmission (bubonic plague) are difficult because of the substantial uncertainties about rat and flea ecology during this period.

We have created and assembled---and made available with this paper---machine-readable databases comprising thousands of handwritten records spanning more than 300 years.  People living in London in 1348 or 1665 could not have imagined how these records might be used hundreds of years later.  The time series that have emerged have allowed us to reveal and characterize large changes in patterns of infectious disease transmission.  In concert with other forms of information \cite{Cumm+16,Abbo17}, these records should continue to contribute to a broader and more insightful view of population-level processes in early eras of history.

\hypertarget{Methods}{}
\mysectionsamepage{Methods}

\paragraph{Growth rate estimates}

We have previously developed methods to 
use mortality data to estimate initial epidemic growth rates \cite{Ma+14}.
The na\"ive approach of fitting a straight line to the relationship between logarithmic death counts and time at the beginning of the epidemic depends sensitively on the fitting window and produces overly narrow confidence intervals.  Instead, we use maximum likelihood estimation to fit a curve that resembles the shape of a deterministic epidemic model (see \hyperlink{phenmod}{Phenomenological Model} below) and compute likelihood profile confidence intervals \cite{Bolk08}; this approach yields robust estimates of initial growth rates and accurate assessments of uncertainty.  All variation around the fitted line is assumed to arise from  observation errors.  See Ma \etal \cite{Ma+14} for methodological details.

In contrast to Ma \etal \cite{Ma+14}, who assumed a Poisson error model to fit simulations that were themselves generated with that assumption, the results shown here applied a negative binomial error model, which allows for the broader range of variation found in real data.  We initially tried both logistic and Richards \cite{Rich59} models for the epidemic curve (\ie the expected mean of the data) and both Poisson and negative binomial models for the distribution around the mean. These variants gave qualitatively similar conclusions about the pattern of growth rates, but we found that the Richards model was too unstable to fit reliably across all epidemics, so we reverted to fitting the logistic model for all epidemics. Similarly, the negative binomial model gave unstable results when the distributions were too close to Poisson; when the negative binomial dispersion parameter was estimated as $> 10^4$, we reverted to a Poisson fit (see \supp\ for further details of the optimization).

In order to summarize the differences between epochs, we fitted a linear mixed model (using the \code{glmmTMB} package in R \cite{Brook+17}) to the estimates of log growth rates. To account for differing precision of the estimates across epidemics, we approximate the standard error of the estimates as the width of the 95\% confidence interval divided by 3.92 (the width in standard errors of a 95\% Normal CI); square it to find the variance; and scale the residual variance for each epidemic curve by this observed variance. This procedure is equivalent to weighting each epidemic curve by the inverse of its variance. We quantified the difference in epidemic between epochs based on the predicted difference in $\log(r)$ between wills in each epoch.

The permutation test $p$-value (0.016) reinforces our confidence that late-epoch epidemics were faster than early-epoch epidemics; however, since it is larger than the parametric $p$-values derived from the model fit (which are $< 0.001$), it suggests that the confidence intervals shown in \cref{F:combdata} may be too narrow.

In light of recent work investigating the possibility of initial epidemic growth that is sub-exponential, we also considered fits based on the more general growth model of Chowell and co-workers \cite{Vibo+16,Chow+16,ChamEarn16}.  This approach leads to slightly slower initial growth estimates, but does not change our conclusion that the later plagues were much faster than the earlier plagues.

\hypertarget{phenmod}{}
\paragraph{Phenomenological model}

Following Ma \etal \cite{Ma+14}, we model the \emph{cumulative} mortality curve (or will-accumulation curve; henceforth we refer to wills or death records as ``mortality'') using a logistic model with an added baseline linear growth rate,
%
\begin{linenomath*}
\begin{equation}
c(t) = b t + \frac{K}{\big[1 + \big((K/c_0) - 1\big)e^{-rt}\big]} \,.
\label{E:Richards}
\end{equation}
\end{linenomath*}
%
Here, $c_0=c(0)$ is the initial cumulative mortality, $r$ is the initial exponential growth rate, and $K$ is the final size of the epidemic ($\lim_{t\to\infty}[c(t)-bt]$). In the fitting procedure we use the unitless parameter $x_0=c_0/K$; all parameters are fitted on a transformed, unconstrained scale (\ie $\log(r)$, $\log(K)$, and $\textrm{arctanh}(x_0)$). The derivative $c'(t)$ is the instantaneous mortality rate.  To avoid the inherent correlations between observations in the cumulative curve, we fit the observed data to the differences from (1): \mbox{$\Delta c(t) = c(t+\Delta t)-c(t)$}, where $\Delta t$ is the observation interval (one day for wills and one week for mortality data). The parameter $b$ represents baseline (non-plague) mortality. For the LBoM fits we set $b=0$ (since the LBoM data include only plague mortality); for the wills and parish-register fits we estimated $b$ by computing the average observed mortality rate over the two years preceding and following the outbreak window.

As discussed above, we initially fitted a Richards model, which extends the logistic equation with a shape parameter, but found that we could not get stable fits to the smallest/noisiest epidemics; we never found qualitative differences between growth rates estimated from logistic and Richards fits.  Ma \etal found that a \emph{delayed} logistic curve, which adds an exponentially distributed delay between infection and mortality, estimated growth rates from mortality curves with less bias. However, the delay between infection and mortality in our data is extremely uncertain---it would represent another parameter that would have to be estimated from noisy data. For wills data, the delay could even be negative, \ie wills could have been written out of fear of \emph{future} infection. Furthermore, we found little difference between the Richards and logistic fits---which Ma \etal found to have opposite biases; hence these methods are likely to bracket the growth rate estimated by the delayed logistic.

\paragraph{Fitting windows}

How well a given model can fit a time series may depend on the time window used for fitting \cite{ChamEarn16}.  In order to avoid choosing a window separately for each individual epidemic, we considered window selection criteria that could be applied uniformly to all epidemics without first examining the data. In all cases we defined the end of the fitting window to be one data point past the observed peak date (the date of maximum observed deaths).  For wills and parish registers, we began the fitting window at the start of the outbreak window (see \hyperlink{outbreak.year}{below}); for LBoM data, we started immediately \emph{after} the last observation before the peak for which the mortality rate was $\le 1/50$ of the peak mortality rate. For two epidemics for which the data are particularly noisy (1593 and 1625), we were unable to find any simple heuristic that would automatically choose a good fitting window, and resorted to choosing the fitting window by eye.

The fitting windows used for all our fits are listed in \supp \cref{tab:windows,tab:windowsminor}. Although we experimented with different methods for determining the fitting window (e.g. using a threshold of 1/100 of the peak mortality rate, or using the observation after the last local minimum before the peak), 
we emphasize that our choices were entirely determined by seeking ``good'' fits (judged by visual inspection), and not by the pattern of estimated initial growth rates. Moreover, our conclusion that later plagues were much faster than the earlier plagues is robust to choices we made about window selection.

\hypertarget{outbreak.year}{}
\paragraph{Outbreak years and outbreak windows}

We refer to the calendar year in which an epidemic begins as the \emph{outbreak year}.  For each epidemic, we define an \emph{outbreak window} that is the maximum time interval to be considered for fitting.  For the 14th century epidemics, we defined the outbreak window to be the full outbreak year, except for outbreak year 1348 when the epidemic spanned both 1348 and 1349 (so we took the outbreak window to be the full length of both calendar years).  For the late epoch epidemics, we defined the outbreak windows as sequences of consecutive weeks during which at least one plague death was listed in the LBoM.  See \supp \cref{tab:windows,tab:windowsminor}.

\hypertarget{data.and.code}{}
\paragraph{Data and code availability}

Our \texttt{epigrowthfit} R package for estimating initial growth
rates, which includes all the data used in this paper, is available at
\url{https://github.com/davidearn/epigrowthfit}.  The data can also be
obtained from the International Infectious Disease Data Archive
(IIDDA, \url{http://iidda.mcmaster.ca}).  The additional code required
to reproduce all of our results is available at
\url{https://github.com/davidearn/plague_growth}.

\mysectionsamepage{Acknowledgements}

The plague mortality data were photographed and/or entered by Seth Earn, Kelly Hancock, Claire Lees, James McDonald, David Price and Hannah Price.  Valerie Hart facilitated work at the Guildhall Library in London.  Josephine Earn provided historical context.  Non-exponential comparison fits were conducted by David Champredon.  Sang Woo Park assisted in finding published estimates of auxiliary epidemic parameters. Alex Bushby assisted with compiling probate dates. Sigal Balshine, Ann Carmichael and David Price made helpful comments.  Early stages of this research were supported by a J.\,S.~McDonnell Foundation Research Award to DE (\url{https://www.jsmf.org/grants/d.php?id=2006014}).  All the authors were supported by Discovery grants from the Natural Sciences and Engineering Research Council of Canada (NSERC). HP was supported by the Social Sciences and Humanities Research Council of Canada (SSHRC). SHARCnet (\url{http://www.sharcnet.ca}) provided computational resources.

### 2020-06-12 library.R

@@ -1,0 +1,35 @@
+library(broom.mixed)
+library(colorspace)
+library(cowplot)
+library(dfoptim)
+library(directlabels)
+library(dplyr)
+library(emdbook)
+library(emmeans)
+library(epigrowthfit)
+library(ggplot2)
+library(ggstance)
+library(glmmTMB)
+library(grid)
+library(gtable)
+library(Hmisc)
+library(huxtable)
+library(lubridate)
+library(methods)
+library(multcomp)
+library(nloptr)
+library(optimx)
+library(plyr)
+library(purrr)
+library(RColorBrewer)
+library(tibble)
+library(tidyr)
+library(tikzDevice)
+library(viridis)
+library(xtable)
+
+stopifnot(
+    packageVersion("glmmTMB")>="0.2.2.9000" ## GH: bbolker/glmmTMB,
+  , packageVersion("bbmle")>="1.0.21.9001"  ## GH: bbolker/bbmle
+)
+

### 2020-06-12 supp.Rnw
Struggling to add things

\documentclass[12pt,table]{article}
%% knitr requires documentclass to be the first line of the file
%% Created: 19 Mar 2017
%% Changed: 22 Apr 2017
%%          30 Apr 2017
\input{nlipreamble.tex}

%% we need to cross-reference to main manuscript file:
\usepackage{xr-hyper}
\externaldocument{ms}

%%\definecolor[named]{verylightgray}{rgb}{0.95,0.95,0.95}
\definecolor[named]{verylightgray}{rgb}{0.93,0.93,0.93}

\newcommand{\mysection}[1]{\clearpage\section*{#1}\addcontentsline{toc}{section}{#1}}
\newcommand{\mysectionsamepage}[1]{\section*{#1}\addcontentsline{toc}{section}{#1}}
%% http://tex.stackexchange.com/questions/316510/hyperlink-within-caption-to-a-figure
%% if used in a caption, need short content \caption[...]{ ... \Methodslink ... }
\newcommand{\Methodslink}{\href[page=9]{ms.pdf}{\emph{Methods}}}
\newcommand\ttbackslash{{\tt\char`\\}}
\newcommand{\macro}[1]{{\tt\ttbackslash#1}}

%% put S in front of equation and figure names in \supp
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}

%% tables:
\usepackage{tabularx} % for \newcolumntype

%%\title{Faster growth of later plagues in the second millenium\\\emph{Electronic Supplementary Material}}
\title{Acceleration of plague outbreaks in the second
  pandemic\\\emph{Electronic \supp}}

\author{David J.\,D.\ Earn$^{a,c,e}$, 
Junling Ma$^{f}$, 
Hendrik N.\ Poinar$^{b,c,d,e}$,\\
Jonathan Dushoff$^{a,c,e}$,
and Benjamin M.\ Bolker$^{a,c,e}$\\
\href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
\small $^a$Department of Mathematics \& Statistics,\\
\small $^b$McMaster Ancient DNA Centre, Department of Anthropology,\\
\small $^c$Department of Biology, $^d$Department of Biochemistry,\\
\small and $^e$Michael G.\ deGroote Institute for Infectious Disease Research,\\
\small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
\small $^f$Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}

\begin{document}

\linenumbers

\null\vspace{-1.5cm}

\vbox{
\maketitle

\tableofcontents
}

\section{Historical human population estimates for London}

%
\begin{table}[h!]
\caption{Population estimates for London previously published by Finlay \cite{Finl81,FinlShea86}\cite[p.\,108]{Kryl11} and Creighton \cite[p.\,660]{Crei65}.  These data are plotted in \cref{F:london_pop}.}
\label{tab:pop}
\begin{center}
%% \input{analysis/tables/supppop.tex} % generated by analysis/tables/supppop.R
%% DE: I'm building the (somewhat different) table myself.
\renewcommand{\arraystretch}{1.0}
\begin{tabular}{c|r|l}
%%\hline
{\bfseries Year} &
{\bfseries Population} &
{\bfseries Source}\\
\hline
  1340 &  45,000 & Finlay \\ 
  1500 &  50,000 & Finlay \\ 
  1550 & 110,000 & Finlay \\ 
  1600 & 180,000 & Finlay \\ 
  1603 & 250,000 & Creighton \\
  1625 & 320,000 & Creighton \\
  1650 & 340,000 & Finlay \\ 
  1665 & 460,000 & Creighton \\
  1700 & 445,000 & Finlay \\ 
\end{tabular}
\end{center}
\end{table}
%

%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/london_pop.pdf}
\end{center}
\caption{Historical population of London, England, as estimated by Finlay \cite{Finl81,FinlShea86} (\cref{tab:pop}).  Earlier estimates of Creighton \cite[p.\,660]{Crei65} are shown for the specific plague years 1603, 1625 and 1665.}
\label{F:london_pop}
\end{figure}
%

\FloatBarrier
\section{Disease-related data sources}

\cref{tab:datasources} lists our data sources.  Below we provide a few
additional comments and details not described in the text of the main
paper.

%% requires tabularx package:
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\begin{table}[h!]
\caption[]{Disease-related data used in this paper.}
\label{tab:datasources}
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{c|c|L{3cm}|c|L{3.5cm}}
%%\hline
{\bfseries Epoch} &
{\bfseries Year range} &
{\bfseries Data type} &
{\bfseries Frequency} &
{\bfseries Source} %%&
%%{\bfseries Description}
 \\
\hline
Early & 1340--1380 & Last Wills and Testaments & Daily & Court of Husting \\\hline
Late & 1540--1680 & Last Wills and Testaments & Daily & Prerogative Court of Canterbury (PCC) \\
 & 1540--1680 & Mortality from all causes & Weekly & Parish records\\
 & 1563--1666 & Mortality from plague & Weekly & London Bills of Mortality (LBoM)
\end{tabular}
\end{center}
\end{table}

\subsection{Last wills and testaments}

\paragraph{London 1348--1375:} 
Cohn \cite{Cohn03} presents his monthly counts of wills (proved by the Court of Husting, London \cite{Shar1889}) in his Figures 7.33 (1348), 7.34 (1361) and 7.35 (1375); Cohn does not show data for 1368.  Monthly aggregations of our daily counts agree nearly perfectly with Cohn's plots for the three epidemics he studied.  Most of the slight discrepancies are probably attributable to a small number of wills that were written during the plague epidemics but probated much later (so not included by Cohn \cite{Cohn03}).

\paragraph{London 1384--1678:} The wills proved by the Prerogative Court of Canterbury (PCC) provide a substantial sample of London wills, but certainly not all the wills written by Londoners at the time, as the following quote from the National Archives website makes clear.

\begin{quote}
\hspace{-5pt}``Most people who left a will used the appropriate church court. The Prerogative Court of Canterbury was the highest church court in England and Wales until 1858, when the national court was established, but even in the late 1850s it was only proving about 40\% of the national total of 21,653 wills.

Until 1858 there were more than 200 church courts each of which kept separate registers of wills -- there was no central index.

\dots

Wills proved in the Prerogative Court of Canterbury (PCC) mainly relate to testators resident in the south of England, although all parts of England and Wales are represented in the records.''

\rightline{--~\href{http://www.nationalarchives.gov.uk/help-with-your-research/research-guides/wills-or-administrations-before-1858/}{The UK National Archives}}
\end{quote}

Before the 17th century, the number of PCC wills is too small to detect an epidemic signal (see the PCC wills plotted in the top two panels of \cref{F:plot3sourcesmajors}, which we have not used for growth rate estimates).  Consequently, studying potential plague epidemics in the 15th century---after the period when many London wills were probated in the Court of Husting and before many London wills were probated in the Court of Canterbury---would probably require analysis of the records of a substantial fraction of church courts.  Creighton \cite[Ch.\,IV]{Crei65} gives numerous historical references that indicate specific plague epidemics in London between 1375 and 1540.  Unfortunately, none of these sources provides an outbreak time series, so there is no possibility of estimating epidemic growth rates.

\subsection{Will dates vs.\ probate dates}

The Calendar of Wills probated in the Court of Husting \cite{Shar1889} is organized by court dates, so every will is associated with a definite date on which it was probated; in contrast, only 64\% of the wills provide information on the date of writing \cite{Bush19}. Despite this lower sample size and the fact that wills were in some cases written long before death, dates of writing actually provide a much more plausible representation of the epidemic patterns in the 14th century than probate dates (\cref{F:willsVSprobates}).  Graphs in the main text (\cref{F:plot3sources,F:timeseries,F:combdata}) are based on counts of wills written.  See \hyperlink{mainsec:Data.wills}{``Wills and testaments''} in the Data section of the main text for further discussion.

%
\begin{figure}[h!]
\begin{center}
  \includegraphics[width=\textwidth]{analysis/plots/probate_dates.pdf}
\end{center}
\caption{Counts of wills written vs.\ wills probated during each of
  the four 14th century plague epidemics in London, based on wills
  probated in the Court of Husting \cite{Shar1889}. Smooth lines represent loess fits
with span=0.5 (shaded regions are 95\% confidence intervals).}
\label{F:willsVSprobates}
\end{figure}
%

\subsection{London Bills of Mortality (LBoM)}

\paragraph{London 1563--1583:} The earliest weekly plague mortality records are tabulated by Creighton \cite{Crei65} for 1563--1564 \cite[p.\,305]{Crei65} and 1578--1583 \cite[p.\,341--344]{Crei65}.

\paragraph{London 1593:} The extant records of weekly plague mortality in London in 1593 (\cref{F:plot3sourcesmajors}, second panel) are implausible.  One possibility is that many earlier deaths were added to the counts for two of the weeks in July, but there is no way to be sure.  Creighton \cite[pp.\,351--360]{Crei65} discusses all the data ever found for the epidemic of plague in 1592--1593, states that the weekly data probably originate from marginal notes in a broadside of 1603, and comments (p.\,354) that ``the weekly mortalities in it for those weeks that had little plague are an absurdity for 1593.  Whatever the source of this table, it is not genuine for 1593 \dots''

\paragraph{London 1603--1680:}
Creighton \cite{Crei65} tabulates data for 1606--1610
\cite[p.\,494]{Crei65} and 1636 \cite[p.\,530]{Crei65}.  Bell \cite{Bell1665} and Creighton \cite{Crei65} both tabulate weekly mortality from 1605 to the end of 1665, which includes the majority of the Great Plague in 1665--1666 \cite[p.\,662]{Crei65}.  We entered all weekly mortality data from 1662 to 1680 directly from the London Bills of Mortality.

\subsection{Parish registers}

Cummins \emph{et al.} \cite{Cumm+16} obtained parish register data from \href{https://www.ancestry.com}{ancestry.com} and have made the weekly counts of deaths available at \url{http://neilcummins.com}.  For the 16th and 17th century epidemics, the parish data provide a third source.  For 1593, these are the only reliable data (we consider them to be reliable because they are part of a continuous weekly time series from 1538 onwards).

\subsection{Major vs minor plague epidemics}

\cref{F:all1600s.scaled} shows all weekly reports of plague deaths from the LBoM from 1563 to 1666, scaled by population size (estimated by interpolating from \cref{tab:pop}).  Epidemics with peak plague mortality above 5 per 1000 individuals per week were classified as \emph{major}.  With the exception of 1593 (for which we do not use LBoM data; see above), all these major epidemics peak above 14 on this scale; all the other (\emph{minor}) epidemics peak below 3.  The raw plague mortality data are shown in \cref{F:all1600s}.

We classify all the 14th century epidemics as `major', although we do not have appropriate data to distinguish major from minor epidemics in this early epoch.

Weekly time series for all the major epidemics are shown in \cref{F:plotwills14c,F:plot3sourcesmajors}.

%%\newcommand{\mywidth}{0.8\textwidth}
\newcommand{\mywidth}{\textwidth}
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\mywidth]{analysis/plots/plotallscaled.crop.pdf}
\end{center}
\caption{Plague deaths per 1000 persons per week, from 1563 to 1666.
Epidemics that exceeded 5 on this scale were classified as \emph{major}.
Analysis of the major epidemics is shown in \cref{tab:params,F:timeseries,F:combdata}.  Corresponding analyses of all of the minor epidemics are presented
in \cref{tab:minorparams,F:seasonalgrowth}.
}
\label{F:all1600s.scaled}
\end{figure}
%

\begin{figure}[h!]
\begin{center}
%%\scalebox{0.6}{
\includegraphics[width=\mywidth]{analysis/plots/plotall.crop.pdf}
%%}
\end{center}
\caption{All plague deaths reported in London from 1563 to 1666.}
\label{F:all1600s}
\end{figure}
%

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/plotwills14c.pdf}
\end{center}
\caption[]{Weekly numbers of wills written during plague epidemics in London, England, in the 14th century.
Vertical grey lines indicate the ``outbreak window'' specified in \cref{tab:windows}.}
\label{F:plotwills14c}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/plot3sourcesmajors.pdf}
\end{center}
\caption[]{Weekly deaths and numbers of wills written during major plague epidemics in London, England, since 1540.  For each epidemic, vertical grey lines indicate the beginning and end of the period during which deaths from plague were listed in the bills of mortality.  The left (right) scale refers to deaths (wills).}
\label{F:plot3sourcesmajors}
\end{figure}

%%\clearpage
\section{Generation interval for bubonic plague}\label{sec:gentime}

For pneumonic plague, sufficiently detailed data exist for a number of modern outbreaks to allow an estimate of the latent and infectious periods (and hence of the generation interval distribution) \cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given value of $r$. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next epidemic cycle becomes infectious is:

\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}

From \cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}

Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.

\section{Relationship between rat density and $\R_0$ (in rats)}

One factor that might have contributed to the observed increase in
epidemic growth rates in London is the density of rats.  We do not
have data that allow us to estimate rat densities in Medieval England,
but we can ask---all else being equal---by what factor would rat
density have to have changed in order to account for the observed
change in growth rates?

In the idealized situation in which the generation interval is exponentially distributed (as in the standard SIR model) then \cite{WallLips07}, as mentioned in the Discussion in the main text, 
%
\begin{linenomath*}
\begin{equation}\label{E:rofR}
r = \frac{\R_0-1}\gentime \,,
\end{equation}
\end{linenomath*}
%
where $\gentime$ is the mean generation interval.  Consequently, if this simple relationship holds and the mean generation interval \emph{does not change}, then a change in growth rate $r_1\to r_2$ implies a change in basic reproduction number $\R_{0,1}\to\R_{0,2}$, where
%
\begin{linenomath*}
\begin{equation}\label{E:1to2}
\frac{\R_{0,2}}{\R_{0,1}} = \frac{r_2\gentime+1}{r_1\gentime+1} \,.
\end{equation}
\end{linenomath*}
%
If we assume the estimate of $\gentime=18$ days obtained in
\cref{sec:gentime} above, and the MLE growth rates from early and late wills
listed in \cref{tab:params}, the relative change in reproduction
number that needs explanation is
<<rat_R0_calc,echo=FALSE>>=
ratR0 <- function(r){ (r / 365 * 18) + 1 }
load("analysis/fits/.epochsum.RData")
r1 <- rvals["fit","early"]
r2 <- rvals["fit","late"]
R01 <- ratR0(r1)
R02 <- ratR0(r2)
R0ratio <- R02/R01
@ 
%
\begin{linenomath*}
\begin{equation}\label{E:Rratio}
\frac{\R_{0,2}}{\R_{0,1}} 
\approx
\frac
{\Sexpr{round(r2,1)} \cdot \frac{18}{365} + 1}
{ \Sexpr{round(r1,2)} \cdot \frac{18}{365} + 1}
\approx
\frac{\Sexpr{round(R02,2)}}{\Sexpr{round(R01,2)}} 
\approx \Sexpr{round(R0ratio,2)}
\,.
\end{equation}
\end{linenomath*}
%
where the second subscript (1 or 2) denotes the early or late epoch, respectively.

<<KGparams,echo=FALSE,cache=FALSE>>=
a <- 4e-3
KR1 <- 250
KGfac <- 1 - exp(-a*KR1)
@

Keeling and Gilligan \cite[p.\,2226]{KeelGill00b} relate the basic
reproduction number to rat density via
%
\begin{linenomath*}
\begin{equation}\label{E:KGR0}
\R_0 = \frac{\beta_{\rm R} K_{\rm F}}{d_{\rm F}}\Big[
1 - \exp{(-aK_{\rm R})} \Big] \,,
\end{equation}
\end{linenomath*}
%
where $\beta$ and $K$ denote transmission rate and carrying capacity, respectively, and the subscripts F and R denote fleas and rats, respectively.  \cref{E:KGR0} implies
%
\begin{linenomath*}
\begin{equation}\label{E:R0ratio}
\frac{\R_{0,2}}{\R_{0,1}} = 
\frac{1 - \exp{(-aK_{{\rm R},2})}}{1 - \exp{(-aK_{{\rm R},1})}} \,,
\end{equation}
\end{linenomath*}
%
which, for $aK_{{\rm R},j}\ll1$, simplifies to
%
\begin{linenomath*}
\begin{equation}\label{E:aKsmall}
\frac{\R_{0,2}}{\R_{0,1}}
\;\approx\; 
\frac{K_{{\rm R},2}}{K_{{\rm R},1}}
\quad.
\end{equation}
\end{linenomath*}
%
The approximation (appropriate in the limit $aK\to0$) provides a lower
bound to the true relationship, indicating that rat density would have
had to increase by at least a factor of  to account for a similar
increase in $\R_0$.

\cref{F:KvsR} shows the exact relationship \eqref{E:R0ratio} for
several values of $aK_{{\rm R},1}$, together with the approximation
for small $aK$ \eqref{E:aKsmall}.  Keeling and Gilligan \cite[Table 1,
p.\,2221]{KeelGill00b} adopt values of $a$ and $K_{{\rm R}}$ that
yield $aK_{{\rm R}}=10$.  If this is the correct order of magnitude
for $aK_{{\rm R},1}$ then \cref{E:R0ratio} implies that no increase in
rat density would be sufficient to yield an increase in $\R_0$ by a
factor of \Sexpr{round(R0ratio,1)} (or any factor detectably greater than 1).

%
\begin{figure}[h!]
\begin{center}
 \includegraphics[width=\textwidth]{analysis/plots/KvsR.pdf}
\end{center}
\caption{Relative change in rat carrying capacity $K_{\rm R}$ as a
  function of relative change in basic reproduction number $\R_0$.
  Solid curves are based on \cref{E:R0ratio}.  The dotted red line is
  based on \cref{E:aKsmall}.  The \Sexpr{round(R0ratio,2)}-fold increase in $\R_0$ that
  requires explanation is indicated in yellow.}
\label{F:KvsR}
\end{figure}
%

\input{analysis/fits/modeldefs.tex}
\section{Effects of rat ecology on growth rates}

In the Discussion in the main text, we listed ``ecological and
demographic changes'' as a possible cause of acceleration of plague
epidemics.  Could changes in rat ecology plausibly have contributed to
changes in epidemic growth rates?

As a vector-borne disease, the rate of bubonic plague spread is primarily affected by the flea-rat ratio.
In the traditional Ross-MacDonald model for vector-borne epidemics, $\rzero$ is proportional to the square of vector-host ratio because vectors must independently bite hosts twice (once to become infected and a second time to infect a susceptible host). In the rat-flea model, $\rzero$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
From Keeling and Gilligan's plague model \cite{KeelGill00b}, the expected change in $r$ for a change in flea-rat ratio from $K_{F0}$ to $K_F$ is $(K_F/K_{F0})^S$, where $S$ is the sensitivity ($\approx 1.5$). Thus in order to a see a fivefold increase in $r$ we would need a change of $5^{(1/S)} \approx \Sexpr{round(5^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.

As noted in the main text, rat density has an additional, indirect effect on growth rate $r$ and reproduction number $\rzero$: increasing rat density will increase the probability that fleas leaving dying rats will find new, susceptible rat hosts. To crudely estimate the magnitude of this effect, if we again consider Keeling and Gilligan's model \cite{KeelGill00b} then the maximum possible effect of rat density on $\rzero$ by this mechanism would be proportional (if fleas have very low success in finding new rat hosts). That is, in order for $\rzerob$ to increase from $\approx \rzeroearlybest$ in the 14th century to $\approx \rzerolatebest$ in the late epoch, rat density would have to increase by at least $\ratpct\%$. If fleas were already fairly successful at finding hosts in the 14th century, then changes in rat density would be expected to have only a small effect on $\rzero$ in rats.

\section{Analysis of delays between wills and other sources}

\input{analysis/fits/wills_delay_defs.tex}

In order to determine the relative timing of epidemics recorded by different sources, we interpolated the parish data (which are recorded at different times from the LBoM) to find values corresponding to the LBoM dates. We quantify the relative timing in two different ways, (1) measuring the maximum cross-correlation and the lag at which this cross-correlation occurs, and (2) measuring the time difference between the epidemic peaks. As expected, the parish records are strongly correlated with LBoM (maximum correlation \parishccfmaxmin--\parishccfmaxmax) and approximately synchronous (CCF lag \parishccflagmin--\parishccflagmax\ weeks, peak lag \parishpeaklagmin--\parishpeaklagmax\ weeks). The wills data are more weakly correlated (maximum correlation \willsccfmaxmin--\willsccfmaxmax) and more delayed (CCF lag \willsccflagmin--\willsccflagmax\ weeks, peak lag \willspeaklagmin--\willspeaklagmax\ weeks).

%% analysis/fits/wills_delay.R

\begin{table}[h!]
\caption{Observed cross-correlation and delay between London Bills of
  Mortality (LBoM) records and other sources. All lags measured in
  weeks.  Rows corresponding to wills data are highlighted in grey.}
\begin{center}
  \input{analysis/tables/wills_delay_tab.tex}
\end{center}
\label{tab:delays}
\end{table}

\section{Details of numerical optimization}

Robustly fitting phenomenological epidemic curves to small, noisy data
sets proved to be surprisingly challenging\footnote{See ``monsters in
  the basement'',
  \url{https://redpenblackpen.tumblr.com/post/145304820562/monsters-in-the-basement}.}. In
the course of developing the full model we experimented with a range
of optimizers --- specifically the BFGS and Nelder-Mead options for
R's \code{optim} function as well as R's \code{nlminb} function; we
also tried an approach that iterated back and forth between
Nelder-Mead and BFGS until convergence was achieved or the fit was
sufficiently stable. We chose \code{nlminb} because it gave the best
results (highest log likelihood) for point estimates. In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, but we found that the scaled-tangent link proposed by \cite{Ma+14} ($\eta = \tan((x/2-1) \cdot \pi/2)$) declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).

With this approach we failed to achieve good (convergent) fits for only two combinations of source and outbreak year, both for minor outbreaks (London bills for 1578, Canterbury wills for 1581); these cases are excluded from the tables and figures shown below.

In the future, we would suggest that further work on reparameterization, including parameterizing the logistic by the time at which half of the final size is achieved rather than by the initial number infected, and possibly the reparameterization and regularization methods suggested by \cite{SmirChow17} for the Richards model, would be useful.

\FloatBarrier
%%\clearpage
\section{Supplementary tables cited in the main text}

\begin{table}[h!]
    \cprotect\caption{Summary statistics for the model estimating differences across epochs. All parameters are in units of log(growth rate)/year.
      The model includes fixed effects of epoch (early [14th c.]
      vs. late [16th - 17th c.]) and source (wills, parish, London
      Bills of Mortality) and a random effect of outbreak
      year. Variability for each observation is assumed to be
      proportional to the uncertainty in its $\log(r)$ estimate (see
      main text, \emph{Growth rate estimates}). 
      In Wilkinson-Rogers notation, the model formula is: \verb!log.r ~ epoch + source + (1|outbreak.year), disp= ~ 1+offset(log(sdvals^2))!. Parameter estimates are given on the $\log(r)$ scale;
      Wald confidence intervals are given in parentheses.
}  \begin{center}
    \input{analysis/tables/epochsum.tex}
  \end{center}
\label{tab:summarystats}
\end{table}

\begin{table}
  \caption[Minor outbreak parameter estimates]{Parameter estimates for minor plague epidemics. Plague epidemics with $\hat r<1/\mathrm{yr}$ or $\hat r>100/\mathrm{yr}$, representing unreliable fits, are excluded; confidence limits that are $<0.1$ or $>200$ are replaced with {\tt NA}. See \Methodslink.}
  \footnotesize
  \rowcolors{2}{white}{verylightgray}
  \begin{center}
    \input{analysis/tables/minorparams.tex}
  \end{center}
  \label{tab:minorparams}
\end{table}

\begin{table}
  \caption[Fitting windows]{Outbreak windows and fitting windows used in fitting the major plague epidemics.  See \Methodslink.}
  \rowcolors{2}{white}{verylightgray}
  \input{analysis/tables/suppwindows.tex}
  \label{tab:windows}
\end{table}

\begin{table}
  \caption[Fitting windows (minors)]{The start and end dates for the fitting windows used in fitting the minor plague epidemics.  See \Methodslink.}
  \scriptsize
  \rowcolors{2}{white}{verylightgray}
  \begin{center}
    \input{analysis/tables/suppwindowsminor.tex}
  \end{center}
  \label{tab:windowsminor}
\end{table}

\FloatBarrier
\section{Supplementary figure cited in the main text}

%
\begin{figure}[h!]
\begin{center}
\scalebox{0.9}{
  \includegraphics{analysis/plots/seasonalgrowth.pdf}
}
\end{center}
\caption{Growth rate estimates for all the late epoch (1540--1680) London plague epidemics.  The growth rates shown for the major epidemics are the same as those in the right panel of \cref{F:combdata}. 
}
\label{F:seasonalgrowth}
\end{figure}
%

%%\clearpage
%%\bibliographystyle{vancouver}
%%\bibliographystyle{Science}
\bibliographystyle{pnas-new}
\bibliography{plague}
\addcontentsline{toc}{section}{References}

\end{document}

### 2020-06-12 analysis/fits/comb.R
Main MS; nothing pulled in except by chaining (there should be few or no excess files)

@@ -1,0 +1,87 @@
+## combine data and fits
+output = ".comb.RData"
+
+library(dplyr)
+library(tidyr)
+library(methods) # needed for as(); needs to go *before* epigrowthfit!
+library(epigrowthfit)
+
+## testing GOF
+## summary(fitList[[2]][[1]])
+## trace("gof",sig="epigrowthfit",browser)
+## summary(fitList[[2]][[1]],gof_agg="7 days")
+
+ss_majors <-(
+    summary.fitList(fitList,level=2)
+    %>% drop_bad
+    ## convert source labels to match previous fits (lump wills together)
+    %>% mutate(source= mod_source(source))
+    %>% select(source,outbreak.year,starts_with("growthrate"),gof)
+    %>% add_epoch
+)
+## summary(fitList[["London bills"]][["1593"]])
+## data (time series of wills/plague deaths)
+dd_majors <- (summary.fitList(fitList,level=2,type="deaths")
+    %>% drop_bad
+    %>%  mutate(source=mod_source(source))
+)
+
+## model predictions
+pp_majors <- (summary.fitList(fitList,level=2,type="pred", confint_pred=FALSE)
+    %>% drop_bad
+    %>%  mutate(source=mod_source(source))
+)
+
+if (FALSE) {
+    ## issues with 1593? which source?
+    ## testing
+    library(ggplot2)
+    ggplot(pp_majors,
+           ## subset(pp_majors,as.numeric(outbreak.year)<1400),
+           aes(time,deaths,ymin=lwr,ymax=upr))+
+        geom_line(aes(colour=source))+
+        geom_ribbon(aes(fill=source),alpha=0.3)+
+        facet_wrap(~outbreak.year,scale="free")+
+        scale_y_log10()
+    predict(fitList[["Canterbury wills"]][["1625"]],confint=TRUE)
+    pp1 <- filter(pp_majors ,outbreak.year=="1")
+    debug(bbmle:::pop_pred_samp)
+    debug(emdbook::dmvnorm)
+    debug(MASS::mvrnorm)
+    trace("predict",sig="epigrowthfit",browser)
+
+}
+
+pp_wills_7day <- pp_majors %>%
+    filter(source=="wills") %>%
+    ## keep source in group_by to retain it in results
+    group_by(outbreak.year,source) %>%
+    do(aggsum(time=.$time,deaths=.$deaths,period="7 days")) %>%
+    ungroup
+
[…]

### 2020-06-12 analysis/fits/doFits.R

@@ -1,0 +1,114 @@
+library(epigrowthfit)
+library(optimx)
+library(dfoptim)
+
+## Default arguments
+def_args <- list(model="logistic",
+                 max_theta=1e4,
+                 epitype="major",
+                 optimizer="nlminb",
+                 optimfun=NULL,
+                 optCtrl=list(eval.max=1000,iter.max=1000),
+                 method="Nelder-Mead", ## ignored unless optimizer="optim"
+                 x0link="atan",
+                 start_list=NULL
+                 )
+
+for (i in seq_along(def_args)) {
+    n <- names(def_args)[i]
+    if (!exists(n)) assign(n,def_args[[i]])
+}
+
+## epitype should be set upstream (by make) or else in arguments
+
+if (!exists("epitype")) stop("epitype must be defined")
+cat(sprintf("fitting %s epidemics (raw data)\n",epitype))
+
+options(epigrowthfit.model=model)
+
+## set window_defs to beginning of outbreak year, except 1593 and 1625
+##  (hand-set)
+window_defs <- read.csv("window_defs.csv")
+window_starts <- with(window_defs,
+	ifelse(outbreak.year %in% c(1593,1625),window.start,start)
+)
+
+## fit.epidemics() depends on *NAMES* of first= argument
+names(window_starts) <- window_defs$outbreak.year
+
+london_bills_sub = dplyr::filter(london_bills
+	, (severity == epitype)
+	, (outbreak.year<=1665)
+)
+canterbury_wills_sub = dplyr::filter(canterbury_wills, severity == epitype)
+london_parish_sub = dplyr::filter(london_parish, severity == epitype)
+
+options(epigrowthfit.optimizer=optimizer,
+        epigrowthfit.optCtrl=optCtrl,
+        epigrowthfit.method=method,
+        epigrowthfit.optimfun=optimfun)
+
+## slightly different settings for each data set (ugh)
+mnoconst <- get_model(model,link_vals=c(x0=x0link))
+mconst <- get_model(model,add_const=TRUE,link_vals=c(x0=x0link))
+## London bills: plague deaths only, no need for baseline
+## (but use data-driven window start rather than starting from
+##   beginning of outbreak year)
+LBOM_arglist <- list(
+    max_theta=max_theta,
+    drop_mle2_call=FALSE,
+    model=mnoconst)  ## can use mostly defaults
+## Canterbury wills data: not aggregated, all-cause mortality
+## change data, model; add baseline, window starts, deaths_var
+CW_arglist <- LBOM_arglist
[…]

### 2020-06-12 analysis/fits/epidemics.R

@@ -1,0 +1,268 @@
+library(methods)
+library(emdbook)
+library(epigrowthfit)
+library(dplyr)
+
+## default is warn=0.  warn=1 gives 'real time' warnings rather than
+##  saving warnings and printing ">50 warnings occurred ..."
+## good for debugging, but right now we have lots of hard-to-suppress
+##  warnings emanating from lbeta ("underflow occurred in 'lgammacor'")
+## options(warn=1)
+
+# pick the subset of data corresponding to an epidemic
+#' @param epidemic the epidemic defined by set.epidemic
+#' @param data a dataframe with columns time and plague.deaths
+pick.epidemic <- function(epidemic, data) {
+  subset(data, outbreak.year == epidemic)
+}
+
+#' aggregate a time series of deaths into monthly deaths
+#' @param data the epidemic defined by set.epidemic
+aggregate.monthly <- function(data) {
+  if (nrow(data) == 1) return(data)
+  stack = data[1,]
+  push <- function(row) {
+    if (row$time - stack$time > 27/366) { ## FIXME: what does 27/366 signify?
+      summed = stack
+      stack = row
+      return(list(stack=stack, pop=summed))
+    } 
+    stack$plague.deaths = stack$plague.deaths + row$plague.deaths
+    list(stack=stack, pop=data.frame())
+  }
+  aggregated = data.frame()
+  # only aggregate to monthly if we have more than 1 row of data
+  for (i in 2:nrow(data)) {
+    out = push(data[i, ])
+    stack = out$stack
+    aggregated = rbind(aggregated, out$pop)
+  }
+  rbind(aggregated, stack)
+}
+
+extract_baseline <- function(data, yr, unit=c("year","step"),
+                             debug=FALSE) {
+    unit <- match.arg(unit)
+    ## cases per reporting step ...
+    bb <- attr(data,"baseline")[as.character(yr)]
+    if (debug) cat("baseline (1): ",bb,"\n")
+    if (unit=="year") {
+        ## time step ...
+        dt <- diff(data$time)[1]
+        if (debug) cat("dt (1): ",dt,"\n")
+        ## -> cases/year
+        bb <- bb/dt
+        if (debug) cat("baseline (2): ",bb,"\n")
+    }
+    return(bb)
+}
+
+#' initialize fitEpidemic
+#' @param data the plague data, with columns time and plague.deaths and epidemics
+#' @param first optional named list of hard-coded parameters for window start
+#' (as index of position within outbreak year); if specified as a single value,
[…]

### 2020-06-12 analysis/fits/epochsum.R

@@ -1,0 +1,149 @@
+library(methods)
+
+library(glmmTMB)
+if (packageVersion("glmmTMB")<"0.2.2.0") {
+    stop("need latest version of glmmTMB: ",
+         "devtools::install_github('glmmTMB/glmmTMB/glmmTMB')")
+}
+library(epigrowthfit)
+library(emmeans)
+library(multcomp)
+
+## predicted values for wills
+dd <- data.frame(epoch=c("early","late"),source="wills",
+                 outbreak.year=NA,sdvals=NA)
+logrvals <- predict(f0,newdata=dd,se.fit=TRUE)
+
+e1 <- emmeans(f0, ~epoch)
+##
+## est df
+ddf <- summary(e1)$df[1]
+ciwid <- qt(0.975,ddf) ## 2.22
+logrtab <- with(logrvals, rbind(fit, lower=fit - ciwid*se.fit,
+                                upper=fit + ciwid*se.fit))
+colnames(logrtab) <- c("early","late")
+
+## The following three tables contain est,lower,upper for each
+## epoch (early or late), based solely on wills:
+rvals <- exp(logrtab)
+R0vals <- R0(rvals)
+Zvals <- finalsize(R0vals)
+
+## crude estimates based on generation time of 18 days = 0.05 year
+R0valsb <- rvals*0.05+1
+Zvalsb <- finalsize(R0valsb)
+
+## R0 ests based on all sources
+R0all <- R0(exp(as.matrix(summary(e1)[,c("emmean","lower.CL","upper.CL")])))
+## match R0vals format
+R0all <- t(R0all)
+dimnames(R0all) <- dimnames(R0vals)
+
+Zall <- finalsize(R0all)
+
+ci_ur <- suppressWarnings(confint(f0,method="uniroot"))
+
+## profile CIs slightly wider
+## cc <- exp(confint(f0))
+## v <- setNames(round(cc["cond.epochlate",],1),c("lwr","upr","est"))
+
+## use t-stat CIs; makes 'v' and 'vall' equivalent
+csd <- summary(f0$sdr, "fixed")[, "Std. Error"]
+csd <- csd[names(csd)=="beta"]
+pp <- fixef(f0)[["cond"]]
+w <- which(names(pp)=="epochlate")
+v <- round(exp(c(est=pp[[w]],
+                 lwr=pp[[w]]-ciwid*csd[[w]],
+                 upr=pp[[w]]+ciwid*csd[[w]])),1)
+
+## compute difference based on all sources
[…]

### 2020-06-12 analysis/fits/permtest.R

@@ -1,0 +1,68 @@
+library(methods)
+library(dplyr)
+
+library(epigrowthfit)
+library(glmmTMB)
+ss <- (summary.fitList(fitList,level=2)
+    ## log-transform ... and rename ...
+    %>% mutate_at(vars(starts_with("growthrate")),log)
+    %>% dplyr::rename(log.r=growthrate.value,
+                      log.r.min=growthrate.lower,
+                      log.r.max=growthrate.upper)
+    %>% select(outbreak.year,source,starts_with("log.r"))
+    %>% add_epoch_EL()
+    %>% mutate(sdvals=(log.r.max-log.r.min)/diff(qnorm(c(0.025,0.975))))
+    %>% drop_bad ## no Canterbury/1563,1593; LBoM/1593
+    ## lump canterbury/husting (*after* drop_bad!)
+    %>% mutate(source=mod_source(source))
+)
+
+f0 <- fitfun(ss,method="glmmTMB")
+s0 <- p_sumfun(f0)
+permtab <- allperms(ss)
+obsval <- s0$estimate  ## observed log-difference in growthrates
+hist(permtab$estimate,col="gray",breaks="FD")
+## pvalue <- mean(abs(obsval)<=abs(permtab$estimate))
+perm_count1 <- sum(obsval<=permtab$estimate) ## 1-sided, directional
+perm_pvalue1 <- mean(obsval<=permtab$estimate) ## 1-sided, directional
+perm_count <- 2*sum(obsval<=permtab$estimate) ## 1-sided, directional
+perm_pvalue <- 2*perm_pvalue1
+
+## predictions (no re.form for glmmTMB)
+
+pframe <- (with(ss,expand.grid(epoch=unique(epoch),
+                     source=unique(source),
+                     outbreak.year=NA,
+                     sdvals=1)))
+pred0 <- predict(f0,newdata=pframe,se.fit=TRUE)
+
+preds <- (pframe
+    %>%   mutate(growthrate.value=pred0$fit,
+                 growthrate.lower=pred0$fit-1.96*pred0$se.fit,
+                 growthrate.upper=pred0$fit+1.96*pred0$se.fit)
+    %>% mutate_at(vars(starts_with("growthrate")),exp)
+    %>% mutate(doublingtime.value=doublingTime(growthrate.value),
+               ## NB lower/upper switched here (inverted scale)
+               doublingtime.lower=doublingTime(growthrate.upper),
[…]

### 2020-06-12 analysis/fits/set_nlminb_major.R

@@ -1,0 +1,7 @@
+## default optimizer (uses nlminb)
+model <- "logistic"
+max_theta <- 1e4
+epitype <- "major"
+optimizer <- "nlminb"
+optCtrl <- list(eval.max=1000,iter.max=1000)
+

### 2020-06-12 analysis/fits/simfuns.R

@@ -1,0 +1,216 @@
+library(dplyr)
+library(purrr)
+library(epigrowthfit)
+
+## extract 'sampling' design from data; which outbreak years
+##  do we have which sources for?
+
+get_oy <- function(x,src) {
+    if ("severity" %in% names(x)) {
+        x <- filter(x,severity=="major")
+    }
+    x2 <- (x
+        %>% select(outbreak.year)
+        %>% unique()
+        %>% mutate(source=src)) 
+    return(x2)
+}
+plague_years <-
+    (map2(list(london_parish,london_bills,husting_wills,canterbury_wills),
+         c("parish","LBoM","wills","wills"),
+         get_oy)
+        %>% bind_rows()
+        %>% mutate(epoch=factor(ifelse(outbreak.year<1400,"early","late")))
+        %>% arrange(outbreak.year)
+    )
+
+true_pars <- list(
+    ## fixed effects (intercept; early vs late; wills vs. parish, LBoM)
+    beta=c(`(Intercept)`=2.5,epochlate=0.5,sourceparish=0.25,sourceLBoM=0.5), 
+    sigma_yr=0.5,  ## among-year sd
+    sdvals=c(mu=-1,sd=0.5) ## log-normal params for residual error
+)
+## reorganize fixed effects into data frame format for later merging/analysis
+true_vals <- tibble(term=names(true_pars$beta),
+                    true_val=true_pars$beta)
+
+
+simfun <- function(epochlate=true_pars$beta["epochlate"], ## early/late diff
+                   beta=true_pars$beta,         ## all fixed parameters
+                   sigma_yr=true_pars$sigma_yr, ## among-year var
+                   sdvals=true_pars$sdvals,     ## log-mean/sd of residual sd
+                   data=plague_years,
+                   checkpoint=FALSE) {
+    res <- data
+    n <- nrow(data)
+    ## substitute specified value ...
+    beta["epochlate"] <- epochlate
+    ## could use lme4::simulate(), but it doesn't work very well
+    ## with variable variances, so ...
+    ## construct model matrices for fixed effects and random effects
+    X <- model.matrix(~epoch + source,  data=data)
+    Z <- model.matrix(~factor(outbreak.year)-1, data=data)
+    ## year effects
+    b <- rnorm(ncol(Z),mean=0,sd=sigma_yr)
+    ## residual SDs
+    sdvals <- rlnorm(n,
[…]

### 2020-06-12 analysis/plots/3axes.R

@@ -1,0 +1,131 @@
+#' testing 3-axis solutions
+#' cf. cowplot::axis_canvas,
+library(cowplot)
+library(gtable)
+#' @examples
+#' set.seed(101)
+#' dd <- data.frame(x=rnorm(20),y=rnorm(20))
+#' library(ggplot2)
+#' gg0 <- ggplot(dd)
+#' g1A <- gg0 + geom_point(aes(x,y))
+#' g1B <- gg0 + geom_point(aes(x,10*y))
+#' g1C <- gg0 + geom_point(aes(x,10*y)) + geom_vline(xintercept=-Inf)
+#' g12 <- combine_axes(g1A,g1C,add_pos="l")
+#' print(g12)  ## only need plot() if we return a gtable
+#' g12T <- combine_axes(g0,g1,add_pos="l",return_gtable=TRUE)
+#' g123T <- combine_axes(g12T,g1,add_pos="l")
+#' g12L <- combine_axes(g2,g1,add_pos="l",pad_inner=5)
+#' g12LS <- combine_axes(g2,g1,add_pos="l",pad_inner=3)
+#' print(g12L)
+#' @param extra_plot ggplot with elements to extract & paste
+#' @param base_plot ggplot or gtable to extend
+#' @param pattern regexp of elements to extract and attach; if not specified, look for axis and ylab on the plot side corresponding to \code{add_pos}
+#' @param add_pos which side to add elements to?
+#' @param tag ??
+#' @param pad ??
+#' @param pad_inner ??
+#' to right-hand side (at the moment) of base_plot
+combine_axes <- function(base_plot,extra_plot,
+                         pattern=NULL,
+                         add_pos="r",
+                         tag="1",
+                         pad=NULL,
+                         pad_inner=NULL,
+                         debug=FALSE,
+                         return_gtable=FALSE,
+                         xtheme=NULL) {
+    if (is.null(pattern)) {
+        pattern <- sprintf("(axis-%s|ylab-%s)",add_pos,add_pos)
+    }
+    if (!inherits(base_plot,"gtable")) {
+        x0 <- cowplot::plot_to_gtable(base_plot)
+    } else {
+        x0 <- base_plot
+    }
+    ## add space before axis
+    if (!is.null(pad_inner)) {
+        x0 <- gtable_add_cols(x0,
+                   pos=if (add_pos=="l") 0 else -1,
+                   widths=grid::unit(pad_inner,"cm"))
+    }
[…]

### 2020-06-12 analysis/plots/combdata.R

@@ -1,0 +1,295 @@
+library(epigrowthfit)
+library(ggplot2); theme_set(theme_bw())
+library(dplyr)
+library(directlabels)
+library(tikzDevice)
+library(cowplot)
+library(gtable)
+library(glmmTMB)
+
+## plot growth rate results from various different data sources
+args <- commandArgs(trailingOnly=TRUE)
+
+## args <- NULL ## interactive/debugging
+defaults = list(output="",width=7,height=4,size=0.5,incl_ref=FALSE,
+                ci_type="pointrange")
+
+texname <- paste0(rtargetname, ".tex")
+
+output = if (length(args) < 1) texname else args[1]
+width = if (length(args) < 2) defaults[[2]] else as.numeric(args[2])
+height = if (length(args) < 3) defaults[[3]] else as.numeric(args[3])
+size =  if (length(args) < 4 ) defaults[[4]] else as.numeric(args[4])
+incl_ref = if (length(args) < 5 ) defaults[[5]] else as.logical(args[5])
+ci_type = if (length(args) < 6 ) defaults[[6]] else args[6]
+
+hack_labs <- TRUE
+
+cat(output,width,height,size,"\n")
+
+## modify widths of specified columns
+## (not used: messes up ggplot format)
+hack_widths <- function(g,width=1,unit="cm",pattern="axis-t-[23]") {
+    gt <- plot_to_gtable(g)
+    inds <- grep(pattern,gt$layout$name)
+    n <- length(inds)
+    if (length(width)<n) width <- rep(width,length.out=n)
+    cols <- gt$layout$l[inds] ## assume 1 col per element
+    gt$widths[cols] <- grid::unit(width,rep(unit,n))
+    return(ggdraw(gt))
+}
+
+pp <- preds %>% switch_epoch()
+
+if (ci_type=="pointrange") {
+    ## modify epoch-summary data to add to outbreak-level data
+    levels(pp$epoch) <-  c("early","late")
+    ## FIXME: fragile, depends on order
+    ## was: paste(levels(pp$epoch),"\n(sum)")
+    pp <- (pp
+        ## growthrates now exponentiated upstream, in test.Rx
+        ##    %>% mutate_at(vars(starts_with("growthrate")),exp)
+        ## epoch now == (late, late, early, late)
+        ## rather than (14th c., 16th-17th c.)
+        %>% mutate(outbreak.year=as.character(epoch))
+    )
+    ss_majors <- (pp
[…]

### 2020-06-12 analysis/plots/plot.R

@@ -1,0 +1,297 @@
+library(ggplot2)
+stopifnot(packageVersion("ggplot2")>="2.2.0")
+library(RColorBrewer)
+require(dplyr)  ## for mutate(), filter()
+
+months.name = c("Jan", "Apr", "Jul", "Oct")
+days <- c(0, 31, 59, 90, 120, 151, 181, 212, 243, 273, 304, 334, 365)
+
+
+theme.fit <- (theme_bw() + ## black and white rather than grey default background
+           theme(panel.spacing=grid::unit(0,"lines"), ## squash panels together
+                 axis.text.y = element_text(size=6),  ## smaller y axis tick labels
+                 strip.text.y = element_text(angle=0), ## horizontal strip labels
+                 ## get rid of grid lines
+                 ## panel.grid.major.x=element_blank(),
+                 panel.grid.minor.x=element_line(linetype="dotted",colour=grey(0.8)),
+                 panel.grid.major.y=element_blank(),
+                 panel.grid.minor.y=element_blank()
+           )
+)
+
+#' figure for death fit
+#' @param fit fitted object
+#' @importFrom ggplot2 ggplot aes facet_grid scale_x_continuous scale_y_continuous geom_line geom_point
+#' @export
+figure.deaths.fit <- function(fit,fake_data=TRUE) {
+    ## only if more than one type plotted on the figure
+    bp3 <- brewer.pal(3,"Set1")
+    myblue <- bp3[2]
+    myred <- bp3[1]
+    ## set point defaults so that single-class version will work correctly
+    update_geom_defaults("point", list(fill=myred, shape=21, colour="black" ))
+    fitted_plus_obs <- length(unique(fit$type))>1
+    if (fitted_plus_obs) {
+        fit$type <- factor(fit$type,levels=c("raw","fitted"),
+                           labels=c("observed","fitted"))
+        figure <- ggplot(data = fit, aes(x=time, y=deaths, fill=type,
+                                         shape=type, color=type)) +
+            scale_fill_manual(name="",values=c(myred,NA))+
+            scale_color_manual(values=c("black","blue"),name="") +
+            scale_shape_manual(name="",values=c(21,NA))
+    } else {
+        ## single type (no colour, shape aesthetics)
+        figure <- ggplot(data = fit, aes(x=time, y=deaths))
+    }
+    if (fake_data) {
+        ## ugly, ugly, ugly
[…]

### 2020-06-12 analysis/plots/plot_all.R

@@ -1,0 +1,142 @@
+## show bills, wills and parish registered so we can see the major
+## epidemics are evident simultaneously no matter what source we
+## examine
+args <- commandArgs(trailingOnly=TRUE)
+
+texname <- paste0(rtargetname, ".tex")
+defaults = list(file="",standAlone=TRUE,width=6.5,height=3)
+
+file = if (length(args) < 1) texname else args[1]
+standAlone = if (length(args) < 2) defaults$standAlone else as.logical(args[2])
+width = if (length(args) < 3) defaults$width else as.numeric(args[3])
+height = if (length(args) < 4) defaults$height else as.numeric(args[4])
+
+cat(width,height,"\n")
+
+library(epigrowthfit)
+library(dplyr)
+library(lubridate)
+library(tikzDevice)
+library(ggplot2)
+theme_set(theme_bw())
+
+tex_inputs <- Sys.getenv("TEXINPUTS")
+Sys.setenv(TEXINPUTS = paste(getwd(), tex_inputs, sep = .Platform$path.sep))
+tikz("crown.tex")
+
+aggWeeks <- 4
+aggUnit <- paste(aggWeeks,"weeks")
+myylab <- if (aggUnit=="1 week") "Weekly counts" else paste(aggWeeks, "-weekly counts")
+
+h_wills <- aggregate_wills(husting_wills_individual,
+                         include_all=TRUE, aggregation=aggUnit)
+c_wills <- aggregate_wills(canterbury_wills_individual,
+                           include_all=TRUE, aggregation=aggUnit)
+
+all_wills <- bind_rows(h_wills,c_wills) %>%
+    select(date,time,nwills,outbreak.year) %>%
+    rename(deaths="nwills")
+
+LBoM <- london_bills %>%
+    select(time,plague.deaths) %>%
+    rename(deaths="plague.deaths")
+
+parish <- london_parish_all %>%
+    select(time,total) %>%
+    rename(deaths="total")
+
+if (aggWeeks > 1) {
+  ## we need to aggregate the LBoM and parish time series further
+  parish <- with(parish,aggsum(time=time,deaths=deaths,period=aggUnit))
+  LBoM <- with(LBoM,aggsum(time=time,deaths=deaths,period=aggUnit))
+}
+
+dd <- bind_rows(list(LBoM=LBoM,wills=all_wills,
+                     parish=parish),.id="source")  %>%
+    filter((1340<time & time<1380) | (1540<time & time<1680)) %>%
+    ## slightly different version of add_epoch (time-based)
[…]

### 2020-06-12 analysis/plots/setcolours.R

@@ -1,0 +1,38 @@
+## FIX: the source name vector defined here is not a colour setting...
+
+## colours etc. for points:
+library(RColorBrewer)
+mycolourpal <- brewer.pal(12, "Paired")
+mygrey <- grey(0.3) # hex string representation
+colvec1 <- c(parish=mycolourpal[2],
+            LBoM=mycolourpal[6],
+            wills=mygrey)
+pchvec1 <- c(parish=16,LBoM=17,wills=15)
+long_names <- c(wills="PCC Wills",parish="Parish registers",
+                LBoM="Plague deaths")
+
+## legend order
+vec_order <- c("wills","parish","LBoM")
+colvec1 <- colvec1[vec_order]
+long_names <- long_names[vec_order]
+pchvec1 <- pchvec1[vec_order]
+
+## ggplot doesn't like mix of hex and named colours, and
+## doesn't like the colvec elements to be named!
+## order here is parish, LBoM, wills:
+rgb2hex <- function(r,g,b) rgb(r, g, b, maxColorValue = 255)
+col2hex <- function(col) rgb2hex(t(col2rgb(col)))
+colvec <- col2hex(colvec1)
+## Note: the above fanciness is not necessary if colvec1
+##       is all hex.  But it is necessary if colvec1 is a
+##       mixture of hex and named colours like "black".
+
+## The above works as desired, but Ben notes:
+## if you want to get the order right using
+##      scale_colour_manual(breaks=colvecnames,values=colvec)
+## you need to specify a breaks= vector that matches the names in
+## data.source and is in the right order ...
+## Hence: the order in this source name vector matters!
+mysource.name <- c(acm.parish.const="Parish registers",
+                   plague.deaths.LBoM="Plague deaths",
+                   nwills.const="Wills")

### 2020-06-12 analysis/plots/timeseries.R

@@ -1,0 +1,315 @@
+library(tikzDevice)
+library(dplyr)
+library(cowplot)
+library(ggplot2)
+	theme_set(theme_bw()) ## FIXME: don't hardcode? (coded in plot.R?)
+library(directlabels)
+library(epigrowthfit) ## for parish data
+
+
+args = commandArgs(trailingOnly=TRUE)
+## names for sanity, not actually used
+defaults = list(output="",
+                exclude_years=NULL,
+                corner_labs=TRUE,
+                standAlone=TRUE,
+                width=8,height=4,
+                ## max outbreak time in years (chop 1665 time series)
+                max_yrs=1.5,
+                add_CI=FALSE
+                )
+output = if (length(args) < 1) paste0(rtargetname, ".tex") else args[1]
+exclude_years = if (length(args) < 2) defaults[[2]] else eval(parse(text=args[2]))
+corner_labs = if (length(args) < 3) defaults[[3]] else as.logical(args[3])
+standAlone = if (length(args) < 4) defaults[[4]] else as.logical(args[4])
+width = if (length(args) < 5) defaults[[5]] else as.numeric(args[5])
+height = if (length(args) < 6) defaults[[6]] else as.numeric(args[6])
+max_yrs = if (length(args) <7) defaults$max_yrs else as.numeric(args[7])
+add_CI = if (length(args) <8) defaults$add_CI else as.numeric(args[8])
+
+
+if (!interactive()) tikz(output, width=width, height=height,
+                         standAlone=standAlone)
+
+## make sure each outbreak year lasts for a max of max_yrs
+## in particular, this chops off the second half of the 1665-1667
+## outbreak year, which we don't want anyway
+chop_time <-
+    (. %>% group_by(outbreak.year)
+        %>% filter((time-time[1])<max_yrs)
+        %>% ungroup()
+    )
+
+## self-explanatory, I hope ...
+dd_majors3 <- (dd_majors2
+    %>% filter(!outbreak.year %in% exclude_years)
+    %>% chop_time()
+    %>% add_epoch()
+)
+
+
+## fake data for expanding time axes
+## guessing at algorithm: rule is really "expand if tick mark missing"
+## (1564, 1604), "and not too much if it's present")
+dd_expand <- (dd_majors3
+    %>% group_by(epoch,outbreak.year)
+    %>% summarise(time=if (min(time) %% 1 <0.5 &&
+                           diff(range(time))<1) floor(min(time)) else min(time))
+)
+
[…]

### 2020-06-12 analysis/tables/params.R

@@ -1,0 +1,101 @@
+library(methods)
+library(emdbook)
+library(epigrowthfit)
+library(dplyr)
+library(xtable)
+
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) paste0(rtargetname, ".tex") else args[1]
+
+rfun <- function(n) {
+    gsub(".*\\.","",n)
+}
+## testing
+## summary(fitList[[1]][[1]],plague_R0=TRUE)
+## trace("summary",sig="epigrowthfit", browser)
+## debug(epigrowthfit:::summary.epigrowthfit)
+ss <- (summary.fitList(fitList,level=2,plague_R0=TRUE)
+    %>% drop_bad()
+    %>% select(source,outbreak.year,
+           ends_with("value"),
+           ends_with("lower"),
+           ends_with("upper"),
+           "gof")
+)
+          
+ss2 <- (preds
+    %>% as_tibble()
+    %>% mutate(outbreak.year=as.character(epoch),
+               source=
+              case_when(source=="wills" & epoch=="early" ~ "Husting wills",
+                        source=="wills" & epoch=="late" ~ "Canterbury wills",
+                        source=="LBoM"  ~ "London bills",
+                        source=="parish" ~ "London parish"))
+    %>% select(-epoch)
+    %>% bind_rows(ss)
+    %>% mutate(outbreak.year = factor(outbreak.year,
+             levels=c(sort(unique(outbreak.year[substr(outbreak.year,1,1)=="1"])),
+                      "early","late")))
+)
+
+conv <- function(colname) {
+    ss2 %>% select(source,outbreak.year,starts_with(colname)) %>%
+        rename_at(vars(starts_with(colname)),rfun) %>%
+        rename(epidemic="outbreak.year") %>%
+        arrange(epidemic,source)
+}
+
+growth.rate <- conv("growthrate") %>%
+    mutate(value=round(value,1), lower=round(lower,1), upper=round(upper,1))
+doubling.time <- conv("doublingtime") %>%
+    mutate(value=round(value,1), lower=round(lower,1), upper=round(upper,1))
+gof <- conv("gof") %>% rename(value="gof") %>% mutate(lower=NA,upper=NA)
+R0 <- conv("R0")
+final.size <- conv("finalsize")
+
+data <- list(growth.rate = growth.rate,
+             doubling.time = doubling.time,
+             R0 = R0,
[…]

### 2020-06-12 analysis/tables/table.R

@@ -1,0 +1,101 @@
+require(xtable)
+
+#' translate column names to latex text
+#' @param col the column name in the dataframe
+col.translate <- function(col) {
+  switch(col, 
+         source="Source",
+         epidemic="Epidemic",
+         growth.rate = "Growth rate",
+         R0="$\\mathcal{R}_0$",
+         doubling.time="Doubling time",
+         finalsize="Final size")
+}
+
+#' create the text for a table column from fitting results
+#' @param variable a variable in the fitting results, containing columns value, lower and upper
+column <- function(variable, digits=2, percent=FALSE) {
+    value <- paste("%1.", digits, "f", if (percent) "\\%%" else "", sep="")
+    if (all(is.na(variable$lower))) {
+        format <- value
+    } else {
+        format <- paste(value, " \\quad (", value, ", ", value, ")", sep="")
+    }
+    no_limits <- all(is.na(variable$lower)) && all(is.na(variable$upper))
+    fmtfun <- function(v, l, u){
+        if (no_limits) {
+            values = v
+        } else {
+            values = c(v, l, u)
+        }
+        if (is.na(v)) return("NA")
+        if (round(v,digits)<10) format <- paste0("\\ \\,",format)
+        if (percent) values <- values * 100
+        do.call(sprintf, c(list(format), values))
+    }
+    mapply(fmtfun, variable$value, variable$lower, variable$upper)
+}
+
+#' print a latex table for the fitting results, with selected columns
+#' @param fit the fitting results
+#' @param columns the components of fits that will be printed as colummns
+#' @param colnames the column names in the latex table
+#' @param digits a vector of number of digits for each column
+#' @param percent a vector of logical indicating if each column is a percentage
+#' @param comment the latex comment to be added
+#' @param file the file to print
+#' @return NULL
+latexTable <- function(fit, columns, colnames, digits=1, percent=FALSE, comment=NULL, file) {
+  n <- length(columns)
+  if (length(colnames) != n)
+    stop("the columns and colnames must have the same length.")
+  if (length(digits) == 1) digits <- rep(digits, n)
+  if (length(percent) == 1) percent <- rep(percent, n)
+  if (length(digits) != n)
+    stop("the columns and digits must have the same length.")
+  if (length(percent) != n)
[…]

### 2020-06-12 analysis/fits/data.R
Everything looks to be here and products are made. Not sure about chaining.

@@ -1,0 +1,210 @@
+library(dplyr)
+
+#' add an epoch column to data
+#' @param data a dataframe object with an epidemic column
+#' @return a dataframe object
+add_epoch <- function(data) {
+  m <- mutate(data,epoch=factor(ifelse(epidemic %in% c("1348", "1361", "1375"),
+                                       "$<1400$ CE","$>1400$ CE")))
+  return(m)
+}
+
+## #' the generic method for extracting exponential growth rate
+## setGeneric(
+##    "growthRate",
+##    def = function(fit) {
+##      standardGeneric("growthRate")
+##    }
+## )
+
+#' extract growth rate from an fitEpidemics object
+setMethod(
+  "growthRate",
+  signature="fitEpidemics",
+  definition = function(obj) {
+    data <- data.frame()
+    fit <- obj@epidemics
+    for (epidemic in names(fit)) {
+        data <- rbind(data, 
+                      cbind(as.data.frame(t(fit[[epidemic]]@growthRate)),
+                            epidemic = epidemic))
+    }
+    add_epoch(data)
+}
+)
+
+#' @export
+#' compute goodness of fit for each epigrowthfit object in a fitEpidemics object
+setMethod(
+  "gof",
+  "fitEpidemics",
+  definition = function(obj) {
+    data <- data.frame()
+    fit <- obj@epidemics
+    for (epidemic in names(fit)) {
+        gofline <- gof(fit[[epidemic]])
+        data <- rbind(data, data.frame(value=gofline,lower=NA,upper=NA,epidemic))
+    }
+    add_epoch(data)
+  }
+)
+
+#' compute sampled CI
+#' @param data a dataframe with columns value, lower and upper
+#' @return a vector
+sampledCI <- function(data) {
+  # if there is a single sample run, do not sample
+  if (nrow(data) == 1) return(data[1, c("value", "lower", "upper")])
+  # otherwise, sample from a normal distribution centered at value, 
+  # with the standard deviation as a quarter of the 95% CI
+  value <- mean(data$value)
+  rn <- c()
+  for (i in 1:nrow(data)) {
+    if (any(is.na(data[i,]))) next
+    sd <- (data$upper[[i]] - data$lower[[i]]) / 4
+    rn <- c(rn, rnorm(1000, mean=data$value[[i]], sd=sd))
+  }
+  data.frame(value = value,
+             lower = quantile(rn, probs=0.025)[[1]],
[…]

### 2020-06-12 analysis/fits/set_minors.R

@@ -1,0 +1,8 @@
+model <- "logistic"
+max_theta <- 1e4
+epitype <- "minor"  
+optimizer <- "nlminb"
+optimfun <- NULL
+x0link <- "logit"
+optCtrl <- list(eval.max=1000,iter.max=1000)
+method <- NULL

### 2020-06-12 analysis/fits/wills_delay.R

@@ -1,0 +1,56 @@
+library(epigrowthfit)
+library(dplyr)
+library(tidyr)
+
+## only have parish records so uninteresting to measure delay times
+yrs <- unique(as.numeric(dd_majors$outbreak.year))
+yrs <- yrs[yrs>1500 & yrs != 1593]
+res <- list()
+for (oy in yrs) {
+    dd1 <- (dd_majors
+        %>% filter(outbreak.year==as.character(oy))
+        %>% select(-outbreak.year)
+        %>% split(.$source)
+    )
+    focal <- "LBoM"
+    ## interpolate to LBoM
+    for (s in c("parish","wills","LBoM")) {
+        if (nrow(dd1[[s]])>0) {
+            ## CCF computation
+            if (s!=focal) {
+                adat <- ( with(dd1[[s]],
+                               approx(time,deaths,xout=dd1[[focal]]$time))
+                    %>% setNames(c("time","deaths"))
+                    %>% as.data.frame()
+                    %>% cbind(LBoM=dd1[[focal]]$deaths)
+                    %>% na.omit()
+                )
+                cc <- ccf(adat$deaths,adat$LBoM)
+                cc_peak <- cc$lag[which.max(cc$acf)]
+                cc_val <- max(cc$acf)
+            } else {
+                ## no CCF comparison for focal source
+                cc_peak <- cc_val <- NA
+            }
+            tvals <- seq(min(dd1[[s]]$time),max(dd1[[s]]$time),length.out=250)
+            spl <- with(dd1[[s]],stats::smooth.spline(time,deaths))
+            smooth_vals <- predict(spl,tvals)$y
+            smooth_peak <- tvals[which.max(smooth_vals)]
+            res <- c(res,
+                     list(data.frame(source=s,outbreak.year=oy,
+                                LBoM_lag=cc_peak,
+                                cor_max=cc_val,
+                                smooth_peak,
+                                stringsAsFactors=FALSE)))
+        }
+    }
+}
+res <- (bind_rows(res)
+    %>% group_by(outbreak.year)
+    ## peak difference in weeks
+    %>% mutate(peak_diff=52*(smooth_peak-smooth_peak[source=="LBoM"]))
+    %>% filter(source!="LBoM")
+)
+
+saveRDS(res,file="wills_delay.rds")
+

### 2020-06-12 analysis/fits/wills_delay_defs.R

@@ -1,0 +1,43 @@
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) "wills_delay_defs.tex" else args[1]
+
+library(dplyr)
+res <- readRDS(input_files[[1]])
+
+parish_cormax_range <- res %>% filter(source=="parish") %>% pull(cor_max) %>% range()
+parish_ccflag_range <- res %>% filter(source=="parish") %>% pull(LBoM_lag) %>% range()
+parish_peaklag_range <- res %>% filter(source=="parish") %>% pull(peak_diff) %>% range()
+
+wills_cormax_range <- res %>% filter(source=="wills") %>% pull(cor_max) %>% range()
+wills_ccflag_range <- res %>% filter(source=="wills") %>% pull(LBoM_lag) %>% range()
+wills_peaklag_range <- res %>% filter(source=="wills") %>% pull(peak_diff) %>% range()
+
+catt <- function(...,f=file) {
+    cat(...,file=f,append=TRUE)
+}
+
+latexout <- function(x,nm,fmt="%1.1f") {
+    catt(sprintf("\\newcommand{\\%s}{%s}\n",nm,sprintf(fmt,x)))
+}
+
+unlink(file)
+
+catt("% parish range of max CCF with LBoM\n")
+latexout(parish_cormax_range[1],"parishccfmaxmin","%1.2f")
+latexout(parish_cormax_range[2],"parishccfmaxmax","%1.3f")
+catt("% parish range of CCF lag from LBoM\n")
+latexout(parish_ccflag_range[1],"parishccflagmin","%d")
+latexout(parish_ccflag_range[2],"parishccflagmax","%d")
+catt("% parish range of peak lag from LBoM\n")
+latexout(parish_peaklag_range[1],"parishpeaklagmin")
+latexout(parish_peaklag_range[2],"parishpeaklagmax")
+
+catt("% wills range of max CCF with LBoM\n")
+latexout(wills_cormax_range[1],"willsccfmaxmin","%1.2f")
+latexout(wills_cormax_range[2],"willsccfmaxmax","%1.2f")
+catt("% wills range of CCF lag from LBoM\n")
+latexout(wills_ccflag_range[1],"willsccflagmin","%d")
+latexout(wills_ccflag_range[2],"willsccflagmax","%d")
+catt("% wills range of peak lag from LBoM\n")
+latexout(wills_peaklag_range[1],"willspeaklagmin")
+latexout(wills_peaklag_range[2],"willspeaklagmax")

### 2020-06-12 analysis/plots/KvsR.R

@@ -1,0 +1,52 @@
+library(tikzDevice)
+
+texname = paste0(rtargetname, ".tex")
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) texname else args[1]
+standAlone = if (length(args) < 2) TRUE else as.logical(args[2])
+width = if (length(args) < 3) 6 else as.numeric(args[3])
+height = if (length(args) < 4) 4 else as.numeric(args[4])
+cat(width,height,"\n")
+
+tikz(file, width=width, height=height,
+                         standAlone=standAlone)
+
+Kratfun <- function(Rratio, aKR1=1) {
+  return((-1/(aKR1))*log(1 - Rratio*(1 - exp(-aKR1))))
+}
+##Rratio <- seq(1.01,1.6/1.1,length=100)
+## Rratio <- seq(1.01, 1.9,length=1000)
+Rratio <- seq(1.01, 2.1,length=1000)
+Krat <- Kratfun(Rratio, aKR1=1)
+plot(Rratio, Krat, type="n", bty="L", las=1,
+     ylim=c(1,3),
+     ##xlab=expression(R[0][2] / R[0][1]), ylab=expression(K[R][2] / K[R][1]))
+     xlab="${\\mathcal R}_{0,2}/{\\mathcal R}_{0,1}$",
+     ylab="$K_{\\rm R,2}/K_{\\rm R,1}$")
+grid()
+lines(Rratio, Krat, lwd=4)
+Krat <- Kratfun(Rratio, aKR1=2)
+lines(Rratio, Krat, lwd=4, col="blue")
+## approx:
+lines(Rratio, Rratio, lwd=4, col="red", lty="dotted")
+## exact:
+Krat <- Kratfun(Rratio, aKR1=0.5)
+lines(Rratio, Krat, lwd=4, col="red")
+titlestring <- "$a \\, K_{\\rm R,1}$"
+legend("topleft",bty="n", title=titlestring, lwd=4, 
+       col=c("blue","black","red"), legend=c(2,1,0.5))
+
+## minimum rat density ratio is same as R ratio
+## FRYed from supp.Rnw
+ratR0 <- function(r){ (r / 365 * 18) + 1 }
+r1 <- rvals["fit","early"]
+r2 <- rvals["fit","late"]
+R01 <- ratR0(r1)
+R02 <- ratR0(r2)
+R0ratio <- R02/R01
+
[…]

### 2020-06-12 analysis/plots/london_pop.R

@@ -1,0 +1,57 @@
+library(epigrowthfit)
+library(dplyr)
+library(tikzDevice)
+
+texname = paste0(rtargetname, ".tex")
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) texname else args[1]
+standAlone = if (length(args) < 2) TRUE else as.logical(args[2])
+width = if (length(args) < 3) 6 else as.numeric(args[3])
+height = if (length(args) < 4) 4 else as.numeric(args[4])
+cat(width,height,"\n")
+
+tikz(file, width=width, height=height,
+                         standAlone=standAlone)
+
+pop <- (pop_london
+    %>% select(year, low, high, total)
+    %>% filter(year >= 1300 & year <= 1700)
+)
+
+## margins:  [default is c(b,l,t,r) = c(5, 4, 4, 2) + 0.1]
+par(mar=c(3,5,2,2))
+
+## should avoid scientific notation in tick labels:
+## https://stackoverflow.com/questions/5963047/do-not-want-scientific-notation-on-plot-axis
+
+plot(total ~ year, data=pop, type="n",
+     ylim=c(0,470000), yaxs="i",
+     bty="L", las=1, 
+     xlab="",
+     ylab="",
+     yaxt="n", # make my own with nice tick labels
+     main="Estimated human population of London"
+     )
+axis(side=2, at=(0:4)*10^5, labels=c("0","$10^5$", paste("$", 2:4, "\\times 10^5$")), las=1)
+
+grid()
+
+lines(total ~ year, data=pop,
+     type="o", lwd=2, pch=21, bg="cyan", cex=2, xpd=NA
+     )
+
+##lines(low ~ year, data=pop, lty="dotted", col="grey")
+##lines(high ~ year, data=pop, lty="dotted", col="grey")
+
+## Data from Creighton 1891, Vol 1, p. 660:
+creighton <- data.frame( year=c(1603,1625,1665), total=c(250000,320000,460000) )
+
+lines(total ~ year, data=creighton,
+     type="o", lwd=2, pch=21, bg="grey", cex=2, xpd=NA
+     )
+
+legend("topleft", bty="n",
+       legend=c("Creighton", "Finlay"),
+       lwd=2, pch=21, pt.bg=c("grey", "cyan"), pt.cex=2)
+
+dev.off()

### 2020-06-12 analysis/plots/plot.setup.R

@@ -1,0 +1,40 @@
+library(tikzDevice)
+
+#' plot setup function
+#' @param plot.file file name
+#' @param width width
+#' @param height height
+#' @param pointsize pointsize
+#' @param theme ggplot theme
+#' @
+#' @importFrom tikzDevice tikz
+#' @importFrom tools file_ext
+#' @importFrom grDevices dev.off pdf png
+#' @usage plot.setup(plot.file, width=5, height=4, pointsize=12)
+#' @export plot.setup
+plot.setup <- function(plot.file, width=5, height=4, pointsize=12,
+                       theme = theme_bw,
+                       pal = "Set1"
+                       ) {
+		  dev.off()
+        theme_set(theme_bw())
+        scale_colour_discrete <- function(...,palette=pal) {
+            scale_colour_brewer(...,palette=pal)
+        }
+	if (plot.file == "") return()
+	ext = tail(strsplit(plot.file, "\\.")[[1]], n=1)
+	switch (ext,
+	        pdf = pdf(plot.file, width=width, height=height, pointsize=pointsize),
+	        tex = tikz(plot.file, standAlone=TRUE, width=width, height=height, 
+                     pointsize=pointsize),
+	        png = png(plot.file, width=width, height=height, units="in", 
+	                  pointsize=pointsize)
+	)
+}
+
+#' plot closing function
+#' @param plot.file file name
+#' @export plot.close
+plot.close <- function(plot.file) {
+	if (plot.file != "") dev.off()
+}

### 2020-06-12 analysis/plots/plot3sources.R

@@ -1,0 +1,147 @@
+library(epigrowthfit)
+library(dplyr)
+library(lubridate)
+library(tikzDevice)
+library(RColorBrewer)
+
+args = commandArgs(trailingOnly=TRUE)
+
+texname <- paste0(rtargetname, ".tex")
+## testing
+## args <- c("tmp.tex","FALSE","9","7")
+file = if (length(args) < 1) texname else args[1]
+standAlone = if (length(args) < 2) TRUE else as.logical(args[2])
+width = if (length(args) < 3) 12 else as.numeric(args[3])
+height = if (length(args) < 4) 8 else as.numeric(args[4])
+cat(width,height,"\n")
+
+aggWeeks <- 4
+aggUnit <- paste(aggWeeks,"weeks")
+myylab <- if (aggUnit=="1 week") "Weekly counts" else paste(aggWeeks, "-weekly counts")
+
+wills <- aggregate_wills(canterbury_wills_individual,
+                         include_all=TRUE, aggregation=aggUnit)
+LBoMvar <- "plague.deaths"
+##LBoMvar <- "all.cause.deaths"
+LBoM <- london_bills[,c("time",LBoMvar)]
+parishVar <- "total"
+parish <- london_parish_all[,c("time",parishVar)]
+
+if (aggWeeks > 1) {
+  ## we need to aggregate the LBoM and parish time series further
+  cat("plot3sources.R: Aggregating LBoM and parish...\n")
+  LBoMorig <- LBoM
+  parishOrig <- parish
+  parish <- with(parish,aggsum(time=time,deaths=parish[,parishVar],period=aggUnit))
+  names(parish)[-1] <- names(parishOrig)
+  LBoM <- with(LBoM,aggsum(time=time,deaths=LBoM[,LBoMvar],period=aggUnit))
+  names(LBoM)[-1] <- names(LBoMorig)
+}
+
+tmin <- 1540
+tmax <- 1680
+ymax <- max(LBoM[,LBoMvar], parish$total, na.rm=TRUE)
+ymin <- 0.9
+
+make_polygon <- function(yr, col="yellow") {
+  xmin <- yr - 1
+  xmax <- yr + 2
+  x <- c(xmin, xmax, xmax, xmin, xmin)
+  y <- c(ymin, ymin, ymax, ymax, ymin)
+  polygon(x, y, col=col, border=NA)
+  xmid <- yr + 0.5
[…]

### 2020-06-12 analysis/plots/plot3sourcesmajors.R

@@ -1,0 +1,209 @@
+## show bills, wills and parish registered so we can see the major
+## epidemics are evident simultaneously no matter what source we
+## examine
+
+library(epigrowthfit)
+library(dplyr)
+library(lubridate)
+library(tikzDevice)
+
+args = commandArgs(trailingOnly=TRUE)
+texname <- paste0(rtargetname, ".tex")
+file = if (length(args) < 1) texname else args[1]
+
+aggWeeks <- 1
+aggUnit <- paste(aggWeeks,"weeks")
+myylab <- if (aggUnit=="1 weeks") "" else paste(aggWeeks, "-weekly counts")
+
+wills <- aggregate_wills(canterbury_wills_individual,
+                         include_all=TRUE, aggregation=aggUnit)
+LBoMvar <- "plague.deaths"
+##LBoMvar <- "all.cause.deaths"
+LBoM <- london_bills[,c("time",LBoMvar)]
+parishVar <- "total"
+parish <- london_parish_all[,c("time",parishVar)]
+
+if (aggWeeks > 1) {
+  ## we need to aggregate the LBoM and parish time series further
+  LBoMorig <- LBoM
+  parishOrig <- parish
+  parish <- with(parish,aggsum(time=time,deaths=parish[,parishVar],period=aggUnit))
+  names(parish)[-1] <- names(parishOrig)
+  LBoM <- with(LBoM,aggsum(time=time,deaths=LBoM[,LBoMvar],period=aggUnit))
+  names(LBoM)[-1] <- names(LBoMorig)
+}
+
+tmin <- 1540
+tmax <- 1680
+ymax <- max(LBoM[,LBoMvar], parish$total, na.rm=TRUE)
+ymin <- 0.9
+
+make_polygon <- function(yr, col="yellow") {
+  xmin <- yr - 1
+  xmax <- yr + 2
+  x <- c(xmin, xmax, xmax, xmin, xmin)
+  y <- c(ymin, ymin, ymax, ymax, ymin)
+  polygon(x, y, col=col, border=NA)
+  xmid <- yr + 0.5
+  ##text(xmid, ymax, label=yr, font=2, xpd=NA)
+  mtext(text=yr, side=3, font=2, at=xmid, xpd=NA, cex=cex.text)
+}
+
+## REMIND MYSELF OF THE pch VALUES
[…]

### 2020-06-12 analysis/plots/plotall.R

@@ -1,0 +1,22 @@
+library(epigrowthfit)
+library(ggplot2)
+library(tikzDevice)
+
+args = commandArgs(trailingOnly=TRUE)
+file = pdfname
+file = paste0(rtargetname, ".tex")
+
+data = subset(plague, aggregation=="weekly" & place == "London")
+figure <- ggplot(data=data, mapping=aes(x=time, y=plague.deaths, 
+                                        color=severity, group=outbreak.year)) +
+    scale_y_continuous("weekly plague deaths",
+                       expand=c(0,0)) + # multiplicative and additive expansion of scale
+    expand_limits(y=7500) +
+    ##scale_x_continuous("year")+
+    scale_x_continuous("")+
+    geom_line() +
+    scale_colour_brewer(palette="Set1") # first colour is red
+
+plot.setup(file, width=4, height=2)
+print(figure)
+plot.close(file)

### 2020-06-12 analysis/plots/plotallscaled.R

@@ -1,0 +1,36 @@
+library(epigrowthfit)
+library(ggplot2)
+library(tikzDevice)
+
+file = pdfname
+file = paste0(rtargetname, ".tex")
+
+data = subset(plague, aggregation=="weekly" & place == "London")
+data$scaled = 1000*data$plague.deaths/data$population
+figure <- ggplot(data=data, mapping=aes(x=time, y=scaled, color=severity, group=outbreak.year)) +
+    ##scale_y_continuous("weekly plague/other causes deaths") +
+    ##scale_y_continuous("weekly plague deaths \\,\\emph{per capita}") +
+    scale_y_continuous("weekly plague deaths\nper 1000 population",
+                       expand=c(0,0)) + # multiplicative and additive expansion of scale
+    expand_limits(y=19.9) +
+    ##scale_x_continuous("year")+
+    scale_x_continuous("")+
+    geom_line() +
+    scale_colour_brewer(palette="Set1") # first colour is red
+
+plot.setup(file, width=4, height=2)
+print(figure)
+plot.close(file)
+
+## get peak values
+majors <- c(1563, 1593, 1603, 1625, 1665)
+nmajors <- length(majors)
+NAs <- rep(NA,nmajors)
+dfpeaks <- data.frame(row.names=majors, peak=NAs, scaled=NAs)
+for (oy in majors) {
+  dd <- subset(data,outbreak.year==oy)
+  this.peak <- max(dd$plague.deaths, na.rm=TRUE)
+  this.peak.scaled <- max(dd$scaled, na.rm=TRUE)
+  dfpeaks[as.character(oy),] <- c(this.peak, this.peak.scaled)
+}
+print(dfpeaks)

### 2020-06-12 analysis/plots/plotwills14c.R

@@ -1,0 +1,159 @@
+## show 14th century Husting wills in style of plot3sourcesmajors.R
+
+library(epigrowthfit)
+library(dplyr)
+library(lubridate)
+library(tikzDevice)
+
+texname = paste0(rtargetname, ".tex")
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) texname else args[1]
+
+aggWeeks <- 1
+aggUnit <- paste(aggWeeks,"weeks")
+myylab <- if (aggUnit=="1 weeks") "" else paste(aggWeeks, "-weekly counts")
+
+wills <- aggregate_wills(husting_wills_individual,
+                         include_all=TRUE, aggregation=aggUnit)
+
+tmin <- 1300
+tmax <- 1400
+ymax <- max(wills[,"nwills"], na.rm=TRUE)
+ymin <- 0.9
+
+## REMIND MYSELF OF THE pch VALUES
+if (interactive()) {
+  vec <- 1:25
+  plot(vec,vec,type="n")
+  grid()
+  points(vec,vec,pch=vec,cex=2)
+}
+
+majors <- c(1348,1361,1368,1375)
+is.major <- function(time) {
+  res <- rep(FALSE, length(time))
+  for (yr in majors) {
+    res <- res | (time > yr & time < yr+1)
+  }
+  return(res)
+}
+
+## get ymin and ymax for each major epidemic:
+yrange_defs <- epidemic_defs %>%
+  filter(severity=="major", outbreak.year<1500) %>%
+  mutate(wmin=NA,wmax=NA,ymin=NA,ymax=NA)
+for (yr in majors) {
+  iyr <- which(yrange_defs[,"outbreak.year"]==yr)
+  t1 <- yrange_defs[iyr,"start"]
+  t2 <- yrange_defs[iyr,"end"]
+  wtmp <- wills %>% filter(time >= t1 & time <= t2)
+  w1 <- min(wtmp[["nwills"]], na.rm=TRUE)
+  w2 <- max(wtmp[["nwills"]], na.rm=TRUE)
+  y1 <- w1
+  y2 <- w2
+  yrange_defs[iyr,"wmin"] <- w1
+  yrange_defs[iyr,"wmax"] <- w2
+  yrange_defs[iyr,"ymin"] <- y1
+  yrange_defs[iyr,"ymax"] <- y2
+}
+print(yrange_defs)
+
+## MAKE THE PLOT
+
[…]

### 2020-06-12 analysis/plots/probate_dates.R

@@ -1,0 +1,109 @@
+library(epigrowthfit)
+library(dplyr)
+source("../fits/epidemics.R")
+library(ggplot2); theme_set(theme_bw())
+library(tikzDevice)
+
+args = commandArgs(trailingOnly=TRUE)
+file = pdfname
+file = paste0(rtargetname, ".tex")
+
+## FIXME: this should probably be independent
+if (require(willsr)) {
+    ## probate dates and times in husting_wills_individual are not clean/complete
+    ww <- (willsr::wills
+        %>% select(ProbateDate)
+        %>% mutate(time=lubridate::decimal_date(ProbateDate))
+    )
+    saveRDS(ww, file = "../../data/probate.rds")
+} else {
+    ww <- readRDS("../../data/probate.rds")
+}
+
+## get info on wills-written dates from epigrowthfit
+## (some repetition from analysis/plots/timeseries.R here)
+
+L1 <- load('../fits/.epidemics.RData')
+L2 <- load('../fits/.comb.RData')  ## incl dd_majors2
+
+## utility: extract time range of each outbreak year
+get_rng <- (. %>% group_by(outbreak.year)
+    %>% select(date)
+    %>% summarise(min=min(date),max=max(date))
+    %>% ungroup()
+)
+
+dd_majors3 <- (dd_majors2
+    %>% filter(time<1400)
+    %>% select(-c(source,time))
+    %>% mutate_at("outbreak.year",as.character)
+    %>% rename(count="deaths")
+)
+
+ed_early <- (filter(epidemic_defs,end<1400)
+    %>% mutate_at("outbreak.year",as.character)
+    %>% full_join(get_rng(dd_majors3))
+)
+
+probate_extend <- 1 ## extra years
+
+
+## for a given outbreak year (row of the epidemic definitions table),
+##  extract the relevant probate wills and aggregate them weekly,
+##  starting from the first day of the *written wills* aggregation
+##  previously constructed ...
+get_epid_counts <- function(ee) {
+    ## take values up to the next probate *after* the extension period
+    next_probate <- head(ww$time[ww$time>ee[["end"]]]+probate_extend,1)
+    ww0 <- ww[get_range(c(ee[["start"]],next_probate),
+                        data=ww, maxlen=Inf),]
+    get_date_count(ww0,
+                   unit="1 week",
+                   date_col="ProbateDate",
+                   count_col="probate_count",
[…]

### 2020-06-12 analysis/plots/seasonalgrowth.R

@@ -1,0 +1,82 @@
+library(epigrowthfit)
+library(cowplot)
+library(ggplot2); theme_set(theme_bw())
+library(dplyr)
+library(purrr)
+library(tikzDevice)
+
+defaults = list(output="",width=7,height=4)
+
+width <- defaults[["width"]]
+height <- defaults[["height"]]
+
+texname <- paste0(rtargetname, ".tex")
+
+args = commandArgs(trailingOnly=TRUE)
+file = pdfname
+file = paste0(rtargetname, ".tex")
+
+## hack since we want to import 'fitList' from both minors and majors ...
+## minors$ to distinguish from minors_combo_logit (etc.)
+## [FIXME: more thorough renaming]
+minorList <- envir_list[[grep("minors$", names(envir_list))]]$fitList
+minorSum <- sumfun(minorList,level=2)
+majorList <- envir_list[[grep("major", names(envir_list))]]$fitList
+majorSum <- sumfun(majorList,level=2)
+
+## slightly redundant with conv info from summary/summary.fitList/sumfun
+conv_info <- map_dfr(minorList,get_conv,.id="source")
+conv_info$conv
+
+bad <- conv_info %>% filter(conv==1)
+bad$msg
+stopifnot(nrow(bad)==2)  ## make sure we're only skipping 2 ...
+minorList[["Canterbury wills"]][["1581"]]@mle2@details
+
+##%>% select(source,outbreak.year,conv,msg)
+par(mfrow=c(1,2))
+for (i in seq(nrow(bad))) {
+    plot(minorList[[bad$source[i]]][[bad$outbreak.year[[i]]]])
+}
+
+## hack since we loaded data *including pdfname*?
+dd <- (bind_rows(minor=minorSum,major=majorSum,.id="severity")
+    %>% filter(outbreak.year>1500)
+    %>% mutate(source=mod_source(source))  ## recode source names
+    %>% drop_bad()
+    %>% filter(conv==0)
+    %>% select(outbreak.year,source,
+               severity,starts_with("growthrate"))
+)
+
+labfun <- function(brks) {
+    f <- split(brks,substr(brks,1,2))
+    f2 <- lapply(f,
+           function(x) {
+        x[-1] <- paste0("'",substr(x[-1],3,nchar(x[-1])))
+		  x[[1]] <- sprintf("\\textbf{%s}", x[[1]])
+        return(x)
+    })
+    return(unlist(f2))
+}
+
[…]

### 2020-06-12 analysis/tables/epochsum.R

@@ -1,0 +1,30 @@
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) paste0(rtargetname, ".tex") else args[1]
+
+## memisc::mtable
+## texreg
+## huxtable
+library(huxtable)
+library(glmmTMB)
+library(broom.mixed)
+
+cc <- c("Intercept (14th-c. wills log growth rate)"="(Intercept)",
+        "epoch (late vs. early)"="epochlate",
+        "source (parish vs. wills)"="sourceparish",
+        "source (LBoM vs. wills)"="sourceLBoM")
+
+h0 <- huxreg(" "=f0, # give model blank name so we don't get '(1)'
+             tidy_args=list(effects="fixed",conf.int=TRUE),
+             coefs=cc,
+             error_format = '({conf.low} -- {conf.high})',
+             stars=NULL, ## no sig. stars
+             error_pos="right",
+             statistics="nobs" # don't include logLik and AIC
+             )
+names(h0) <- c("","estimate","95% CI")
+h1 <- add_colnames(h0)
+cat(to_latex(h1,tabular_only=TRUE),
+    file=file,append=FALSE)
+
+
+

### 2020-06-12 analysis/tables/minorparams.R

@@ -1,0 +1,60 @@
+library(methods)
+library(emdbook)
+library(epigrowthfit)
+library(dplyr)
+library(xtable)
+
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) paste0(rtargetname, ".tex") else args[1]
+
+rfun <- function(n) {
+    gsub(".*\\.","",n)
+}
+
+ss <- (summary.fitList(fitList,level=2)
+    %>% drop_bad()
+)
+
+## check bad fits.
+## don't need gof>0 requirement
+with(ss,table(source,cut(growthrate.value,c(-Inf,1,100,Inf)),gof>0))
+
+bad_to_na <- function(x,lwr=-Inf,upr=Inf) {
+    x[x<lwr | x>upr] <- NA
+    return(x)
+}
+
+# filter out/NA-replace bad values
+ss <- (ss
+    %>% filter(gof>0,growthrate.value>1,growthrate.value<100,conv==0)
+    %>% mutate_at(vars(starts_with("growthrate")),
+                  funs(bad_to_na(.,0.1,200)))
+    %>% mutate_at(vars(starts_with("doubling")),
+                  funs(bad_to_na(.,1,1000)))
+)
+
+conv <- function(colname) {
+    ss %>% select(source,outbreak.year,starts_with(colname)) %>%
+        rename_at(vars(starts_with(colname)),rfun) %>%
+        rename(epidemic="outbreak.year") %>%
+        arrange(epidemic,source)
+}
+
+growth.rate <- conv("growthrate")
+doubling.time <- conv("doublingtime")
+gof <- conv("gof") %>% rename(value="gof") %>% mutate(lower=NA,upper=NA)
+R0 <- conv("R0")
+final.size <- conv("finalsize")
+
+data <- list(growth.rate = growth.rate,
+             doubling.time = doubling.time,
+             goodness.of.fit = gof)
+
+latexTable(
+  data, 
+  columns = c("growth.rate", "doubling.time", "goodness.of.fit"),
+  colnames = c("Growth rate [$1/\\text{year}]$", "Doubling time [days]",
+               "\\ \\ $R^2$"),
+  comment = "generated by analysis/tables/minorparams.R",
+  digits = c(1,1,3), # 3 for gof
+  file = file)

### 2020-06-12 analysis/tables/suppwindows.R

@@ -1,0 +1,48 @@
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) "suppwindows.tex" else args[1]
+majors_only = if (length(args) < 2) TRUE else as.logical(args[2])
+
+##library(xtable)
+library(Hmisc) # for latex()
+library(epigrowthfit)
+library(dplyr)
+## source("analysis/data.R")
+
+window_defs <- (summary.fitList(fitList,key="type")
+    %>% select(type,outbreak.year,
+               fulltime.min,fulltime.max,
+               wintime.min,wintime.max)
+    %>% arrange(fulltime.min)  ## arrange BEFORE converting to char!
+    %>% mutate_at(grep("\\.(min|max)",names(.)),
+                  funs(format(lubridate::date_decimal(.),"%d %b %Y")))
+    %>% rename("outbreak year"=outbreak.year,
+               "fitting window (start)"=wintime.min,
+               "fitting window (end)"=wintime.max,
+               "outbreak window (start)"=fulltime.min,
+               "outbreak window (end)"=fulltime.max)
+)
+    
+
+cat("% created by analysis/tables/suppwindows.R\n", file=file)
+
+## I (DE) find xtable too inflexible for what we need so am
+## using Hmisc::latex instead.
+## table = xtable(window_defs)
+## align(table) = "|l|l|l|l|l|l|l|"
+## print(table,
+##       append = TRUE,
+##       floating = FALSE,
+##       include.rownames = FALSE, 
+##       comment = TRUE, 
+##       sanitize.text.function = function(x) {x},
+##       file=file)
+
+ltx <- latex(window_defs,
+      multicol=TRUE, # allow \multicolumn in header
+      rowname=NULL,
+      cgroup=c("", "Outbreak Window", "Fitting Window"),
+      colheads=c("Source","Outbreak Year", "start", "end", "start", "end"),
+      col.just = c("l","c",rep("r",4)),
+      table.env=FALSE,
+      file=file
+      )

### 2020-06-12 analysis/tables/suppwindowsminor.R

@@ -1,0 +1,34 @@
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) "suppwindowsminor.tex" else args[1]
+majors_only = if (length(args) < 2) TRUE else as.logical(args[2])
+
+library(Hmisc) # for latex()
+library(epigrowthfit)
+library(dplyr)
+
+window_defs <- (summary.fitList(fitList,key="type")
+    %>% select(type,outbreak.year,
+               fulltime.min,fulltime.max,
+               wintime.min,wintime.max)
+    %>% arrange(fulltime.min)  ## arrange BEFORE converting to char!
+    %>% mutate_at(grep("\\.(min|max)",names(.)),
+                  funs(format(lubridate::date_decimal(.),"%d %b %Y")))
+    %>% rename("outbreak year"=outbreak.year,
+               "fitting window (start)"=wintime.min,
+               "fitting window (end)"=wintime.max,
+               "outbreak window (start)"=fulltime.min,
+               "outbreak window (end)"=fulltime.max)
+)
+    
+
+cat("% created by analysis/tables/suppwindows.R\n", file=file)
+
+ltx <- latex(window_defs,
+      multicol=TRUE, # allow \multicolumn in header
+      rowname=NULL,
+      cgroup=c("", "Outbreak Window", "Fitting Window"),
+      colheads=c("Source","Outbreak Year", "start", "end", "start", "end"),
+      col.just = c("l","c",rep("r",4)),
+      table.env=FALSE,
+      file=file
+      )

### 2020-06-12 analysis/tables/wills_delay_tab.R

@@ -1,0 +1,45 @@
+args = commandArgs(trailingOnly=TRUE)
+file = if (length(args) < 1) "wills_delay_tab.tex" else args[1]
+
+
+library(xtable)
+library(dplyr)
+## leave out second line (\\newline) stuff/shorten
+## labels for now; will only work
+## with paragraph column format (e.g. p{3cm}), but I'm not sure how
+## that will play with the formatting of the rest of the column. I
+## think if I were coding this directly I would use \multicolumn{}
+## to modify the header format ... ?
+## https://tex.stackexchange.com/questions/152066/color-first-row-in-a-xtable-generated-table
+res <- (readRDS("../fits/wills_delay.rds")
+    %>% rename("outbreak year"=outbreak.year,
+               ## "max \\newline cross-correlation"=cor_max,
+               "max CCF"=cor_max,
+               ## "CCF lag \\newline (rel. to LBoM)"=LBoM_lag,
+               "CCF lag"=LBoM_lag,
+               "peak"=smooth_peak,
+               ## "peak lag \\newline (rel. to LBoM)"=peak_diff)
+               "peak lag"=peak_diff)
+)
+
+rws <- which(res$source=="wills")-1
+col <- rep("\\rowcolor[gray]{0.9}", length(rws))
+cat("% created by analysis/tables/wills_delay_tab.R\n", file=file)
+table <- xtable(res)
+
+## align(table) = "|l|l|l|l|l|l|l|"
+## n.b.: this breaks things:
+display(table) <- c("s","s","d","d","f","f","f")
+digits(table) <- c(NA,NA,0,0,3,2,2)
+## https://stackoverflow.com/questions/11109489/xtable-header-manipulation
+## help?
+print(table,
+      append = FALSE,
+      floating = FALSE,
+      include.rownames = FALSE, 
+      comment = TRUE, 
+      sanitize.text.function = function(x) {x},
+      file=file,
+      add.to.row = list(pos = as.list(rws), command = col))
+
+

### 2020-06-12 analysis/plots/plot_all.R
Finally got crown where I want it?

@@ -91,7 +91,7 @@ begin      end         dsource  nx fac  label size
 1551-01-01 1551-12-31  parish   0 5.5  sickness 1.5
 1557-01-01 1559-12-31  parish   0 3  flu                2
 1649-01-30 1660-06-29  parish   1 2  Interregnum  2
-1558-11-17 1558-11-17  parish   0 2.2    \\includegraphics[width=15pt]{crown.png} 2",
+1558-11-17 1558-11-17  parish   0 2.2    \\includegraphics[width=15pt]{../../images/crown.png} 2",
 colClasses=c("Date","Date","character","numeric","numeric","character",
              "numeric"))
 

### 2020-06-12 analysis/plots/plot_all.R
What just happened (mysterious failure to commit)?

@@ -91,7 +91,7 @@ begin      end         dsource  nx fac  label size
 1551-01-01 1551-12-31  parish   0 5.5  sickness 1.5
 1557-01-01 1559-12-31  parish   0 3  flu                2
 1649-01-30 1660-06-29  parish   1 2  Interregnum  2
-1558-11-17 1558-11-17  parish   0 2.2    \\includegraphics[width=15pt]{../../images/crown.png} 2",
+1558-11-17 1558-11-17  parish   0 2.2    \\includegraphics[width=15pt]{crown.png} 2",
 colClasses=c("Date","Date","character","numeric","numeric","character",
              "numeric"))
 

### 2020-06-12 supp.Rnw
amsync

@@
 ratR0 <- function(r){ (r / 365 * 18) + 1 }
-load("analysis/fits/.epochsum.RData")
+load("analysis/fits/epochsum.RData")
 r1 <- rvals["fit","early"]

### 2020-06-12 analysis/fits/set_nlminb_logit_major.R
Some cribbed things were not kept!

@@ -1,0 +1,8 @@
+model <- "logistic"
+max_theta <- 1e4
+epitype <- "major"
+optimizer <- "nlminb"
+optimfun <- NULL
+x0link <- "logit"
+optCtrl <- list()
+method <- NULL

### 2020-06-12 analysis/plots/plot_allsources.R
change line order in legend

@@ -111,6 +111,8 @@ event_tab2 <- (event_tab
                                   time<=end+time_eps),"deaths")))
 )                   
 
+colvec1 <- colvec1[c("parish","wills","LBoM")]  ## change order
+brkvec <- c("parish","wills","LBoM")
 plot_all <- (ggplot(dd,aes(x=time,y=deaths,colour=source))
     + geom_rect(data=oy_dat,
               aes(group=outbreak.year,
@@ -121,9 +123,9 @@ plot_all <- (ggplot(dd,aes(x=time,y=deaths,colour=source))
               alpha=0.5,colour=NA)
     + geom_line(aes(linetype=source),size=1)
     + geom_point(aes(shape=source),size=0.5)
-    + scale_shape_manual(values=c(wills=NA,parish=NA,LBoM=15))
-    + scale_linetype_manual(values=c(wills=1,parish=1,LBoM=NA))
-    + scale_colour_manual(values=colvec1)
+    + scale_shape_manual(values=c(parish=NA,wills=NA,LBoM=15),breaks=brkvec)
+    + scale_linetype_manual(values=c(parish=1,wills=1,LBoM=NA),breaks=brkvec)
+    + scale_colour_manual(values=colvec1,breaks=brkvec)
     + scale_y_log10(breaks=c(1,10,100,1000,10000))
     + labs(x="date",y="wills or deaths / 4 weeks")
     + facet_grid(.~epoch,scale="free",space="free")

### 2020-06-12 submit.pl
Almost done!

@@ -1,0 +1,10 @@
+use strict;
+use 5.10.0;
+
+while(<>){
+	s|\{.*[^{/]*/(.*)[.]pdf|{$1.eps|;
+	s|\{.*[^{/]*/(.*)[.]jpg|{$1.eps|;
+	print;
+}
+## < \includegraphics[%% height=\pixht]{WillExample1644_page86_1.eps}
+## > \includegraphics[height=\pixht]{images/WillExample1644_page86_1.pdf}

### 2020-06-13 install_pkgs.R
package loading/checking stuff

@@ -1,0 +1,12 @@
+## install all packages needed by library.R
+pkgs <- scan("required_packages", comment="#", what=character(0), quiet=TRUE)
+i1 <- installed.packages()
+pkgs <- pkgs[!pkgs %in% rownames(i1)]
+install.packages(pkgs)
+library(remotes)
+install_github("bbolker/bbmle")
+install_github("davidearn/epigrowthfit",dependencies=TRUE)
+
+
+
+

### 2020-06-13 library.R

@@ -1,35 +1,9 @@
-library(broom.mixed)
-library(colorspace)
-library(cowplot)
-library(dfoptim)
-library(directlabels)
-library(dplyr)
-library(emdbook)
-library(emmeans)
-library(epigrowthfit)
-library(ggplot2)
-library(ggstance)
-library(glmmTMB)
-library(grid)
-library(gtable)
-library(Hmisc)
-library(huxtable)
-library(lubridate)
-library(methods)
-library(multcomp)
-library(nloptr)
-library(optimx)
-library(plyr)
-library(purrr)
-library(RColorBrewer)
-library(tibble)
-library(tidyr)
-library(tikzDevice)
-library(viridis)
-library(xtable)
-
-stopifnot(
-    packageVersion("glmmTMB")>="0.2.2.9000" ## GH: bbolker/glmmTMB,
-  , packageVersion("bbmle")>="1.0.21.9001"  ## GH: bbolker/bbmle
-)
+p <- scan("required_packages", comment="#", what=character(0), quiet=TRUE)
+p <- c(p, "epigrowthfit", "bbmle")
+res <- suppressMessages(vapply(p,require,character.only=TRUE,logical(1)))
+if (any(!res)) {
+    missing <- p[!res]
+    stop("please install these missing packages: ",
+         paste(missing, collapse=", "))
+}
 

### 2020-06-13 install_pkgs.R
add weights pkg; add repos to pkg install

@@ -1,4 +1,5 @@
 ## install all packages needed by library.R
+options(repos=c(CRAN="https://cloud.r-project.org"))
 pkgs <- scan("required_packages", comment="#", what=character(0), quiet=TRUE)
 i1 <- installed.packages()
 pkgs <- pkgs[!pkgs %in% rownames(i1)]

### 2020-06-13 analysis/plots/probate.R
Separate willsr pipeline

@@ -1,0 +1,8 @@
+library(willsr)
+
+## probate dates and times in husting_wills_individual are not clean/complete
+ww <- (willsr::wills
+	%>% select(ProbateDate)
+	%>% mutate(time=lubridate::decimal_date(ProbateDate))
+)
+saveRDS(ww, file = rdsname)

### 2020-06-13 analysis/plots/probate_dates.R

@@ -8,17 +8,7 @@ args = commandArgs(trailingOnly=TRUE)
 file = pdfname
 file = paste0(rtargetname, ".tex")
 
-## FIXME: this should probably be independent
-if (require(willsr)) {
-    ## probate dates and times in husting_wills_individual are not clean/complete
-    ww <- (willsr::wills
-        %>% select(ProbateDate)
-        %>% mutate(time=lubridate::decimal_date(ProbateDate))
-    )
-    saveRDS(ww, file = "../../data/probate.rds")
-} else {
-    ww <- readRDS("../../data/probate.rds")
-}
+ww <- readRDS("../../data/probate.rds")
 
 ## get info on wills-written dates from epigrowthfit
 ## (some repetition from analysis/plots/timeseries.R here)

### 2020-06-13 analysis/plots/probate.R
willsr spun out into a side-track

@@ -1,4 +1,5 @@
 library(willsr)
+library(dplyr)
 
 ## probate dates and times in husting_wills_individual are not clean/complete
 ww <- (willsr::wills

### 2020-06-13 analysis/plots/seasonalgrowth.R
seasonalgrowth fix

@@ -28,9 +28,18 @@ majorSum <- sumfun(majorList,level=2)
 conv_info <- map_dfr(minorList,get_conv,.id="source")
 conv_info$conv
 
-bad <- conv_info %>% filter(conv==1)
+bad <- (conv_info
+    ## skip these fits because they are *sometimes* bad
+    %>% filter((outbreak.year=="1578" & source=="London bills") |
+               (outbreak.year=="1581" & source=="Canterbury wills") |
+               conv==1)
+)
+
 bad$msg
-stopifnot(nrow(bad)==2)  ## make sure we're only skipping 2 ...
+if (nrow(bad)>2) {
+    print(bad)
+    stop("unexpected convergence failures")
+}
 minorList[["Canterbury wills"]][["1581"]]@mle2@details
 
 ##%>% select(source,outbreak.year,conv,msg)

### 2020-06-14 analysis/plots/seasonalgrowth.R
mark LBoM/1582 as sometimes bad

@@ -32,11 +32,12 @@ bad <- (conv_info
     ## skip these fits because they are *sometimes* bad
     %>% filter((outbreak.year=="1578" & source=="London bills") |
                (outbreak.year=="1581" & source=="Canterbury wills") |
+               (outbreak.year=="1582" & source=="London bills") |
                conv==1)
 )
 
 bad$msg
-if (nrow(bad)>2) {
+if (nrow(bad)>3) {
     print(bad)
     stop("unexpected convergence failures")
 }

### 2020-06-18 supp.Rnw
more detail on link functions; three vs two bad fits

@@
 results (highest log likelihood) for point estimates.
-In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, but we found that the scaled-tangent link proposed by \cite{Ma+14} ($\eta = \tan((x/2-1) \cdot \pi/2)$) declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).
+In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, i.e.
+fitting most of the parameters (growth rate $r$, background mortality/wills rate $b$, final size $K$, Richards shape parameter $s$) on the log scale with a scaled-tangent link ($\eta = \tan(\pi/2 \cdot (x/2-1))$) for the initial condition $x_0$.
+However, we found that the scaled-tangent link declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).
 
-With this approach we failed to achieve good (convergent) fits for only two combinations of source and outbreak year, both for minor outbreaks (London bills for 1578, Canterbury wills for 1581); these cases are excluded from the tables and figures shown below.
+With this approach we failed to achieve good (convergent) fits for only three combinations of source and outbreak year, all for minor outbreaks (London bills for 1578 and 1582, Canterbury wills for 1581); these cases are excluded from the tables and figures shown below.
 
@@
 %%\clearpage
-%%\bibliographystyle{vancouver}
-%%\bibliographystyle{Science}
 \bibliographystyle{pnas-new}

### 2020-06-19 supp.Rnw
two clarifications in generation interval section

@@
 Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval.
-However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next epidemic cycle becomes infectious is:
-
+However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
+%%
 \begin{equation}
@@
 \end{equation}
-
-From \cite{LaPerriere+2009}, we can gather that
+%%
+The second line above takes account of the fact that a flea leaves
+its rat host when, and only when, the rat dies.  From
+\cite{LaPerriere+2009}, we can gather that
 \begin{itemize}
@@
 \end{itemize}
-
+%%
 Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.
@@
 
-One factor that might have contributed to the observed increase in
-epidemic growth rates in London is the density of rats.  We do not
-have data that allow us to estimate rat densities in Medieval England,
-but we can ask---all else being equal---by what factor would rat
-density have to have changed in order to account for the observed
-change in growth rates?
+One factor that might have contributed to the observed increase in epidemic growth rates in London is the density of rats.
+We do not have data that allow us to estimate rat densities in Medieval England, but we can ask---all else being equal---by what factor would rat density have to have changed in order to account for the observed change in growth rates?
 

### 2020-06-19 ms_text.tex
Main MS; nothing pulled in except by chaining (there should be few or no excess files)
more natural affiliation order

@@
 
-\author{David J.\,D.\ Earn$^{a,c,e}$, 
+\author{David J.\,D.\ Earn$^{a,b,c}$, 
 Junling Ma$^{f}$, 
 Hendrik N.\ Poinar$^{b,c,d,e}$,\\
-Jonathan Dushoff$^{a,c,e}$,
-and Benjamin M.\ Bolker$^{a,c,e}$\\
+Jonathan Dushoff$^{a,b,c}$,
+and Benjamin M.\ Bolker$^{a,b,c}$\\
 \href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
-\small $^a$Department of Mathematics \& Statistics,\\
-\small $^b$McMaster Ancient DNA Centre, Department of Anthropology,\\
-\small $^c$Department of Biology, $^d$Department of Biochemistry,\\
-\small and $^e$Michael G.\ deGroote Institute for Infectious Disease Research,\\
+\small $^a$Department of Mathematics \& Statistics, $^b$Department of Biology, \\
+\small $^c$Michael G.\ deGroote Institute for Infectious Disease Research,\\
+\small $^d$McMaster Ancient DNA Centre, Department of Anthropology,\\
+\small and $^e$Department of Biochemistry,\\
 \small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
@@
 \\ \vspace{0.5cm}
-\includegraphics[width=\textwidth]{analysis/plots/plot_all.pdf}
-%% \includegraphics[width=\textwidth]{plot_all.eps}
+\includegraphics[width=\textwidth]{analysis/plots/plot_allsources.pdf}
+%% \includegraphics[width=\textwidth]{plot_allsources.eps}
 \end{center}
@@
   \rowcolors{2}{white}{gray}
-  \input{analysis/tables/fakeparams.tex}
+  \input{analysis/tables/params.tex}
 
@@
 Valerie Hart facilitated work at the Guildhall Library in London.
-Josephine Earn provided historical context.
 Non-exponential comparison fits were conducted by David Champredon.
@@
 SHARCnet (\url{http://www.sharcnet.ca}) provided computational resources.
+David Earn dedicates this paper to Josephine Earn, who enthusiastically discussed the history of London (and plagues) with him over many years.

### 2020-07-03 ms_text.tex
additional keywords
amsync

@@
 
-\leftline{plague $|$ London $|$ epidemic growth rate}
+\leftline{plague $|$ London $|$ epidemic growth rate $|$ reproduction number $|$ SARS-CoV-2 $|$ COVID-19}
 
@@
 When available, the LBoM are our most reliable records (except for the epidemic of 1593; see \supp).
-\item[Parish registers] After 1538 \cite[p.\,54]{Slac90}, London parish registers provide information on mortality.
+\item[Parish registers] After 1538 \cite[p.\,54]{Slac90}\cite{Wall04}, London parish registers provide information on mortality.
 %%, although these are not categorized by cause. 

### 2020-07-04 ms_text.tex
minor typos

@@
 Currently available data do not provide enough information to infer the mode of plague transmission in any given epidemic; nevertheless, order-of-magnitude estimates of epidemic parameters suggest that the observed slow growth rates in the 14th century are inconsistent with direct (pneumonic) transmission.
-We discuss the potential roles of demographic and ecological factors such as climate change or human or rat population density in driving the observed acceleration.
+We discuss the potential roles of demographic and ecological factors, such as climate change or human or rat population density, in driving the observed acceleration.
 }  
@@
 We can use the distribution of the generation interval (the times elapsed between the moment a host is infected and the times at which they infect others) to compute $\rzero$ from $r$ \cite{WallLips07}.
-If we only know the \emph{mean} generation interval $\gentime$, we can still approximate $\rzero \approx r \gentime +1$, by assuming that the generation intervals are exponentially distributed; this approximation underestimates $\rzero$ in the typical case where the generation interval distribution is less variable than an exponential distribution with the same mean.
+If we know only the \emph{mean} generation interval $\gentime$, we can still approximate $\rzero \approx r \gentime +1$, by assuming that the generation intervals are exponentially distributed; this approximation underestimates $\rzero$ in the typical case where the generation interval distribution is less variable than an exponential distribution with the same mean.
 
@@
 The growth rates increased \foldval in the late \emph{vs.}\ early epochs: $r$ increased \rdiffest $\times$ [95\% CI $(\rdifflwr\times, \rdiffupr\times)$] (see \Methodslink).
-The permutation test results support the statistical significance of this difference; the observed order of outbreaks yields the most extreme difference in growth rates out of all \nperm\ possible orderings, giving a two-tailed $p$ value of $\permfrac=\permpval$.
+The permutation test results support the statistical significance of this difference; the observed order of outbreaks yields the most extreme difference in growth rates out of all \nperm\ possible orderings, giving a two-tailed $p$-value of $\permfrac=\permpval$.
 
@@
 Why the later plague epidemics were faster than the earlier ones is not clear.
-Nevertheless, to provide some context, we consider below a number of mechanisms which \emph{could} have contributed to increases in epidemic growth rates.
+Nevertheless, to provide some context, we consider below a number of mechanisms that \emph{could} have contributed to increases in epidemic growth rates.
 We then consider what our estimated epidemic growth rates suggest about the mode of transmission during the various London plague epidemics.
@@
 reconciling the predicted attack rates with observed mortality would require strong effects of heterogeneity; a large fraction of individuals resistant to plague due to prior non-fatal plague infections; a large proportion of asymptomatic or sub-clinical infections; or a decrease in transmission rate as the plague spread (perhaps due to behavioural avoidance of transmission).
-The effects of heterogeneity and behavioural avoidance may indeed have been stronger in the late epoch, as a result of concentration of the wealthy in central London and tendency for them to flee London during plague epidemics \cite[p.\,4]{Cumm+16}.
+The effects of heterogeneity and behavioural avoidance may indeed have been stronger in the late epoch, as a result of concentration of the wealthy in central London and a tendency for them to flee London during plague epidemics \cite[p.\,4]{Cumm+16}.
 While we still consider direct (pneumonic) transmission in  late-epoch London plague epidemics unlikely because of the strong effects that would be needed to reduce the expected mortality rates to the observed levels, we cannot rule it out.

### 2020-07-05 ms_text.tex
minor changes during 'final' proofread

@@
 By considering the additional (extreme) assumptions that \hypertarget{4assumptions}{} (1) the population was homogeneously mixed; (2) the population was initially 100\% susceptible; (3) there were no asymptomatic infections; and (4) all infected individuals died (\ie the CFP was 100\%), all of which increase the observed attack rate, we can calculate an upper bound on the total mortality we would expect to observe in an epidemic with a particular transmission mode.
+%%
+\david{I meant that if CFP$=1$ then all infected individuals get
+  detected by mortality statistics, hence this would yield larger
+  observed attack rate than if CFP$<1$.  Not sure this is
+  sufficiently clear.}
 
@@
 For wills data, the delay could even be negative, \ie wills could have been written out of fear of \emph{future} infection.
-Furthermore, we found little difference between the Richards and logistic fits---which Ma \etal found to have opposite biases; hence these methods are likely to bracket the growth rate estimated by the delayed logistic.
+Furthermore, because we found little difference between the Richards and logistic fits---which Ma \etal found to have opposite biases---growth rates estimated by these two methods are likely to bracket the growth rate estimated by the delayed logistic.
 
@@
 The fitting windows used for all our fits are listed in \supp \cref{tab:windows,tab:windowsminor}.
-Although we experimented with different methods for determining the fitting window (e.g.
-using a threshold of 1/100 of the peak mortality rate, or using the observation after the last local minimum before the peak),
-we emphasize that our choices were entirely determined by seeking ``good'' fits (judged by visual inspection), and not by the pattern of estimated initial growth rates.
+Although we experimented with different methods for determining the fitting window (e.g., starting the window from the observation after the last local minimum before the peak, or based on a threshold of $1/100$ of the peak mortality rate), we emphasize that our choices were entirely determined by seeking ``good'' fits (judged by visual inspection), and not by the pattern of estimated initial growth rates.
 Moreover, our conclusion that later plagues were much faster than the earlier plagues is robust to choices we made about window selection.

### 2020-07-05 ms_text.tex
improved (hopefully) last paragraph of Approach section

@@
 We can use independent estimates of the generation interval to estimate $\rzero$ (and hence the theoretical attack rate) for London epidemics under different transmission modes.
-By considering the additional (extreme) assumptions that \hypertarget{4assumptions}{} (1) the population was homogeneously mixed; (2) the population was initially 100\% susceptible; (3) there were no asymptomatic infections; and (4) all infected individuals died (\ie the CFP was 100\%), all of which increase the observed attack rate, we can calculate an upper bound on the total mortality we would expect to observe in an epidemic with a particular transmission mode.
-%%
-\david{I meant that if CFP$=1$ then all infected individuals get
-  detected by mortality statistics, hence this would yield larger
-  observed attack rate than if CFP$<1$.  Not sure this is
-  sufficiently clear.}
+Given the limited information that we have, we consider the additional (extreme) assumptions that \hypertarget{4assumptions}{} (1) the population was homogeneously mixed; (2) the population was initially 100\% susceptible; (3) there were no asymptomatic infections; and (4) all infected individuals died (\ie the CFP was 100\%).
+All of these assumptions increase the observed attack rate (as detected from wills or deaths), so we can calculate an upper bound on the total mortality we would expect to observe in an epidemic with a particular transmission mode.
 

### 2020-07-08 ms_text.tex
changed are apparently'' to appear to be''

@@
 In 1563 and 1593, the number of PCC wills is too small to extract a signal from the noise, but the data for the later three outbreaks verify that the numbers of wills written follow the plague epidemic patterns evident in the mortality records (\cref{F:timeseries}).
-Based on cross-correlation analysis and differences between epidemic peak times, will counts appear to have lagged mortality recorded in parish registers and LBoM by approximately 3--5 weeks; our inferences about epidemic growth rates are apparently robust to this lag (see \supp).
+Based on cross-correlation analysis and differences between epidemic peak times, will counts appear to have lagged mortality recorded in parish registers and LBoM by approximately 3--5 weeks; our inferences about epidemic growth rates appear to be robust to this lag (see \supp).
+%%
+\ben{I *think* all I meant (no longer 100\% sure) is that, since we get reasonably consistent numbers across the different record types, that suggests that lag doesn't have a big effect.
+This is of course a weak argument, as we could imagine that we would get different growth rates from the different record types *if* they were not differently lagged, and the lag effects are canceling out the true differences.}
+\david{changed ``are apparently'' to ``appear to be''}
 

### 2020-07-09 analysis/plots/plot_all.R
Revert "change line order in legend"
This reverts commit 2ab15d1fc2b16571498c623a886873156c7265e6.

@@ -111,8 +111,6 @@ event_tab2 <- (event_tab
                                   time<=end+time_eps),"deaths")))
 )                   
 
-colvec1 <- colvec1[c("parish","wills","LBoM")]  ## change order
-brkvec <- c("parish","wills","LBoM")
 plot_all <- (ggplot(dd,aes(x=time,y=deaths,colour=source))
     + geom_rect(data=oy_dat,
               aes(group=outbreak.year,
@@ -123,9 +121,9 @@ plot_all <- (ggplot(dd,aes(x=time,y=deaths,colour=source))
               alpha=0.5,colour=NA)
     + geom_line(aes(linetype=source),size=1)
     + geom_point(aes(shape=source),size=0.5)
-    + scale_shape_manual(values=c(parish=NA,wills=NA,LBoM=15),breaks=brkvec)
-    + scale_linetype_manual(values=c(parish=1,wills=1,LBoM=NA),breaks=brkvec)
-    + scale_colour_manual(values=colvec1,breaks=brkvec)
+    + scale_shape_manual(values=c(wills=NA,parish=NA,LBoM=15))
+    + scale_linetype_manual(values=c(wills=1,parish=1,LBoM=NA))
+    + scale_colour_manual(values=colvec1)
     + scale_y_log10(breaks=c(1,10,100,1000,10000))
     + labs(x="date",y="wills or deaths / 4 weeks")
     + facet_grid(.~epoch,scale="free",space="free")

### 2020-07-09 ms_text.tex
attempting to update make for plot_all.pdf

@@
 Based on cross-correlation analysis and differences between epidemic peak times, will counts appear to have lagged mortality recorded in parish registers and LBoM by approximately 3--5 weeks; our inferences about epidemic growth rates appear to be robust to this lag (see \supp).
-%%
-\ben{I *think* all I meant (no longer 100\% sure) is that, since we get reasonably consistent numbers across the different record types, that suggests that lag doesn't have a big effect.
-This is of course a weak argument, as we could imagine that we would get different growth rates from the different record types *if* they were not differently lagged, and the lag effects are canceling out the true differences.}
-\david{changed ``are apparently'' to ``appear to be''}
 
@@
 \\ \vspace{0.5cm}
-\includegraphics[width=\textwidth]{analysis/plots/plot_allsources.pdf}
+\includegraphics[width=\textwidth]{analysis/plots/plot_all.Routpdf}
 %% \includegraphics[width=\textwidth]{plot_allsources.eps}

### 2020-07-11 supp.Rnw
ref to KeelGill00 for plague natural history
hyperlink

@@
 
+\hypertarget{1593}{}
 \paragraph{London 1593:} The extant records of weekly plague mortality in London in 1593 (\cref{F:plot3sourcesmajors}, second panel) are implausible.
@@
 Epidemics with peak plague mortality above 5 per 1000 individuals per week were classified as \emph{major}.
-With the exception of 1593 (for which we do not use LBoM data; see above), all these major epidemics peak above 14 on this scale; all the other (\emph{minor}) epidemics peak below 3.
+With the exception of 1593 (for which we do not use LBoM data; see \hyperlink{1593}{above}), all these major epidemics peak above 14 on this scale; all the other (\emph{minor}) epidemics peak below 3.
 The raw plague mortality data are shown in \cref{F:all1600s}.
@@
 The second line above takes account of the fact that a flea leaves
-its rat host when, and only when, the rat dies.  From
+its rat host when, and only when, the rat dies \cite{KeelGill00}.  From
 \cite{LaPerriere+2009}, we can gather that

### 2020-07-11 supp.Rnw
a few minor tweaks / temporary comments

@@
 
-For pneumonic plague, sufficiently detailed data exist for a number of modern outbreaks to allow an estimate of the latent and infectious periods (and hence of the generation interval distribution) \cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given value of $r$.
+For pneumonic plague, sufficiently detailed data exist for a number of
+modern outbreaks to allow an estimate of the latent and infectious
+periods (and hence of the generation interval distribution)
+\cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given
+value of $r$ \cite{WallLips07}.
 Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval.
@@
 bound to the true relationship, indicating that rat density would have
-had to increase by at least a factor of  to account for a similar
-increase in $\R_0$.
+had to increase by at least a factor of \david{???missing number???}
+to account for a similar increase in $\R_0$.
 
@@
 rat density would be sufficient to yield an increase in $\R_0$ by a
-factor of \Sexpr{round(R0ratio,1)} (or any factor detectably greater than 1).
+factor of \Sexpr{round(R0ratio,2)} (or any factor detectably greater
+than 1).
 
@@
 In the rat-flea model, $\rzero$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
-From Keeling and Gilligan's plague model \cite{KeelGill00b}, the expected change in $r$ for a change in flea-rat ratio from $K_{F0}$ to $K_F$ is $(K_F/K_{F0})^S$, where $S$ is the sensitivity ($\approx 1.5$).
-Thus in order to a see a fivefold increase in $r$ we would need a change of $5^{(1/S)} \approx \Sexpr{round(5^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.
+From Keeling and Gilligan's plague model \cite{KeelGill00b}, the
+expected change in $r$ for a change in flea-rat ratio from $K_{\rm
+  F0}$ to $K_{\rm F}$ is $(K_{\rm F}/K_{\rm F0})^S$, where $S$ is the
+sensitivity ($\approx 1.5$). Thus in order to a see a fivefold
+\david{why FIVEfold?  we need only FOURfold; didn't we have a macro
+  for this?} increase in $r$ we would need a change of $5^{(1/S)} \approx \Sexpr{round(5^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.
 
@@
   Mortality (LBoM) records and other sources. All lags measured in
-  weeks.  Rows corresponding to wills data are highlighted in grey.}
+  weeks.  Rows corresponding to wills data are highlighted in grey.
+  \david{why is the row shading wrong?}
+}
 \begin{center}
@@
 results (highest log likelihood) for point estimates.
-In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, i.e.
-fitting most of the parameters (growth rate $r$, background mortality/wills rate $b$, final size $K$, Richards shape parameter $s$) on the log scale with a scaled-tangent link ($\eta = \tan(\pi/2 \cdot (x/2-1))$) for the initial condition $x_0$.
+In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, i.e., fitting most of the parameters (growth rate $r$, background mortality/wills rate $b$, final size $K$, Richards shape parameter $s$) on the log scale with a scaled-tangent link ($\eta = \tan(\pi/2 \cdot (x/2-1))$) for the initial condition $x_0$.
 However, we found that the scaled-tangent link declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).

### 2020-07-11 supp.Rnw
corrected foldval and calculation of fleas-per-rat requirement

@@
   F0}$ to $K_{\rm F}$ is $(K_{\rm F}/K_{\rm F0})^S$, where $S$ is the
-sensitivity ($\approx 1.5$). Thus in order to a see a fivefold
-\david{why FIVEfold?  we need only FOURfold; didn't we have a macro
-  for this?} increase in $r$ we would need a change of $5^{(1/S)} \approx \Sexpr{round(5^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.
+sensitivity ($\approx 1.5$). Thus in order to a see a \foldval
+increase in $r$ we would need a change of $4^{(1/S)} \approx \Sexpr{round(4^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.
 

### 2020-07-11 supp.Rnw
minor repairs

@@
 bound to the true relationship, indicating that rat density would have
-had to increase by at least a factor of \david{???missing number???}
-to account for a similar increase in $\R_0$.
+had to increase by at least a factor of \Sexpr{round(R0ratio,2)}
+to account for a similar increase \eqref{E:Rratio} in $\R_0$.
 
@@
   weeks.  Rows corresponding to wills data are highlighted in grey.
-  \david{why is the row shading wrong?}
 }

### 2020-07-12 ms_text.tex
fix typo
still struggling
include pdf rather than jpg
compiles without error with PNAS style

@@
 
-\author{David J.\,D.\ Earn$^{a,b,c}$, 
-Junling Ma$^{f}$, 
-Hendrik N.\ Poinar$^{b,c,d,e}$,\\
-Jonathan Dushoff$^{a,b,c}$,
-and Benjamin M.\ Bolker$^{a,b,c}$\\
-\href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
-\small $^a$Department of Mathematics \& Statistics, $^b$Department of Biology, \\
-\small $^c$Michael G.\ deGroote Institute for Infectious Disease Research,\\
-\small $^d$McMaster Ancient DNA Centre, Department of Anthropology,\\
-\small and $^e$Department of Biochemistry,\\
-\small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
-\small $^f$Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}
+% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+\author[a,b,c,1]{David J.\,D.\ Earn}
+\author[f]{Junling Ma} 
+\author[b,c,d,e]{Hendrik Poinar}
+\author[a,b,c]{Jonathan Dushoff}
+\author[a,b,c]{Benjamin M.\ Bolker}
+
+\affil[a]{Department of Mathematics \& Statistics}
+\affil[b]{Department of Biology}
+\affil[c]{Michael G.\ deGroote Institute for Infectious Disease Research}
+\affil[d]{McMaster Ancient DNA Centre, Department of Anthropology}
+\affil[e]{Department of Biochemistry, McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada}
+\affil[f]{Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}
+
+% Please give the surname of the lead author for the running footer
+\leadauthor{Earn}
+
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{Epidemics of plague devastated Europe's population throughout the Medieval and Renaissance periods.
+Genetic studies of small numbers of skeletal remains indicate that the causative agent of all these epidemics was the bacterium \emph{Yersinia pestis}, but such analyses cannot identify overall patterns of transmission dynamics.
+Analysis of thousands of archival records from London, England, reveal that plague epidemics spread much faster in the 17th century than in the 14th century.}
+
+% Please include corresponding author, author contribution and author declaration information
+\authorcontributions{{\color{red}Please provide details of author contributions here.}}
+%%\authordeclaration{{\color{red}Please declare any conflict of interest here.}}
+%%\equalauthors{\textsuperscript{1}A.O.(Author One) and A.T. (Author Two) contributed equally to this work (remove if not applicable).}
+\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: earn@math.mcmaster.ca}
+
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+\keywords{plague $|$ London $|$ epidemic growth rate $|$ reproduction number $|$ SARS-CoV-2 $|$ COVID-19}
+
+\begin{abstract}
+Historical records reveal the temporal patterns of a sequence of plague epidemics in London, England, from the 14th to 17th centuries.
+Analysis of these records shows for the first time that later epidemics spread significantly faster (``accelerated'').
+Between the Black Death of 1348 and the later epidemics that culminated with the Great Plague of 1665, we estimate that the epidemic growth rate increased \foldval.
+Currently available data do not provide enough information to infer the mode of plague transmission in any given epidemic; nevertheless, order-of-magnitude estimates of epidemic parameters suggest that the observed slow growth rates in the 14th century are inconsistent with direct (pneumonic) transmission.
+We discuss the potential roles of demographic and ecological factors, such as climate change or human or rat population density, in driving the observed acceleration.
+\end{abstract}
+
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
@@
 
-\linenumbers
+\maketitle
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
 
-\maketitle
-
-%%\begin{abstract}
-{\bfseries
-\noindent
-Historical records reveal the temporal patterns of a sequence of plague epidemics in London, England, from the 14th to 17th centuries.
-Analysis of these records shows for the first time that later epidemics spread significantly faster (``accelerated'').
-Between the Black Death of 1348 and the later epidemics that culminated with the Great Plague of 1665, we estimate that the epidemic growth rate increased \foldval.
-%%[\rdiffest $\times$, 95\% CI: \rdifflwr $\times$ to \rdiffupr $\times$].  
-%%The cause of this change is unknown; despite rapid recent advances in paleogenomics, it is hard to confirm or reject the possibility of functional evolutionary changes in circulating strains of the plague bacterium \ypestis over this time period.
-Currently available data do not provide enough information to infer the mode of plague transmission in any given epidemic; nevertheless, order-of-magnitude estimates of epidemic parameters suggest that the observed slow growth rates in the 14th century are inconsistent with direct (pneumonic) transmission.
-We discuss the potential roles of demographic and ecological factors, such as climate change or human or rat population density, in driving the observed acceleration.
-}  
-%%\end{abstract}
-
-\bigskip
-
-\leftline{plague $|$ London $|$ epidemic growth rate $|$ reproduction number $|$ SARS-CoV-2 $|$ COVID-19}
-
-\bigbreak
-
-\noindent\fbox{\parbox{0.98\textwidth}{{\bfseries Significance Statement:}
-
-Epidemics of plague devastated Europe's population throughout the Medieval and Renaissance periods.
-Genetic studies of small numbers of skeletal remains indicate that the causative agent of all these epidemics was the bacterium \emph{Yersinia pestis}, but such analyses cannot identify overall patterns of transmission dynamics.
-Analysis of thousands of archival records from London, England, reveal that plague epidemics spread much faster in the 17th century than in the 14th century.}}
-
-\bigbreak
-
+% If your first paragraph (i.e.
+with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
+If this is the case, add \parshape=0 to the end of the list environment.
 \dropcap{P}lague epidemics have afflicted human populations since at least the 6th century \cite{McNe76,Wagn+14}.
@@
 
-\mysection{Data}
+\section*{Data}
 
@@
 \newcommand{\pixht}{0.3\textheight}
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 \hspace{0.4cm}
-\includegraphics[height=\pixht]{images/cripplegate_wellcome_parish.jpg}
+\includegraphics[height=\pixht]{images/cripplegate_wellcome_parish.pdf}
 %% \includegraphics[height=\pixht]{cripplegate_wellcome_parish.eps}
@@
 \\ \vspace{0.5cm}
-\includegraphics[width=\textwidth]{analysis/plots/plot_all.Routpdf}
+\includegraphics[width=\textwidth]{analysis/plots/plot_allsources.pdf}
 %% \includegraphics[width=\textwidth]{plot_allsources.eps}
@@
 \label{F:plot3sources}
-\end{figure}
+\end{figure*}
 %
 
-\mysection{Approach}
+\section*{Approach}
 
@@
 While our collection of high-resolution mortality time series during plague epidemics allows us to estimate epidemic growth rates, we are constrained by the limitations of our data---the time series for each epidemic are short and noisy, especially during the 14th century (\eg the epidemics of 1368 and 1375 peaked at a total of 5 wills written per week).
-Thus we chose to use simple phenomenological models with a small number of parameters that can be fitted to short time series (starting size, growth rate, and total size: see \Methodslink), and do not attempt to separate different sources of variability or estimate the influence of various mechanistic processes (King \etal's exhortation to use epidemic models that separate process from observation noise \cite{King+2015} applies primarily to forecasting, which we are not attempting).
-Furthermore, we primarily consider the initial growth rate $r$, which---unlike the basic reproduction number $\rzero$---can be estimated reliably from mortality data alone without any knowledge of the disease characteristics or natural history \cite{Ma+14}.
+Thus we chose to use simple phenomenological models with a small number of parameters that can be fitted to short time series (starting size, growth rate, and total size: see \Methodslink), and do not attempt to separate different sources of variability or estimate the influence of various mechanistic processes (King \ETAL's exhortation to use epidemic models that separate process from observation noise \cite{King+2015} applies primarily to forecasting, which we are not attempting).
+Furthermore, we primarily consider the initial growth rate $r$, which---unlike the basic reproduction number $\R_0$---can be estimated reliably from mortality data alone without any knowledge of the disease characteristics or natural history \cite{Ma+14}.
 
@@
 
-\paragraph{Estimating $\rzero$ and attack rate}
+\paragraph{Estimating $\R_0$ and attack rate}
 
-Our analysis focuses on the epidemic growth rate $r$, rather than the more commonly investigated basic reproduction number $\rzero$ (the expected number of secondary cases caused by a primary case in a completely susceptible population \cite{AndeMay91}), because estimates of $\rzero$ depend on information about the life history of the pathogen as well as on the epidemic curve itself \cite{WallLips07}.
-However, if we can estimate $\rzero$ then we can predict epidemic outcomes that we are unable to predict from $r$ alone; then, by comparing observed with theoretical outcomes under different scenarios, we can make inferences about modes of transmission.
+Our analysis focuses on the epidemic growth rate $r$, rather than the more commonly investigated basic reproduction number $\R_0$ (the expected number of secondary cases caused by a primary case in a completely susceptible population \cite{AndeMay91}), because estimates of $\R_0$ depend on information about the life history of the pathogen as well as on the epidemic curve itself \cite{WallLips07}.
+However, if we can estimate $\R_0$ then we can predict epidemic outcomes that we are unable to predict from $r$ alone; then, by comparing observed with theoretical outcomes under different scenarios, we can make inferences about modes of transmission.
 
-In particular, if the host population is homogeneously mixed, we can use $\rzero$ to estimate the expected final size $Z$ (the proportion of the population infected by the end of an epidemic, also called the \emph{attack rate}) \cite{KermMcKe27,MaEarn06}.
+In particular, if the host population is homogeneously mixed, we can use $\R_0$ to estimate the expected final size $Z$ (the proportion of the population infected by the end of an epidemic, also called the \emph{attack rate}) \cite{KermMcKe27,MaEarn06}.
 The \emph{observed attack rate} based on mortality records depends in turn on the fraction of infected individuals who die from disease.
@@
 
-We can use the distribution of the generation interval (the times elapsed between the moment a host is infected and the times at which they infect others) to compute $\rzero$ from $r$ \cite{WallLips07}.
-If we know only the \emph{mean} generation interval $\gentime$, we can still approximate $\rzero \approx r \gentime +1$, by assuming that the generation intervals are exponentially distributed; this approximation underestimates $\rzero$ in the typical case where the generation interval distribution is less variable than an exponential distribution with the same mean.
+We can use the distribution of the generation interval (the times elapsed between the moment a host is infected and the times at which they infect others) to compute $\R_0$ from $r$ \cite{WallLips07}.
+If we know only the \emph{mean} generation interval $\gentime$, we can still approximate $\R_0 \approx r \gentime +1$, by assuming that the generation intervals are exponentially distributed; this approximation underestimates $\R_0$ in the typical case where the generation interval distribution is less variable than an exponential distribution with the same mean.
 
-Beyond the information on generation interval that we use to derive $\rzero$ from $r$, the observed attack rate depends on additional characteristics of an epidemic.
+Beyond the information on generation interval that we use to derive $\R_0$ from $r$, the observed attack rate depends on additional characteristics of an epidemic.
 If the population does not mix homogeneously, the final size will typically decrease (\eg \cite[Fig.\,6]{Roll+15}).
@@
 
-We can use independent estimates of the generation interval to estimate $\rzero$ (and hence the theoretical attack rate) for London epidemics under different transmission modes.
+We can use independent estimates of the generation interval to estimate $\R_0$ (and hence the theoretical attack rate) for London epidemics under different transmission modes.
 Given the limited information that we have, we consider the additional (extreme) assumptions that \hypertarget{4assumptions}{} (1) the population was homogeneously mixed; (2) the population was initially 100\% susceptible; (3) there were no asymptomatic infections; and (4) all infected individuals died (\ie the CFP was 100\%).
@@
 
-\mysectionsamepage{Results}
+\section*{Results}
 
@@
 %%\begin{figure}[h!]
-\begin{figure}
+\begin{figure*}
 \begin{center}
@@
 \label{F:timeseries}
-\end{figure}
+\end{figure*}
 %
@@
 %%\begin{figure}[h!]
-\begin{figure}
+\begin{figure*}
 \begin{center} 
@@
 Center panels show mixed-model estimates (see \Methodslink) of overall average growth rate for each epoch (early:\ 14th c.\ \emph{vs.}\ late:\ 16th-17th c.{})\ and data source.
-To aid interpretation, the additional vertical scales show the implied intrinsic reproductive number ($\rzerop$) and epidemic final size ($\Zp$) under the assumption that the generation interval distribution \cite{WallLips07} was the same as that estimated from 20th century pneumonic plague epidemics \cite{GaniLeac04}.
+To aid interpretation, the additional vertical scales show the implied intrinsic reproductive number ($\R_0p$) and epidemic final size ($\Zp$) under the assumption that the generation interval distribution \cite{WallLips07} was the same as that estimated from 20th century pneumonic plague epidemics \cite{GaniLeac04}.
 All estimates and CIs are listed in \cref{tab:params} (which also lists the associated doubling times).}
 \label{F:combdata}
-\end{figure}
+\end{figure*}
 %
 
-\begin{table}[h!] 
+%%\begin{table}[h!] 
+\begin{table*}
   \caption[Estimated $r$ and $1/r$]{Maximum likelihood estimates (MLEs) of the initial exponential growth rate ($r$) and doubling time ($(\log2)/r$) with their 95\% confidence intervals (CIs), obtained from the time series shown in \cref{F:timeseries} (see \Methodslink).
 The goodness of fit measure ($R^2$) is the proportional reduction in the mean squared error, \ie $1-d^2/\sigma^2$ where $d^2$ is the model mean squared error and $\sigma^2$ is the (population) variance of the data; predictions and observations are aggregated to weekly before computing the $R^2$ for wills, for consistency among data sources.
-The implied basic reproduction numbers ($\rzerop$) and attack rates ($\Zp$), assuming the generation interval distribution for Modern pneumonic plague \cite{GaniLeac04}, are also shown.
+The implied basic reproduction numbers ($\R_0p$) and attack rates ($\Zp$), assuming the generation interval distribution for Modern pneumonic plague \cite{GaniLeac04}, are also shown.
 All MLEs and associated CIs are shown in \cref{F:combdata}.
@@
   \footnotesize
-  %%\rowcolors{2}{white}{verylightgray}
-  \rowcolors{2}{white}{gray}
+  \rowcolors{2}{white}{verylightgray}
   \input{analysis/tables/params.tex}
@@
 \label{tab:params}
-\end{table}
+\end{table*}
 
-\FloatBarrier
-\mysection{Discussion}
+\section*{Discussion}
 
@@
 As mentioned above, it is possible that human ectoparasites could have supported indirect plague transmission \cite{Dean+18,Laro+18}.
-While the biological details of this transmission mode are still uncertain, we can guess that the generation interval of this transmission mode would be intermediate between direct and indirect rat-flea transmission, giving rise to estimates of $\rzero$ and $Z$ intermediate between those discussed above.
-Investigations of plague biology in fleas have also raised the possibility that rat transmission may have shorter generation intervals than previously thought, which would lower estimated $\rzero$ values for bubonic plague \cite{Eise+06,Eise+07} (but cf.~\cite{Hinn+17}).
+While the biological details of this transmission mode are still uncertain, we can guess that the generation interval of this transmission mode would be intermediate between direct and indirect rat-flea transmission, giving rise to estimates of $\R_0$ and $Z$ intermediate between those discussed above.
+Investigations of plague biology in fleas have also raised the possibility that rat transmission may have shorter generation intervals than previously thought, which would lower estimated $\R_0$ values for bubonic plague \cite{Eise+06,Eise+07} (but cf.~\cite{Hinn+17}).
 
@@
 \hypertarget{Methods}{}
-\mysectionsamepage{Methods}
+%%\section*{Methods}
+\matmethods{
 
-\paragraph{Growth rate estimates}
+\subsection*{Growth rate estimates}
 
@@
 All variation around the fitted line is assumed to arise from  observation errors.
-See Ma \etal \cite{Ma+14} for methodological details.
+See Ma \ETAL \cite{Ma+14} for methodological details.
 
-In contrast to Ma \etal \cite{Ma+14}, who assumed a Poisson error model to fit simulations that were themselves generated with that assumption, the results shown here applied a negative binomial error model, which allows for the broader range of variation found in real data.
+In contrast to Ma \ETAL \cite{Ma+14}, who assumed a Poisson error model to fit simulations that were themselves generated with that assumption, the results shown here applied a negative binomial error model, which allows for the broader range of variation found in real data.
 We initially tried both logistic and Richards \cite{Rich59} models for the epidemic curve (\ie the expected mean of the data) and both Poisson and negative binomial models for the distribution around the mean.
@@
 
-In order to summarize the differences between epochs, we fitted a linear mixed model (using the \code{glmmTMB} package in R \cite{Brook+17}) to the estimates of log growth rates.
+In order to summarize the differences between epochs, we fitted a linear mixed model (using the \texttt{glmmTMB} package in R \cite{Brook+17}) to the estimates of log growth rates.
 To account for differing precision of the estimates across epidemics, we approximate the standard error of the estimates as the width of the 95\% confidence interval divided by 3.92 (the width in standard errors of a 95\% Normal CI); square it to find the variance; and scale the residual variance for each epidemic curve by this observed variance.
@@
 \hypertarget{phenmod}{}
-\paragraph{Phenomenological model}
+\subsection*{Phenomenological model}
 
-Following Ma \etal \cite{Ma+14}, we model the \emph{cumulative} mortality curve (or will-accumulation curve; henceforth we refer to wills or death records as ``mortality'') using a logistic model with an added baseline linear growth rate,
+Following Ma \ETAL \cite{Ma+14}, we model the \emph{cumulative} mortality curve (or will-accumulation curve; henceforth we refer to wills or death records as ``mortality'') using a logistic model with an added baseline linear growth rate,
 %
@@
 \begin{equation}
-c(t) = b t + \frac{K}{\big[1 + \big((K/c_0) - 1\big)e^{-rt}\big]} \,.
+%%c(t) = b t + \frac{K}{\big[1 + \big((K/c_0) - 1\big)e^{-rt}\big]} \,.
+c(t) = b t + \frac{K}{1 + [(K/c_0) - 1]e^{-rt}} \,.
 \label{E:Richards}
@@
 As discussed above, we initially fitted a Richards model, which extends the logistic equation with a shape parameter, but found that we could not get stable fits to the smallest/noisiest epidemics; we never found qualitative differences between growth rates estimated from logistic and Richards fits.
-Ma \etal found that a \emph{delayed} logistic curve, which adds an exponentially distributed delay between infection and mortality, estimated growth rates from mortality curves with less bias.
+Ma \ETAL found that a \emph{delayed} logistic curve, which adds an exponentially distributed delay between infection and mortality, estimated growth rates from mortality curves with less bias.
 However, the delay between infection and mortality in our data is extremely uncertain---it would represent another parameter that would have to be estimated from noisy data.
 For wills data, the delay could even be negative, \ie wills could have been written out of fear of \emph{future} infection.
-Furthermore, because we found little difference between the Richards and logistic fits---which Ma \etal found to have opposite biases---growth rates estimated by these two methods are likely to bracket the growth rate estimated by the delayed logistic.
+Furthermore, because we found little difference between the Richards and logistic fits---which Ma \ETAL found to have opposite biases---growth rates estimated by these two methods are likely to bracket the growth rate estimated by the delayed logistic.
 
-\paragraph{Fitting windows}
+\subsection*{Fitting windows}
 
@@
 \hypertarget{outbreak.year}{}
-\paragraph{Outbreak years and outbreak windows}
+\subsection*{Outbreak years and outbreak windows}
 
@@
 \hypertarget{data.and.code}{}
-\paragraph{Data and code availability}
+\subsection*{Data and code availability}
 
@@
 
-\mysectionsamepage{Acknowledgements}
+}% end \matmethods
 
-The plague mortality data were photographed and/or entered by Seth Earn, Kelly Hancock, Claire Lees, James McDonald, David Price and Hannah Price.
+\showmatmethods{} % Display the Materials and Methods section
+
+\acknow{The plague mortality data were photographed and/or entered by Seth Earn, Kelly Hancock, Claire Lees, James McDonald, David Price and Hannah Price.
 Valerie Hart facilitated work at the Guildhall Library in London.
@@
 Sang Woo Park assisted in finding published estimates of auxiliary epidemic parameters.
-Alex Bushby assisted with compiling probate dates.
+Alexandra Bushby assisted with compiling probate dates.
 Sigal Balshine, Ann Carmichael and David Price made helpful comments.
@@
 SHARCnet (\url{http://www.sharcnet.ca}) provided computational resources.
-David Earn dedicates this paper to Josephine Earn, who enthusiastically discussed the history of London (and plagues) with him over many years.
+David Earn dedicates this paper to Josephine Earn, who enthusiastically discussed the history of London (and plagues) with him over many years.}
+
+\showacknow{} % Display the acknowledgments section

### 2020-07-12 supp.Rnw
supp now compiles with PNAS SI format

@@
-\documentclass[12pt,table]{article}
+\documentclass[9pt,twocolumn,twoside,lineno,table]{pnas-new}
+% Use the lineno option to display guide line numbers if required.
 %% knitr requires documentclass to be the first line of the file
@@
 %%          30 Apr 2017
-\input{nlipreamble.tex}
 
@@
 
-%%\definecolor[named]{verylightgray}{rgb}{0.95,0.95,0.95}
+\usepackage{xspace}
+\usepackage{cprotect} % allow verbatim in caption
+\usepackage{hhline} % Better horizontal lines in tabulars and arrays
+
+%% cleveref package for convenient hyperrefercing/citing:
+\usepackage[nameinlink,capitalize]{cleveref}
+\crefname{equation}{Equation}{Equations}
+\Crefname{equation}{Equation}{Equations}
+\crefname{figure}{Figure}{Figures}
+\Crefname{figure}{Figure}{Figures}
+
 \definecolor[named]{verylightgray}{rgb}{0.93,0.93,0.93}
 
-\newcommand{\mysection}[1]{\clearpage\section*{#1}\addcontentsline{toc}{section}{#1}}
-\newcommand{\mysectionsamepage}[1]{\section*{#1}\addcontentsline{toc}{section}{#1}}
+\input{analysis/fits/modeldefs.tex}
+\input{macros.tex}
+
 %% http://tex.stackexchange.com/questions/316510/hyperlink-within-caption-to-a-figure
 %% if used in a caption, need short content \caption[...]{ ... \Methodslink ... }
-\newcommand{\Methodslink}{\href[page=9]{ms.pdf}{\emph{Methods}}}
-\newcommand\ttbackslash{{\tt\char`\\}}
-\newcommand{\macro}[1]{{\tt\ttbackslash#1}}
+\renewcommand{\Methodslink}{\href[page=8]{ms.pdf}{\emph{Methods}}}
 
@@
 
-%%\title{Faster growth of later plagues in the second millenium\\\emph{Electronic Supplementary Material}}
-\title{Acceleration of plague outbreaks in the second
-  pandemic\\\emph{Electronic \supp}}
+\templatetype{pnassupportinginfo}
 
-\author{David J.\,D.\ Earn$^{a,b,c}$, 
-Junling Ma$^{f}$, 
-Hendrik N.\ Poinar$^{b,c,d,e}$,\\
-Jonathan Dushoff$^{a,b,c}$,
-and Benjamin M.\ Bolker$^{a,b,c}$\\
-\href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
-\small $^a$Department of Mathematics \& Statistics, $^b$Department of Biology, \\
-\small $^c$Michael G.\ deGroote Institute for Infectious Disease Research,\\
-\small $^d$McMaster Ancient DNA Centre, Department of Anthropology,\\
-\small and $^e$Department of Biochemistry,\\
-\small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
-\small $^f$Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}
+\title{Acceleration of plague outbreaks in the second pandemic}
+
+\author{David J.\,D.\ Earn, Junling Ma, Hendrik Poinar, Jonathan Dushoff, Benjamin M.\ Bolker}
+\correspondingauthor{E-mail: \texttt{earn@math.mcmaster.ca}}
+
+% Please give the surname of the lead author for the running footer
+\leadauthor{Earn}
 
@@
 
-\linenumbers
-
-\null\vspace{-1.5cm}
-
-\vbox{
 \maketitle
 
-\tableofcontents
-}
+%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
+\SItext
 
-\section{Historical human population estimates for London}
+\section*{Historical human population estimates for London}
+
+See \cref{T:london_pop} and \cref{F:london_pop}.
 
 %
-\begin{table}[h!]
-\caption{Population estimates for London previously published by Finlay \cite{Finl81,FinlShea86}\cite[p.\,108]{Kryl11} and Creighton \cite[p.\,660]{Crei65}.
+\begin{table}%[h!]
+\caption[Population estimates for London]{Population estimates for London previously published by Finlay \cite{Finl81,FinlShea86}\cite[p.\,108]{Kryl11} and Creighton \cite[p.\,660]{Crei65}.
 These data are plotted in \cref{F:london_pop}.}
@@
 \end{center}
+\label{T:london_pop}
 \end{table}
@@
 %
-\begin{figure}[h!]
+\begin{figure*}%[h!]
 \begin{center}
@@
 \end{center}
-\caption{Historical population of London, England, as estimated by Finlay \cite{Finl81,FinlShea86} (\cref{tab:pop}).
+\caption[Historical population of London]{Historical population of London, England, as estimated by Finlay \cite{Finl81,FinlShea86} (\cref{tab:pop}).
 Earlier estimates of Creighton \cite[p.\,660]{Crei65} are shown for the specific plague years 1603, 1625 and 1665.}
 \label{F:london_pop}
-\end{figure}
+\end{figure*}
 %
 
-\FloatBarrier
-\section{Disease-related data sources}
+%%\FloatBarrier
+\section*{Disease-related data sources}
 
@@
 \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\begin{table}[h!]
+\begin{table*}%[h!]
 \caption[]{Disease-related data used in this paper.}
@@
 \end{center}
-\end{table}
+\end{table*}
 
-\subsection{Last wills and testaments}
+\subsection*{Last wills and testaments}
 
@@
 
-\subsection{Will dates vs.\ probate dates}
+\subsection*{Will dates vs.\ probate dates}
 
@@
 \end{center}
-\caption{Counts of wills written vs.\ wills probated during each of
+\caption[Wills written vs.\ probated]{Counts of wills written vs.\ wills probated during each of
   the four 14th century plague epidemics in London, based on wills
@@
 \label{F:willsVSprobates}
-\end{figure}
+\end{figure*}
 %
 
-\subsection{London Bills of Mortality (LBoM)}
+\subsection*{London Bills of Mortality (LBoM)}
 
@@
 
-\subsection{Parish registers}
+\subsection*{Parish registers}
 
@@
 
-\subsection{Major vs minor plague epidemics}
+\subsection*{Major vs minor plague epidemics}
 
@@
 \end{center}
-\caption{Plague deaths per 1000 persons per week, from 1563 to 1666.
+\caption[Plague deaths per 1000 per week]{Plague deaths per 1000 persons per week, from 1563 to 1666.
 Epidemics that exceeded 5 on this scale were classified as \emph{major}.
@@
 \label{F:all1600s.scaled}
-\end{figure}
+\end{figure*}
 %
 
-\begin{figure}[h!]
+\begin{figure*}%[h!]
 \begin{center}
@@
 \end{center}
-\caption{All plague deaths reported in London from 1563 to 1666.}
+\caption[Plague deaths]{All plague deaths reported in London from 1563 to 1666.}
 \label{F:all1600s}
-\end{figure}
+\end{figure*}
 %
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[width=\textwidth]{analysis/plots/plotwills14c.pdf}
+\includegraphics[width=0.95\textwidth]{analysis/plots/plotwills14c.pdf}
 \end{center}
@@
 \label{F:plotwills14c}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\begin{figure*}
 \begin{center}
-\includegraphics[width=\textwidth]{analysis/plots/plot3sourcesmajors.pdf}
+\includegraphics[width=0.95\textwidth]{analysis/plots/plot3sourcesmajors.pdf}
 \end{center}
@@
 \label{F:plot3sourcesmajors}
-\end{figure}
+\end{figure*}
 
 %%\clearpage
-\section{Generation interval for bubonic plague}\label{sec:gentime}
+\section*{Generation interval for bubonic plague}\label{sec:gentime}
 
@@
 periods (and hence of the generation interval distribution)
-\cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given
+\cite{GaniLeac04}, which we can use to estimate $\R_0$ for a given
 value of $r$ \cite{WallLips07}.
@@
 
-\section{Relationship between rat density and $\R_0$ (in rats)}
+\section*{Relationship between rat density and $\R_0$ (in rats)}
 
@@
 \end{center}
-\caption{Relative change in rat carrying capacity $K_{\rm R}$ as a
+\caption[Relative change in rat carrying capacity]{Relative change in rat carrying capacity $K_{\rm R}$ as a
   function of relative change in basic reproduction number $\R_0$.
@@
 \label{F:KvsR}
-\end{figure}
+\end{figure*}
 %
 
-\input{analysis/fits/modeldefs.tex}
-\section{Effects of rat ecology on growth rates}
+\section*{Effects of rat ecology on growth rates}
 
@@
 As a vector-borne disease, the rate of bubonic plague spread is primarily affected by the flea-rat ratio.
-In the traditional Ross-MacDonald model for vector-borne epidemics, $\rzero$ is proportional to the square of vector-host ratio because vectors must independently bite hosts twice (once to become infected and a second time to infect a susceptible host).
-In the rat-flea model, $\rzero$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
+In the traditional Ross-MacDonald model for vector-borne epidemics, $\R_0$ is proportional to the square of vector-host ratio because vectors must independently bite hosts twice (once to become infected and a second time to infect a susceptible host).
+In the rat-flea model, $\R_0$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
 From Keeling and Gilligan's plague model \cite{KeelGill00b}, the
@@
 
-As noted in the main text, rat density has an additional, indirect effect on growth rate $r$ and reproduction number $\rzero$: increasing rat density will increase the probability that fleas leaving dying rats will find new, susceptible rat hosts.
-To crudely estimate the magnitude of this effect, if we again consider Keeling and Gilligan's model \cite{KeelGill00b} then the maximum possible effect of rat density on $\rzero$ by this mechanism would be proportional (if fleas have very low success in finding new rat hosts).
+As noted in the main text, rat density has an additional, indirect effect on growth rate $r$ and reproduction number $\R_0$: increasing rat density will increase the probability that fleas leaving dying rats will find new, susceptible rat hosts.
+To crudely estimate the magnitude of this effect, if we again consider Keeling and Gilligan's model \cite{KeelGill00b} then the maximum possible effect of rat density on $\R_0$ by this mechanism would be proportional (if fleas have very low success in finding new rat hosts).
 That is, in order for $\rzerob$ to increase from $\approx \rzeroearlybest$ in the 14th century to $\approx \rzerolatebest$ in the late epoch, rat density would have to increase by at least $\ratpct\%$.
-If fleas were already fairly successful at finding hosts in the 14th century, then changes in rat density would be expected to have only a small effect on $\rzero$ in rats.
+If fleas were already fairly successful at finding hosts in the 14th century, then changes in rat density would be expected to have only a small effect on $\R_0$ in rats.
 
-\section{Analysis of delays between wills and other sources}
+\section*{Analysis of delays between wills and other sources}
 
@@
 
-\begin{table}[h!]
-\caption{Observed cross-correlation and delay between London Bills of
-  Mortality (LBoM) records and other sources. All lags measured in
+\begin{table*}%[h!]
+  \caption[Observed cross-correlation and delay]{Observed cross-correlation and delay between London Bills of Mortality (LBoM) records and other sources. All lags measured in
     weeks.  Rows corresponding to wills data are highlighted in grey.
@@
 \label{tab:delays}
-\end{table}
+\end{table*}
 
-\section{Details of numerical optimization}
+\section*{Details of numerical optimization}
 
@@
 of optimizers --- specifically the BFGS and Nelder-Mead options for
-R's \code{optim} function as well as R's \code{nlminb} function; we
+R's \texttt{optim} function as well as R's \texttt{nlminb} function; we
 also tried an approach that iterated back and forth between
 Nelder-Mead and BFGS until convergence was achieved or the fit was
-sufficiently stable. We chose \code{nlminb} because it gave the best
+sufficiently stable. We chose \texttt{nlminb} because it gave the best
 results (highest log likelihood) for point estimates.
@@
 \FloatBarrier
-%%\clearpage
-\section{Supplementary tables cited in the main text}
+\clearpage
+\section*{Supplementary tables cited in the main text}
 
-\begin{table}[h!]
-    \cprotect\caption{Summary statistics for the model estimating differences across epochs. All parameters are in units of log(growth rate)/year.
+\begin{table*}%[h!]
+    \cprotect\caption[Summary statistics]{Summary statistics for the model estimating differences across epochs. All parameters are in units of log(growth rate)/year.
       The model includes fixed effects of epoch (early [14th c.]
@@
 \label{tab:summarystats}
-\end{table}
+\end{table*}
 
-\begin{table}
+\begin{table*}
   \caption[Minor outbreak parameter estimates]{Parameter estimates for minor plague epidemics.
@@
   \label{tab:minorparams}
-\end{table}
+\end{table*}
 
-\begin{table}
+\begin{table*}
   \caption[Fitting windows]{Outbreak windows and fitting windows used in fitting the major plague epidemics.  See \Methodslink.}
@@
   \label{tab:windows}
-\end{table}
+\end{table*}
 
-\begin{table}
+\begin{table*}
   \caption[Fitting windows (minors)]{The start and end dates for the fitting windows used in fitting the minor plague epidemics.  See \Methodslink.}
-  \scriptsize
+  \large
   \rowcolors{2}{white}{verylightgray}
@@
   \label{tab:windowsminor}
-\end{table}
+\end{table*}
 
 \FloatBarrier
-\section{Supplementary figure cited in the main text}
+\section*{Supplementary figure cited in the main text}
 
 %
-\begin{figure}[h!]
+\begin{figure*}%[h!]
 \begin{center}
@@
 \label{F:seasonalgrowth}
-\end{figure}
+\end{figure*}
 %
 
-%%\clearpage
-\bibliographystyle{pnas-new}
+%%% Add this line AFTER all your figures and tables
+%%\FloatBarrier
+
+%%\dataset{dataset_one.txt}{Type or paste legend here.}
+
+%%\dataset{dataset_two.txt}{Type or paste legend here. Adding longer text to show what happens, to decide on alignment and/or indentations for multi-line or paragraph captions.}
+
+\clearpage
 \bibliography{plague}
-\addcontentsline{toc}{section}{References}
 

### 2020-07-12 supp.Rnw
OK except for supp_crossrefs.tex, which I can no longer include in ms.tex and consequently refs to supp are broken in ms

@@
 % Use the lineno option to display guide line numbers if required.
+
 %% knitr requires documentclass to be the first line of the file
-%% Created: 19 Mar 2017
-%% Changed: 22 Apr 2017
-%%          30 Apr 2017
 
@@
 
-%
-\begin{table}%[h!]
-\caption[Population estimates for London]{Population estimates for London previously published by Finlay \cite{Finl81,FinlShea86}\cite[p.\,108]{Kryl11} and Creighton \cite[p.\,660]{Crei65}.
-These data are plotted in \cref{F:london_pop}.}
-\label{tab:pop}
-\begin{center}
-%% \input{analysis/tables/supppop.tex} % generated by analysis/tables/supppop.R
-%% DE: I'm building the (somewhat different) table myself.
-\renewcommand{\arraystretch}{1.0}
-\begin{tabular}{c|r|l}
-%%\hline
-{\bfseries Year} &
-{\bfseries Population} &
-{\bfseries Source}\\
-\hline
-  1340 &  45,000 & Finlay \\ 
-  1500 &  50,000 & Finlay \\ 
-  1550 & 110,000 & Finlay \\ 
-  1600 & 180,000 & Finlay \\ 
-  1603 & 250,000 & Creighton \\
-  1625 & 320,000 & Creighton \\
-  1650 & 340,000 & Finlay \\ 
-  1665 & 460,000 & Creighton \\
-  1700 & 445,000 & Finlay \\ 
-\end{tabular}
-\end{center}
-\label{T:london_pop}
-\end{table}
-%
-
-%
-\begin{figure*}%[h!]
-\begin{center}
-\includegraphics[width=\textwidth]{analysis/plots/london_pop.pdf}
-\end{center}
-\caption[Historical population of London]{Historical population of London, England, as estimated by Finlay \cite{Finl81,FinlShea86} (\cref{tab:pop}).
-Earlier estimates of Creighton \cite[p.\,660]{Crei65} are shown for the specific plague years 1603, 1625 and 1665.}
-\label{F:london_pop}
-\end{figure*}
-%
-
-%%\FloatBarrier
 \section*{Disease-related data sources}
@@
 paper.
-
-%% requires tabularx package:
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\begin{table*}%[h!]
-\caption[]{Disease-related data used in this paper.}
-\label{tab:datasources}
-\begin{center}
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{c|c|L{3cm}|c|L{3.5cm}}
-%%\hline
-{\bfseries Epoch} &
-{\bfseries Year range} &
-{\bfseries Data type} &
-{\bfseries Frequency} &
-{\bfseries Source} %%&
-%%{\bfseries Description}
- \\
-\hline
-Early & 1340--1380 & Last Wills and Testaments & Daily & Court of Husting \\\hline
-Late & 1540--1680 & Last Wills and Testaments & Daily & Prerogative Court of Canterbury (PCC) \\
- & 1540--1680 & Mortality from all causes & Weekly & Parish records\\
- & 1563--1666 & Mortality from plague & Weekly & London Bills of Mortality (LBoM)
-\end{tabular}
-\end{center}
-\end{table*}
 
@@
 
-%
-\begin{figure*}%[h!]
-\begin{center}
-  \includegraphics[width=\textwidth]{analysis/plots/probate_dates.pdf}
-\end{center}
-\caption[Wills written vs.\ probated]{Counts of wills written vs.\ wills probated during each of
-  the four 14th century plague epidemics in London, based on wills
-  probated in the Court of Husting \cite{Shar1889}. Smooth lines represent loess fits
-with span=0.5 (shaded regions are 95\% confidence intervals).}
-\label{F:willsVSprobates}
-\end{figure*}
-%
-
 \subsection*{London Bills of Mortality (LBoM)}
@@
 
-%%\newcommand{\mywidth}{0.8\textwidth}
-\newcommand{\mywidth}{\textwidth}
-%
-\begin{figure*}%[h!]
-\begin{center}
-\includegraphics[width=\mywidth]{analysis/plots/plotallscaled.crop.pdf}
-\end{center}
-\caption[Plague deaths per 1000 per week]{Plague deaths per 1000 persons per week, from 1563 to 1666.
-Epidemics that exceeded 5 on this scale were classified as \emph{major}.
-Analysis of the major epidemics is shown in \cref{tab:params,F:timeseries,F:combdata}.  Corresponding analyses of all of the minor epidemics are presented
-in \cref{tab:minorparams,F:seasonalgrowth}.
-}
-\label{F:all1600s.scaled}
-\end{figure*}
-%
-
-\begin{figure*}%[h!]
-\begin{center}
-%%\scalebox{0.6}{
-\includegraphics[width=\mywidth]{analysis/plots/plotall.crop.pdf}
-%%}
-\end{center}
-\caption[Plague deaths]{All plague deaths reported in London from 1563 to 1666.}
-\label{F:all1600s}
-\end{figure*}
-%
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=0.95\textwidth]{analysis/plots/plotwills14c.pdf}
-\end{center}
-\caption[]{Weekly numbers of wills written during plague epidemics in London, England, in the 14th century.
-Vertical grey lines indicate the ``outbreak window'' specified in \cref{tab:windows}.}
-\label{F:plotwills14c}
-\end{figure*}
-
-\begin{figure*}
-\begin{center}
-\includegraphics[width=0.95\textwidth]{analysis/plots/plot3sourcesmajors.pdf}
-\end{center}
-\caption[]{Weekly deaths and numbers of wills written during major plague epidemics in London, England, since 1540.
-For each epidemic, vertical grey lines indicate the beginning and end of the period during which deaths from plague were listed in the bills of mortality.
-The left (right) scale refers to deaths (wills).}
-\label{F:plot3sourcesmajors}
-\end{figure*}
-
-%%\clearpage
-\section*{Generation interval for bubonic plague}\label{sec:gentime}
+\hypertarget{sec:gentime}{}
+\section*{Generation interval for bubonic plague}
 
@@
 %
-If we assume the estimate of $\gentime=18$ days obtained in
-\cref{sec:gentime} above, and the MLE growth rates from early and late wills
+If we assume the estimate of $\gentime=18$ days obtained
+\hyperlink{sec:gentime}{above}, and the MLE growth rates from early and late wills
 listed in \cref{tab:params}, the relative change in reproduction
@@
 
-%
-\begin{figure*}%[h!]
-\begin{center}
- \includegraphics[width=\textwidth]{analysis/plots/KvsR.pdf}
-\end{center}
-\caption[Relative change in rat carrying capacity]{Relative change in rat carrying capacity $K_{\rm R}$ as a
-  function of relative change in basic reproduction number $\R_0$.
-  Solid curves are based on \cref{E:R0ratio}.  The dotted red line is
-  based on \cref{E:aKsmall}.  The \Sexpr{round(R0ratio,2)}-fold increase in $\R_0$ that
-  requires explanation is indicated in yellow.}
-\label{F:KvsR}
-\end{figure*}
-%
-
 \section*{Effects of rat ecology on growth rates}
@@
 
-\begin{table*}%[h!]
-  \caption[Observed cross-correlation and delay]{Observed cross-correlation and delay between London Bills of Mortality (LBoM) records and other sources. All lags measured in
-    weeks.  Rows corresponding to wills data are highlighted in grey.
-  }
-\begin{center}
-  \input{analysis/tables/wills_delay_tab.tex}
-\end{center}
-\label{tab:delays}
-\end{table*}
-
 \section*{Details of numerical optimization}
@@
 
-\FloatBarrier
-\clearpage
 \section*{Supplementary tables cited in the main text}
+
+\cref{tab:summarystats,tab:minorparams,tab:windows,tab:windowsminor}.
+
+\section*{Supplementary figure cited in the main text}
+
+\cref{F:seasonalgrowth}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%
+%% FIGURES %%
+%%%%%%%%%%%%%
+
+%
+\begin{figure*}%[h!]
+\begin{center}
+\includegraphics[width=\textwidth]{analysis/plots/london_pop.pdf}
+\end{center}
+\caption[Historical population of London]{Historical population of London, England, as estimated by Finlay \cite{Finl81,FinlShea86} (\cref{tab:pop}).
+Earlier estimates of Creighton \cite[p.\,660]{Crei65} are shown for the specific plague years 1603, 1625 and 1665.}
+\label{F:london_pop}
+\end{figure*}
+%
+
+%
+\begin{figure*}%[h!]
+\begin{center}
+  \includegraphics[width=\textwidth]{analysis/plots/probate_dates.pdf}
+\end{center}
+\caption[Wills written vs.\ probated]{Counts of wills written vs.\ wills probated during each of
+  the four 14th century plague epidemics in London, based on wills
+  probated in the Court of Husting \cite{Shar1889}. Smooth lines represent loess fits
+with span=0.5 (shaded regions are 95\% confidence intervals).}
+\label{F:willsVSprobates}
+\end{figure*}
+%
+
+\newcommand{\mywidth}{\textwidth}
+%
+\begin{figure*}%[h!]
+\begin{center}
+\includegraphics[width=\mywidth]{analysis/plots/plotallscaled.crop.pdf}
+\end{center}
+\caption[Plague deaths per 1000 per week]{Plague deaths per 1000 persons per week, from 1563 to 1666.
+Epidemics that exceeded 5 on this scale were classified as \emph{major}.
+Analysis of the major epidemics is shown in \cref{tab:params,F:timeseries,F:combdata}.  Corresponding analyses of all of the minor epidemics are presented
+in \cref{tab:minorparams,F:seasonalgrowth}.
+}
+\label{F:all1600s.scaled}
+\end{figure*}
+%
+
+\begin{figure*}%[h!]
+\begin{center}
+%%\scalebox{0.6}{
+\includegraphics[width=\mywidth]{analysis/plots/plotall.crop.pdf}
+%%}
+\end{center}
+\caption[Plague deaths]{All plague deaths reported in London from 1563 to 1666.}
+\label{F:all1600s}
+\end{figure*}
+%
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=0.95\textwidth]{analysis/plots/plotwills14c.pdf}
+\end{center}
+\caption[]{Weekly numbers of wills written during plague epidemics in London, England, in the 14th century.
+Vertical grey lines indicate the ``outbreak window'' specified in \cref{tab:windows}.}
+\label{F:plotwills14c}
+\end{figure*}
+
+\begin{figure*}
+\begin{center}
+\includegraphics[width=0.95\textwidth]{analysis/plots/plot3sourcesmajors.pdf}
+\end{center}
+\caption[]{Weekly deaths and numbers of wills written during major plague epidemics in London, England, since 1540.
+For each epidemic, vertical grey lines indicate the beginning and end of the period during which deaths from plague were listed in the bills of mortality.
+The left (right) scale refers to deaths (wills).}
+\label{F:plot3sourcesmajors}
+\end{figure*}
+
+%
+\begin{figure*}%[h!]
+\begin{center}
+ \includegraphics[width=\textwidth]{analysis/plots/KvsR.pdf}
+\end{center}
+\caption[Relative change in rat carrying capacity]{Relative change in rat carrying capacity $K_{\rm R}$ as a
+  function of relative change in basic reproduction number $\R_0$.
+  Solid curves are based on \cref{E:R0ratio}.  The dotted red line is
+  based on \cref{E:aKsmall}.  The \Sexpr{round(R0ratio,2)}-fold increase in $\R_0$ that
+  requires explanation is indicated in yellow.}
+\label{F:KvsR}
+\end{figure*}
+%
+
+%
+\begin{figure*}%[h!]
+\begin{center}
+\scalebox{0.9}{
+  \includegraphics{analysis/plots/seasonalgrowth.pdf}
+}
+\end{center}
+\caption{Growth rate estimates for all the late epoch (1540--1680) London plague epidemics.
+The growth rates shown for the major epidemics are the same as those in the right panel of \cref{F:combdata}.
+}
+\label{F:seasonalgrowth}
+\end{figure*}
+%
+
+{%%%%%%%%%%%%
+%% TABLES %%
+%%%%%%%%%%%%
+
+%
+\begin{table*}%[h!]
+\caption[Population estimates for London]{Population estimates for London previously published by Finlay \cite{Finl81,FinlShea86}\cite[p.\,108]{Kryl11} and Creighton \cite[p.\,660]{Crei65}.
+These data are plotted in \cref{F:london_pop}.}
+\label{tab:pop}
+\begin{center}
+%% \input{analysis/tables/supppop.tex} % generated by analysis/tables/supppop.R
+%% DE: I'm building the (somewhat different) table myself.
+\renewcommand{\arraystretch}{1.0}
+\begin{tabular}{c|r|l}
+%%\hline
+{\bfseries Year} &
+{\bfseries Population} &
+{\bfseries Source}\\
+\hline
+  1340 &  45,000 & Finlay \\ 
+  1500 &  50,000 & Finlay \\ 
+  1550 & 110,000 & Finlay \\ 
+  1600 & 180,000 & Finlay \\ 
+  1603 & 250,000 & Creighton \\
+  1625 & 320,000 & Creighton \\
+  1650 & 340,000 & Finlay \\ 
+  1665 & 460,000 & Creighton \\
+  1700 & 445,000 & Finlay \\ 
+\end{tabular}
+\end{center}
+\label{T:london_pop}
+\end{table*}
+%
+
+%% requires tabularx package:
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\begin{table*}%[h!]
+\caption[]{Disease-related data used in this paper.}
+\label{tab:datasources}
+\begin{center}
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular}{c|c|L{3cm}|c|L{3.5cm}}
+%%\hline
+{\bfseries Epoch} &
+{\bfseries Year range} &
+{\bfseries Data type} &
+{\bfseries Frequency} &
+{\bfseries Source} %%&
+%%{\bfseries Description}
+ \\
+\hline
+Early & 1340--1380 & Last Wills and Testaments & Daily & Court of Husting \\\hline
+Late & 1540--1680 & Last Wills and Testaments & Daily & Prerogative Court of Canterbury (PCC) \\
+ & 1540--1680 & Mortality from all causes & Weekly & Parish records\\
+ & 1563--1666 & Mortality from plague & Weekly & London Bills of Mortality (LBoM)
+\end{tabular}
+\end{center}
+\end{table*}
+
+\begin{table*}%[h!]
+  \caption[Observed cross-correlation and delay]{Observed cross-correlation and delay between London Bills of Mortality (LBoM) records and other sources. All lags measured in
+    weeks.  Rows corresponding to wills data are highlighted in grey.
+  }
+\begin{center}
+  \input{analysis/tables/wills_delay_tab.tex}
+\end{center}
+\label{tab:delays}
+\end{table*}
 
@@
 
-\FloatBarrier
-\section*{Supplementary figure cited in the main text}
-
-%
-\begin{figure*}%[h!]
-\begin{center}
-\scalebox{0.9}{
-  \includegraphics{analysis/plots/seasonalgrowth.pdf}
-}
-\end{center}
-\caption{Growth rate estimates for all the late epoch (1540--1680) London plague epidemics.
-The growth rates shown for the major epidemics are the same as those in the right panel of \cref{F:combdata}.
-}
-\label{F:seasonalgrowth}
-\end{figure*}
-%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2020-07-12 ms_text.tex
author contributions

@@
 % Please include corresponding author, author contribution and author declaration information
-\authorcontributions{{\color{red}Please provide details of author contributions here.}}
+\authorcontributions{{\color{red}
+%%Please provide details of author contributions here.
+DJDE conceived the project and managed the data entry.  
+All authors contributed to design of analyses.
+BMB, JD, JM, and DJDE conducted analyses.
+DJDE and BMB wrote the paper. 
+All authors revised the text and approved the manuscript for submission.
+}}
 %%\authordeclaration{{\color{red}Please declare any conflict of interest here.}}

### 2020-07-13 trimMS.pl
amsync

@@ -3,7 +3,6 @@ use 5.10.0;
 
 while(<>){
 	s|\{.*[^{/]*/(.*[.]pdf)|{$1|;
+	## s|ms|autosub/Earn_etal_ms| if /externaldocument/;
 	print;
 }
-## < \includegraphics[%% height=\pixht]{WillExample1644_page86_1.eps}
-## > \includegraphics[height=\pixht]{images/WillExample1644_page86_1.pdf}

### 2020-07-14 supp.Rnw
change .crop to _crop
New submission pipeline
avoid warning in supp

@@
 \usepackage{xr-hyper}
+\usepackage{grffile}
 \externaldocument{ms}
@@
 number that needs explanation is
-<<rat_R0_calc,echo=FALSE>>=
+<<rat_R0_calc,echo=FALSE,warning=FALSE>>=
 ratR0 <- function(r){ (r / 365 * 18) + 1 }
@@
 \begin{center}
-\includegraphics[width=\mywidth]{analysis/plots/plotallscaled.crop.pdf}
+\includegraphics[width=\mywidth]{analysis/plots/plotallscaled_crop.pdf}
 \end{center}
@@
 %%\scalebox{0.6}{
-\includegraphics[width=\mywidth]{analysis/plots/plotall.crop.pdf}
+\includegraphics[width=\mywidth]{analysis/plots/plotall_crop.pdf}
 %%}

### 2020-08-28 supp.Rnw
Wills and Testaments hyperlink to ms.pdf
insert DOI and get fix footnote hypertarget

@@
 % Use the lineno option to display guide line numbers if required.
+\definecolor[named]{verylightgray}{rgb}{0.93,0.93,0.93}
+\renewcommand{\linenumberfont}{\normalfont\sffamily\tiny\color{lightgray}}
+\urlstyle{tt}
 
@@
 \Crefname{figure}{Figure}{Figures}
-
-\definecolor[named]{verylightgray}{rgb}{0.93,0.93,0.93}
 
@@
 \author{David J.\,D.\ Earn, Junling Ma, Hendrik Poinar, Jonathan Dushoff, Benjamin M.\ Bolker}
-\correspondingauthor{E-mail: \texttt{earn@math.mcmaster.ca}}
+\correspondingauthor{E-mail: {\color{blue}\texttt{earn@math.mcmaster.ca}}
+  \\\bigskip DOI: \texttt{10.1073/pnas.2004904}
+}
 
@@
 Graphs in the main text (\cref{F:plot3sources,F:timeseries,F:combdata}) are based on counts of wills written.
-See \hyperlink{mainsec:Data.wills}{``Wills and testaments''} in the Data section of the main text for further discussion.
+See \href[page=2]{ms.pdf}{``Wills and testaments''} in the Data section of the main text for further discussion.
 
@@
 
-Cummins \emph{et al.} \cite{Cumm+16} obtained parish register data from \href{https://www.ancestry.com}{ancestry.com} and have made the weekly counts of deaths available at \url{http://neilcummins.com}.
+Cummins \emph{et al.} \cite{Cumm+16} obtained parish register data from \href{https://www.ancestry.com}{\texttt{ancestry.com}} and have made the weekly counts of deaths available at \url{http://neilcummins.com}.
 For the 16th and 17th century epidemics, the parish data provide a third source.
@@
 With this approach we failed to achieve good (convergent) fits for only three combinations of source and outbreak year, all for minor outbreaks (London bills for 1578 and 1582, Canterbury wills for 1581); these cases are excluded from the tables and figures shown below.
+%% hack this target because pnas style breaks footnote hyperrefs:
+\hypertarget{Hfootnote.1}{}
 

### 2020-08-28 analysis/plots/london_pop.R
tidied grid

@@ -34,7 +34,8 @@ plot(total ~ year, data=pop, type="n",
      )
 axis(side=2, at=(0:4)*10^5, labels=c("0","$10^5$", paste("$", 2:4, "\\times 10^5$")), las=1)
 
-grid()
+##grid()
+abline(h=1e5*(1:4), v=seq(1350,1700,by=50), col="grey", lty="dotted")
 
 lines(total ~ year, data=pop,
      type="o", lwd=2, pch=21, bg="cyan", cex=2, xpd=NA

### 2020-08-29 supp.Rnw
in caption to Fig S8, mentioned tables where growth rate estimates are listed
tweaked caption to Table S5 and used correct colours for Sources in column 1

@@
   probated in the Court of Husting \cite{Shar1889}. Smooth lines represent loess fits
-with span=0.5 (shaded regions are 95\% confidence intervals).}
+with \texttt{span=0.5} (shaded regions are 95\% confidence intervals).}
 \label{F:willsVSprobates}
@@
 \end{center}
-\caption{Growth rate estimates for all the late epoch (1540--1680) London plague epidemics.
-The growth rates shown for the major epidemics are the same as those in the right panel of \cref{F:combdata}.
-}
+\caption{Growth rate estimates for all the late epoch (1540--1680)
+  London plague epidemics.  The growth rates shown for the major
+  epidemics are the same as those in the right panel of
+  \cref{F:combdata}.  See \cref{tab:params,tab:minorparams}.}
 \label{F:seasonalgrowth}
@@
 \begin{table*}
-  \caption[Minor outbreak parameter estimates]{Parameter estimates for minor plague epidemics.
-Plague epidemics with $\hat r<1/\mathrm{yr}$ or $\hat r>100/\mathrm{yr}$, representing unreliable fits, are excluded; confidence limits that are $<0.1$ or $>200$ are replaced with {\tt NA}.
-See \Methodslink.}
-  \footnotesize
-  \rowcolors{2}{white}{verylightgray}
+  \caption[Minor outbreak parameter estimates]{Parameter estimates for
+    minor plague epidemics (\emph{cf.}\/ \cref{F:seasonalgrowth}).
+    Plague epidemics with $\hat r<1/\mathrm{yr}$ or
+    $\hat r>100/\mathrm{yr}$, representing unreliable fits, are
+    excluded; confidence limits that are $<0.1$ or $>200$ are replaced
+    with {\tt NA}. See \Methodslink.}
+  \footnotesize \rowcolors{2}{white}{verylightgray}
   \begin{center}
+%% To get the colours used for each data type in plots:
+%% source('plague_growth/analysis/plots/setcolours.R')
+%% colvec1
+%%    wills    parish      LBoM 
+%%" #4D4D4D" "#1F78B4" "#E31A1C" 
+%% t(col2rgb(colvec1))
+%%        red green blue
+%% wills   77    77   77
+%% parish  31   120  180
+%% LBoM   227    26   28
+%% signif(t(col2rgb(colvec1))/255,3)
+%%          red green  blue
+%% wills  0.302 0.302 0.302
+%% parish 0.122 0.471 0.706
+%% LBoM   0.890 0.102 0.110
+\definecolor{willsgrey}{rgb}{0.302,0.302,0.302}
+\definecolor{parishblue}{rgb}{0.122,0.471,0.706}
+\definecolor{LBoMred}{rgb}{0.890,0.102,0.110}
     \input{analysis/tables/minorparams.tex}

### 2020-08-29 analysis/tables/table.R
correct source colours now in pipeline

@@ -67,11 +67,11 @@ latexTable <- function(fit, columns, colnames, digits=1, percent=FALSE, comment=
 
   ## ssn = 'simplify source names':
   ## FIX: why does x get altered to numeric by the first ifelse ?!?
-  ## FIX: colours should be exactly the same as in plots
+  ## colours should be exactly the same as in plots [DONE, 29 Aug 2020]
   ssn <- function(x) {
     x <- ifelse(grepl("wills",x), "Wills", x)
-    x <- ifelse(grepl("bills",x), "{\\color{red}LBoM}", x)
-    x <- ifelse(grepl("parish",x),"{\\color{blue}Parish}", x)
+    x <- ifelse(grepl("bills",x), "{\\color{LBoMred}LBoM}", x)
+    x <- ifelse(grepl("parish",x),"{\\color{parishblue}Parish}", x)
     return(x)
   }
   data <- data %>% mutate(Source=ssn(as.character(Source)))

### 2020-08-29 analysis/plots/seasonalgrowth.R
tweak minor fits plot for supp

@@ -52,8 +52,9 @@ for (i in seq(nrow(bad))) {
 ## hack since we loaded data *including pdfname*?
 dd <- (bind_rows(minor=minorSum,major=majorSum,.id="severity")
     %>% filter(outbreak.year>1500)
-    %>% mutate(source=mod_source(source))  ## recode source names
+    ## need to drop_bad() **BEFORE** recoding source names
     %>% drop_bad()
+    %>% mutate(source=mod_source(source))  ## recode source names
     %>% filter(conv==0)
     %>% select(outbreak.year,source,
                severity,starts_with("growthrate"))

### 2022-08-02 supp.Rnw
'in main text'
change I neglected to push on 31 Aug 2020: eqref --> cref in three locations

@@
 %% if used in a caption, need short content \caption[...]{ ... \Methodslink ... }
-\renewcommand{\Methodslink}{\href[page=8]{ms.pdf}{\emph{Methods}}}
+\renewcommand{\Methodslink}{\href[page=8]{ms.pdf}{\emph{Methods}} in main text}
 
@@
 had to increase by at least a factor of \Sexpr{round(R0ratio,2)}
-to account for a similar increase \eqref{E:Rratio} in $\R_0$.
+to account for a similar increase [\cref{E:Rratio}] in $\R_0$.
 
-\cref{F:KvsR} shows the exact relationship \eqref{E:R0ratio} for
+\cref{F:KvsR} shows the exact relationship [\cref{E:R0ratio}] for
 several values of $aK_{{\rm R},1}$, together with the approximation
-for small $aK$ \eqref{E:aKsmall}.  Keeling and Gilligan \cite[Table 1,
+for small $aK$ [\cref{E:aKsmall}].  Keeling and Gilligan \cite[Table 1,
 p.\,2221]{KeelGill00b} adopt values of $a$ and $K_{{\rm R}}$ that

### 2022-08-02 install_pkgs.R
install the version of epigrowthfit used for the PNAS plague growth paper

@@ -6,8 +6,13 @@ pkgs <- pkgs[!pkgs %in% rownames(i1)]
 install.packages(pkgs)
 library(remotes)
 install_github("bbolker/bbmle")
-install_github("davidearn/epigrowthfit",dependencies=TRUE)
-
-
-
-
+##
+## The current version of epigrowthfit is installed via
+##     install_github("davidearn/epigrowthfit",dependencies=TRUE)
+## but this current version is not backwards-compatible and will not
+## work with the code for the PNAS plague growth paper.  Therefore, we
+## install the version that was released with the PNAS paper:
+install_github("davidearn/epigrowthfit",
+               ref = "pnas",
+               dependencies = TRUE,
+               build_vignettes = TRUE)			

### 2022-08-04 install_pkgs.R
modified:   install_pkgs.R with some comments

@@ -1,13 +1,17 @@
 ## install all packages needed by library.R
+
+## install packages listed in required_packages file from CRAN
 options(repos=c(CRAN="https://cloud.r-project.org"))
 pkgs <- scan("required_packages", comment="#", what=character(0), quiet=TRUE)
 i1 <- installed.packages()
 pkgs <- pkgs[!pkgs %in% rownames(i1)]
 install.packages(pkgs)
+
+## install the correct version of bbmle??
 library(remotes)
 install_github("bbolker/bbmle")
-##
-## The current version of epigrowthfit is installed via
+
+## Install an older version of epigrowthfit (provides the package epigrowthfitPNAS)
 ##     install_github("davidearn/epigrowthfit",dependencies=TRUE)
 ## but this current version is not backwards-compatible and will not
 ## work with the code for the PNAS plague growth paper.  Therefore, we

### 2022-08-04 analysis/fits/comb.R
update many R scripts with perl (project not run yet)

@@ -4,7 +4,7 @@ output = ".comb.RData"
 library(dplyr)
 library(tidyr)
 library(methods) # needed for as(); needs to go *before* epigrowthfit!
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 
 ## testing GOF
 ## summary(fitList[[2]][[1]])

### 2022-08-04 analysis/fits/doFits.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(optimx)
 library(dfoptim)
 

### 2022-08-04 analysis/fits/epidemics.R

@@ -1,6 +1,6 @@
 library(methods)
 library(emdbook)
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 
 ## default is warn=0.  warn=1 gives 'real time' warnings rather than

### 2022-08-04 analysis/fits/epochsum.R

@@ -5,7 +5,7 @@ if (packageVersion("glmmTMB")<"0.2.2.0") {
     stop("need latest version of glmmTMB: ",
          "devtools::install_github('glmmTMB/glmmTMB/glmmTMB')")
 }
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(emmeans)
 library(multcomp)
 

### 2022-08-04 analysis/fits/permtest.R

@@ -1,7 +1,7 @@
 library(methods)
 library(dplyr)
 
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(glmmTMB)
 ss <- (summary.fitList(fitList,level=2)
     ## log-transform ... and rename ...

### 2022-08-04 analysis/fits/simfuns.R

@@ -1,6 +1,6 @@
 library(dplyr)
 library(purrr)
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 
 ## extract 'sampling' design from data; which outbreak years
 ##  do we have which sources for?

### 2022-08-04 analysis/fits/wills_delay.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(tidyr)
 

### 2022-08-04 analysis/plots/combdata.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(ggplot2); theme_set(theme_bw())
 library(dplyr)
 library(directlabels)

### 2022-08-04 analysis/plots/london_pop.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(tikzDevice)
 

### 2022-08-04 analysis/plots/plot3sources.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(lubridate)
 library(tikzDevice)

### 2022-08-04 analysis/plots/plot3sourcesmajors.R

@@ -2,7 +2,7 @@
 ## epidemics are evident simultaneously no matter what source we
 ## examine
 
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(lubridate)
 library(tikzDevice)

### 2022-08-04 analysis/plots/plot_allsources.R

@@ -13,7 +13,7 @@ height = if (length(args) < 4) defaults$height else as.numeric(args[4])
 
 cat(width,height,"\n")
 
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(lubridate)
 library(tikzDevice)

### 2022-08-04 analysis/plots/plotall.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(ggplot2)
 library(tikzDevice)
 

### 2022-08-04 analysis/plots/plotwills14c.R

@@ -1,6 +1,6 @@
 ## show 14th century Husting wills in style of plot3sourcesmajors.R
 
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(lubridate)
 library(tikzDevice)

### 2022-08-04 analysis/plots/seasonalgrowth.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(cowplot)
 library(ggplot2); theme_set(theme_bw())
 library(dplyr)

### 2022-08-04 analysis/plots/timeseries.R

@@ -4,7 +4,7 @@ library(cowplot)
 library(ggplot2)
 	theme_set(theme_bw()) ## FIXME: don't hardcode? (coded in plot.R?)
 library(directlabels)
-library(epigrowthfit) ## for parish data
+library(epigrowthfitPNAS) ## for parish data
 
 
 args = commandArgs(trailingOnly=TRUE)

### 2022-08-04 analysis/tables/minorparams.R

@@ -1,6 +1,6 @@
 library(methods)
 library(emdbook)
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(xtable)
 

### 2022-08-04 analysis/tables/params.R

@@ -1,6 +1,6 @@
 library(methods)
 library(emdbook)
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 library(xtable)
 
@@ -13,7 +13,7 @@ rfun <- function(n) {
 ## testing
 ## summary(fitList[[1]][[1]],plague_R0=TRUE)
 ## trace("summary",sig="epigrowthfit", browser)
-## debug(epigrowthfit:::summary.epigrowthfit)
+## debug(epigrowthPNAS:::summary.epigrowthfit)
 ss <- (summary.fitList(fitList,level=2,plague_R0=TRUE)
     %>% drop_bad()
     %>% select(source,outbreak.year,

### 2022-08-04 analysis/tables/suppwindows.R

@@ -4,7 +4,7 @@ majors_only = if (length(args) < 2) TRUE else as.logical(args[2])
 
 ##library(xtable)
 library(Hmisc) # for latex()
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 ## source("analysis/data.R")
 

### 2022-08-04 analysis/tables/suppwindowsminor.R

@@ -3,7 +3,7 @@ file = if (length(args) < 1) "suppwindowsminor.tex" else args[1]
 majors_only = if (length(args) < 2) TRUE else as.logical(args[2])
 
 library(Hmisc) # for latex()
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 
 window_defs <- (summary.fitList(fitList,key="type")

### 2022-08-04 data/get_probate_counts.R

@@ -1,4 +1,4 @@
-library(epigrowthfit)
+library(epigrowthfitPNAS)
 library(dplyr)
 source("../analysis/fits/epidemics.R")
 library(ggplot2); theme_set(theme_bw())

### 2022-08-04 epiPEN.pl

@@ -1,0 +1,4 @@
+while(<>){
+	s/pnas/snap/;
+	print;
+}

### 2022-08-04 install_pkgs.R
package install machinery updates

@@ -4,19 +4,15 @@
 options(repos=c(CRAN="https://cloud.r-project.org"))
 pkgs <- scan("required_packages", comment="#", what=character(0), quiet=TRUE)
 i1 <- installed.packages()
-pkgs <- pkgs[!pkgs %in% rownames(i1)]
+pkgs <- setdiff(pkgs, rownames(i1))
 install.packages(pkgs)
-
-## install the correct version of bbmle??
-library(remotes)
-install_github("bbolker/bbmle")
 
 ## Install an older version of epigrowthfit (provides the package epigrowthfitPNAS)
 ##     install_github("davidearn/epigrowthfit",dependencies=TRUE)
 ## but this current version is not backwards-compatible and will not
 ## work with the code for the PNAS plague growth paper.  Therefore, we
 ## install the version that was released with the PNAS paper:
-install_github("davidearn/epigrowthfit",
+remotes::install_github("davidearn/epigrowthfit",
                ref = "pnas",
                dependencies = TRUE,
                build_vignettes = TRUE)			

### 2022-08-04 library.R

@@ -1,6 +1,8 @@
 p <- scan("required_packages", comment="#", what=character(0), quiet=TRUE)
-p <- c(p, "epigrowthfit", "bbmle")
-res <- suppressMessages(vapply(p,require,character.only=TRUE,logical(1)))
+p <- c(p, "epigrowthfitPNAS")
+res <- suppressPackageStartupMessages(
+    vapply(p, require, quietly = TRUE, character.only=TRUE, logical(1))
+    )
 if (any(!res)) {
     missing <- p[!res]
     stop("please install these missing packages: ",

### 2022-08-04 library.R
merge with Ben?

@@ -1,9 +1,11 @@
 p <- scan("required_packages", comment="#", what=character(0), quiet=TRUE)
-p <- c(p, "epigrowthfit", "bbmle")
+p <- c(p, "epigrowthfitPNAS")
 res <- suppressMessages(vapply(p,require,character.only=TRUE,logical(1)))
 if (any(!res)) {
-    missing <- p[!res]
-    stop("please install these missing packages: ",
-         paste(missing, collapse=", "))
+	missing <- p[!res]
+	stop("please install these missing packages: "
+  		, paste(missing, collapse=", ")
+		, "\nsee install_pkgs.R if you need epigrowthfitPNAS"
+	)
 }
 
